---
title: "Fig_paper"
author: "Anne Baranger"
date: "2026-06-24"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE,message=FALSE)
library(targets)
lapply(c("stringr","tidyr","dplyr","tibble","tidyverse","forcats", "purrr",# data analysis
         "ggplot2","viridis", "ggridges","wesanderson","ggridges","ggrepel","ggnewscale","ggside", # plot
         "factoextra", "nlme","modi","rstan","shinystan","boot", "rstatix","lme4","multcompView","vegan",# stat & models
         "rworldmap" # geographic data #terre, sf
         ),require,character.only=TRUE)
theme_set(theme_bw()) # ggplot theme
common_theme <- theme(
  # legend.position = "none",
  strip.background = element_rect(fill = "white",
                                        colour = "gray12"),
  panel.background = element_rect(fill = "white",
                                  colour = "gray12",
                                  size = 0.5, linetype = "solid"),
  plot.title = element_text(size = 12),
  axis.title = element_text(size = 10),
  axis.text = element_text(size = 10),
  strip.text = element_text(size =10),
  plot.margin = unit(c(0, 0.2, 0, 0.2), "cm")  # Example margin; adjust as needed
)

loadRData <- function(fileName){
  #loads an RData file, and returns it
  load(fileName)
  get(ls()[ls() != "fileName"])
}

```

# Data preparation 

```{r}
# load targets objects
tar_load(c(metric3d_all, metric3d_all_unsliced, metric3d_all_seg,
           regen_metrics_plot, regen_metrics_subplot, plot_df, H_ext_all,
           regen_cat, regen_cat_subplot, var_meta, gini_all))

# def of var_meta
var_meta=tribble(
  ~variable,                ~type,                            ~label,
  "richness",               "Composition",                    "Richness \ntotal",
  "richness_sapling",       "Composition",                    "Richness \nsaplings",
  "H",                      "Composition",                    "Shannon \nindex",
  "cwm_shade",              "Composition",                    "Shade tolerance \n (CWM)",
  "abundance",              "Quantity",                       "Abundance",
  "abundance_cv",           "Quantity",                       "Abundance CV",
  "n_sapling",              "Quantity",                       "Abundance of \nsaplings",
  "h_mean",                 "Quantity",                       "Mean \nheight",
  "pad_tot",                "Quantity",                       "PAD \ntotal",
  "empty_gf",               "Quantity",                       "Gap fraction",
  "Height_increment_mn",    "Quantity",                       "Height \nincrement",
  "H_height",               "Complexity",                     "Vertical \ndistribution",
  "h_cv",                   "Complexity",                     "Height \nvariability",
  "pad_cv",                 "Complexity",                     "PAD \nvariability",
  "ri",                     "Complexity",                     "Rumple index",
  "fhd",                    "Complexity",                     "Foliage height \ndiversity",
  "browsing",               "Other",                          "Browsing",
  "lrb_mean",               "Light quantity",                 "Radiativebudget \nmean (logged)",
  "rb_mean",                "Light quantity",                 "Radiative\nbudget mean",
  "rb_cv",                  "Light horizontal heterogeneity", "Radiative\nbudget CV",
  "rb_sd",                  "Light horizontal heterogeneity", "Radiative\nbudget SD",
  "ext_ref_agg",            "Light quantity",                 "Mean extinction \n at 4.5m",
  "ext_ref_cv",             "Light horizontal heterogeneity", "Mean extinction \n at 4.5m CV",
  "ext_ref_sd",             "Light horizontal heterogeneity", "Mean extinction \n at 4.5m SD",
  "padAbove_agg",           "PAD",                            "PAD above \nregeneration",
  "padAbove_cv",            "PAD",                            "PAD above \nCV",
  "padAbove_sd",            "PAD",                            "PAD above \nSD",
  "padTot_agg",             "PAD",                            "PAD total",
  "padTot_cv",              "PAD",                            "PAD total CV",
  "padTot_sd",              "PAD",                            "PAD total sd",
  "rb_ref_agg",             "Light quantity",                 "RB. at 4.5m",
  "rb_ref_cv",              "Light horizontal heterogeneity", "RB. at 4.5m \nCV",
  "rb_ref_sd",              "Light horizontal heterogeneity", "RB. at 4.5m \nSD",
  "slope_cross_agg",        "Light vertical heterogeneity",   "Light profile slope \nat ext = ",
  "slope_cross_cv",         "Light vertical heterogeneity",   "Light profile slope, CV \nat ext = ",
  "slope_cross_sd",         "Light vertical heterogeneity",   "Light profile slope, SD \nat ext = ",
  "slope_ref_agg",          "Light vertical heterogeneity",   "Light profile slope \nat 4.5m",
  "slope_ref_cv",           "Light vertical heterogeneity",   "Light profile slope, CV\nat 4.5m",
  "slope_ref_sd",           "Light vertical heterogeneity",   "Light profile slope, SD\nat 4.5m",
  "z_cross_agg",            "Light",                          "Height\nat ext = ",
  "z_cross_cv",             "Light",                          "Height, CV\nat ext = ",
  "z_cross_sd",             "Light",                          "Height, SD\nat ext = ",
  "zPAD50_agg",             "PAD",                            "Height at\n 50% of max(PAD)",
  "zPAD50_cv",              "PAD",                            "Height (CV) at\n 50% of max(PAD)",
  "zPAD50_sd",              "PAD",                            "Height (SD) at\n 50% of max(PAD)",
  "gini_h_rb",              "Light horizontal heterogeneity", "Gini index",
  "gini_v_rb",              "Light vertical heterogeneity",   "Gini vert.",
  "bl_diff_abs",            "Light vertical heterogeneity",   "Big leaf dif.",
  "bl_slope_abs",           "Light vertical heterogeneity",   "Big leaf \nslope dif.",
  "PC1",                    "Light",                          "Light PCA axis 1",
  "n_trees_alive",          "Adult",                          "Number of\n living trees",
  "species_richness_alive", "Adult",                          "Adult richness",
  "mean_dbh_alive_cm",      "Adult",                          "Mean diameter",
  "mean_height_alive_m",    "Adult",                          "Mean height",
  "mortality_rate",         "Adult",                          "Mortality",
  "basal_area_alive_m2",     "Adult",                          "Total basal area"
) 


# all tls metrics
tls_metrics = rbind(metric3d_all,metric3d_all_unsliced) %>% 
  left_join(H_ext_all,by=c("plot","subplot","time"))%>% 
  left_join(gini_all,by=c("plot","subplot","time")) %>% 
  mutate(lrb_mean = log(rb_mean))

# inventory 
forest_data_plot=regen_metrics_plot$all %>% 
  left_join(regen_cat) %>% 
  left_join(plot_df)%>%
  mutate(across(c(richness_sapling, n_sapling), ~ replace_na(.x, 0))) %>% 
  mutate(subplot="plot")
forest_data_subplot=regen_metrics_subplot$all %>% 
  left_join(regen_cat_subplot)%>% 
  left_join(plot_df) %>% 
  unique()%>%
  mutate(across(c(richness_sapling, n_sapling), ~ replace_na(.x, 0)))

# function to create dataset
prep_data<-function(df,df_2=NULL,id_cols,plot_level,t=12,s=0,tg=0.9){
  df_f = df %>% 
    filter(if (plot_level=="plot") subplot == "plot" else subplot != "plot",
           time == t,slice_num==s,
           if ("target" %in% colnames(df)) target == .env$tg else TRUE) %>% 
    dplyr::select(any_of(c( c("plot","subplot"),id_cols))) %>% 
    unique()
  if(!is.null(df_2)){
    df_filtered_2 = df_2 %>% 
      filter(
        if (plot_level == "plot") subplot == "plot" else subplot != "plot",
        if ("time" %in% colnames(df_2)) time == .env$t else TRUE,
        if ("slice_num" %in% colnames(df_2)) slice_num == .env$s else TRUE
      ) %>% 
      dplyr::select(any_of(c( c("plot","subplot"),id_cols))) %>% 
      unique()
    df_f=df_f%>% 
      left_join(df_filtered_2,by=c("plot","subplot"))
  }
  return(df_f)
}
```

# Regeneration structure pattern


```{r, fig.width=10, fig.height=6}
# varaibles for the pca
var_pca=c('richness','H','richness_sapling','cwm_shade','abundance','n_sapling','Height_increment_mn','H_height','empty_gf','pad_tot','pad_cv','ri','fhd')

# create dataset 
pca_data_plot= prep_data(tls_metrics,forest_data_plot,var_pca,plot_level = "plot")
pca_data_subplot= prep_data(tls_metrics,forest_data_subplot,var_pca,plot_level = "subplot")


# compute pca on plots
pca_df   <- pca_data_plot %>% 
  dplyr::select(-plot,-subplot) %>% 
  mutate(across(everything(), as.numeric)) %>%
  drop_na()
pca_res  <- prcomp(pca_df, center = TRUE, scale. = TRUE)

# predictions
pca_tot <- bind_rows(pca_data_plot, pca_data_subplot) %>%
  left_join(regen_cat) %>%
  bind_cols(as_tibble(predict(pca_res, .[, colnames(pca_df)])[, 1:2]))
pts <- pca_tot %>%
  filter(subplot=="plot") %>%
  mutate(plot_level = "Plot")

# arrows coordinated
v <- get_pca_var(pca_res)
var_coord <- tibble(
  variable = rownames(v$coord),
  PC1      = v$coord[, 1],
  PC2      = v$coord[, 2],
  contrib  = rowSums(v$contrib[, 1:2]),
  cos2     = rowSums(v$cos2[, 1:2])
  ) %>%
  left_join(var_meta, by = "variable") %>%
  mutate(type = replace_na(type, "Other"))


# plot
ggplot() +
  geom_point(
    data = pts,
    aes(PC1, PC2, shape = plot_level, size = plot_level, color = reg_category)
  ) +
  geom_xsideboxplot(
    data = pts %>% filter(!is.na(reg_category)),
    aes(x = PC1, y = reg_category, fill = reg_category),
    orientation = "y", outliers = FALSE
  ) +
  geom_ysideboxplot(
    data = pts %>% filter(!is.na(reg_category)),
    aes(y = PC2, x = reg_category, fill = reg_category),
    orientation = "x", outliers = FALSE
  )+
  scale_fill_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  scale_shape_manual(name = "Spatial scale", values = c("Plot" = 19, "Subplot" = 1)) +
  scale_size_manual(name = "Spatial scale", values = c("Plot" = 2.5, "Subplot" = 1.5)) +
  scale_color_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  guides(color="none",size="none",shape="none")+
  new_scale_color() +
  geom_segment(
    data = var_coord,
    aes(x = 0, y = 0, xend = PC1*5, yend = PC2*5, color = type),
    arrow = arrow(length = unit(0.20, "cm")), lineend = "round"
  ) +
  geom_text_repel(
    data = var_coord,
    aes(x = PC1*5, y = PC2*5, label = label, color = type),
    size = 3, show.legend = FALSE
  ) +
  scale_color_brewer(name = "Variable type", palette = "Dark2") +
  theme_minimal() +
  # shrink the side panels + hide their axes
  theme(
    # --- main panel: black frame ---
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    
    # --- side panels: small ---
    ggside.panel.scale = 0.15,          # smaller = thinner strips (try 0.10–0.20)
    
    # --- side panels: no grid ---
    ggside.panel.grid = element_blank(),
    
    # --- side panels: no frame/border of their own, hide their axes ---
    ggside.panel.border = element_blank(),
    ggside.panel.background = element_blank(),
    ggside.axis.text  = element_blank(),
    ggside.axis.ticks = element_blank(),
    
  legend.position = "bottom"
  )+
  labs(
    x = paste0("PCA axis 1 (", round(summary(pca_res)$importance["Proportion of Variance","PC1"]*100, 1), "% of variation)"),
    y = paste0("PCA axis 2 (", round(summary(pca_res)$importance["Proportion of Variance","PC2"]*100, 1), "% of variation)")
  ) + 
  coord_cartesian()->c
c
```
- hard to distinguish different regeneration clusters
- variable type are not necessarly colinear, can even be non correlated
- field-derived and lidar-derived variables roughly non correlated, catching different dimensions
- prediction at the subplot scale = add variability


# Light distribution descriptors - plot level

## PCA of light descriptors

```{r}
# variables of the pca
var_pca_light=c("lrb_mean","ext_ref_agg","rb_cv","gini_h_rb","bl_diff_abs","slope_ref_agg","reg_category")

# create dataset
light_pca <- prep_data(tls_metrics,forest_data_plot,var_pca_light,plot_level = "plot",t=12,s=0)  

# compute pca
df_pca=light_pca[,var_pca_light] %>% select(-reg_category)#"rb_ref_agg","padAbove_cv","rb_cv",
pca_light <- prcomp(df_pca,center = TRUE, scale. = TRUE)

light_pca_pred = cbind(light_pca,predict(pca_light,df_pca)[,c(1,2)])


# get arrows
var <- get_pca_var(pca_light)
var_coord <- as.data.frame(var$coord[, 1:2])
colnames(var_coord) <- c("PC1", "PC2")
var_coord$variable <- rownames(var_coord)
var_coord$contrib <- var$contrib[, 1] + var$contrib[, 2]
var_coord$cos2 <- var$cos2[, 1] + var$cos2[, 2]

var_coord=var_coord %>%
  left_join(var_meta,by="variable") %>% 
  mutate(label=case_when(variable=="slope_cross_agg"~paste0(label," 0.9"),
                         variable=="z_cross_agg"~paste0(label," 0.9"),
                         TRUE~label),
         type=if_else(is.na(type),"Other",type))

# plot
ggplot() +
  geom_point(
    data = light_pca_pred,
    aes(PC1, PC2, color = reg_category),
    size=2.5 #,show.legend = FALSE
  ) +
  scale_color_brewer(name = "Regeneration cover", palette = "YlGn",na.value = "grey") +
  
  new_scale_color() +
  
  geom_segment(
    data = var_coord,
    aes(x = 0, y = 0, xend = PC1*5, yend = PC2*5,  color = type), #alpha = contrib,
    arrow = arrow(length = unit(0.20, "cm")),
    lineend = "round"
  ) +
  geom_text_repel(
    data = var_coord,
    aes(x = PC1*5, y = PC2*5, label = label,  color = type), #alpha = contrib,
    size = 3, show.legend = FALSE
  ) +
  # geom_xsideboxplot(
  #   data = light_pca_pred %>% filter(!is.na(reg_category)),
  #   aes(x = PC1, y = reg_category, fill = reg_category),
  #   orientation = "y", outliers = FALSE
  # ) +
  # geom_ysideboxplot(
  #   data = light_pca_pred %>% filter(!is.na(reg_category)),
  #   aes(y = PC2, x = reg_category, fill = reg_category),
  #   orientation = "x", outliers = FALSE
  # )+
  scale_fill_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  scale_color_brewer(name = "Variable type", palette = "Dark2") +
  guides(color="none")+
  
  coord_equal() +
  theme_minimal() +
  guides(
    alpha = "none",
    color = guide_legend(override.aes = list(alpha = 1))
  ) +
  labs(
    x = paste0("PCA axis 1 (", round(summary(pca_light)$importance["Proportion of Variance","PC1"]*100, 1), "% of variation)"),
    y = paste0("PCA axis 2 (", round(summary(pca_light)$importance["Proportion of Variance","PC2"]*100, 1), "% of variation)")
  )+
  scale_y_continuous(breaks = seq(-10, 10, by = 2))+
  scale_x_continuous(breaks = seq(-10, 10, by = 2))+
  theme(
    # --- main panel: black frame ---
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    
    # --- side panels: small ---
    ggside.panel.scale = 0.15,          # smaller = thinner strips (try 0.10–0.20)
    
    # --- side panels: no grid ---
    ggside.panel.grid = element_blank(),
    
    # --- side panels: no frame/border of their own, hide their axes ---
    ggside.panel.border = element_blank(),
    ggside.panel.background = element_blank(),
    ggside.axis.text  = element_blank(),
    ggside.axis.ticks = element_blank()
  )-> pca_light

pca_light

```

## comparative effects of plant material and light distribution on the structure of regeneration

```{r}
light_pca_comp <- prep_data(tls_metrics,forest_data_plot,
                            c(var_pca_light,"padAbove_agg","padAbove_cv"),
                            plot_level = "plot",t=12,s=0)  %>% 
  pivot_longer(cols = -c("plot","subplot","reg_category")) %>%
  filter(!is.na(reg_category)) %>%
  mutate(name=factor(name,levels = c(var_pca_light,"padAbove_agg","padAbove_cv")),
         reg_category = factor(reg_category)) %>%
  group_by(name) %>% mutate(value = as.numeric(scale(value))) %>% ungroup() %>%
  left_join(var_meta, by = c("name" = "variable")) %>% 
  arrange(name) %>%        # type's level order drives it
  mutate(label=case_when(name=="slope_cross_agg"~paste0(label," 0.9"),
                         name=="z_cross_agg"~paste0(label," 0.9"),
                         TRUE~label),
         label = fct_inorder(label))


cld_one <- function(df) {
  lev <- levels(droplevels(df$reg_category))
  if (length(lev) < 2) return(tibble(reg_category = lev, letter = ""))
  
  kw <- kruskal.test(value ~ droplevels(reg_category), data = df)
  if (is.na(kw$p.value) || kw$p.value > 0.05)
    return(tibble(reg_category = lev, letter = "a"))   # no differences
  
  dt <- dunn_test(df, value ~ reg_category, p.adjust.method = "BH")
  pv <- setNames(dt$p.adj, paste(dt$group1, dt$group2, sep = "-"))
  l  <- multcompLetters(pv)$Letters
  tibble(reg_category = names(l), letter = unname(l))
}

cld_df <- light_pca_comp %>%
  group_by(type, label) %>%
  group_modify(~ cld_one(.x)) %>%
  ungroup() %>%
  mutate(reg_category = factor(reg_category, levels = levels(light_pca_comp$reg_category))) %>%
  # assign y coordinates
  left_join(
    light_pca_comp %>% group_by(type, label, reg_category) %>%
      summarise(y = max(value, na.rm = TRUE), .groups = "drop"),
    by = c("type", "label", "reg_category")
  )


ggplot(light_pca_comp, aes(label, value, fill = reg_category)) +
  geom_boxplot(position = position_dodge(width = 0.8, preserve = "single"),
               width=0.5,outlier.shape = NA) +
  geom_point(aes(color=reg_category),
             position = position_jitterdodge(jitter.width = 0.15,
                                             dodge.width = 0.8,
                                             seed = 1),
             size = 1, alpha = 0.8, shape = 16,
             show.legend = FALSE) +
  geom_text(data = cld_df,
            aes(x = label, y = y + 0.25, label = letter, group = reg_category),
            position = position_dodge(width = 0.8),
            inherit.aes = FALSE, size = 3, vjust = 0) +
  scale_fill_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  scale_color_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  facet_grid(~type, scales = "free_x") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.12)))+
  labs(y="Scaled values",x="")+
  theme_minimal(base_size = 9)+
  theme(legend.position = "none")->a
a

cowplot::plot_grid(pca_light+theme(legend.position="left"),a,ncol=2,align = "h", axis = "tb",
                   rel_widths = c(1, 1.5))


```



# Light distribution effect on regeneration structure - plot level

```{r}
explained <- c("richness","richness_sapling","H","cwm_shade","abundance","n_sapling","Height_increment_mn","empty_gf","H_height","fhd","ri","pad_cv")
explaining <- c("lrb_mean","gini_h_rb","slope_ref_agg","bl_diff_abs")

light_effect <- prep_data(tls_metrics,forest_data_plot,c(explained,explaining),plot_level="plot",
                          t=12,s=0,tg=0.9) %>% 
  mutate(slope_ref_agg=-slope_ref_agg)

order=c(var_meta$label[match(explaining,var_meta$variable)],
        var_meta$label[match(explained,var_meta$variable)])


std_coef <- function(d, y, x) { #, covar = "browsing"
  d <- d %>%
    dplyr::select(all_of(c(y, x))) %>% #, covar
    drop_na() %>%
    mutate(across(where(is.numeric), ~ as.numeric(scale(.))))  # factors (e.g. browsing) left alone
  f <- reformulate(c( x), response = y) #covar,
  unname(coef(lm(f, data = d))[x])
}

# --- bootstrap distribution of that coef ---
boot_coef <- function(d, y, x, B = 1000) { #, covar = "browsing"
  d <- d %>% dplyr::select(all_of(c(y, x))) %>% drop_na() %>% # , covar
      mutate(across(where(is.numeric), ~ as.numeric(scale(.)))) 
  n <- nrow(d)
  f <- reformulate(c( x), response = y) # covar,
  map_dbl(seq_len(B), function(b) {
    db <- d[sample(n, n, replace = TRUE), ]  # <- move this line out of the loop to scale once
    unname(coef(lm(f, data = db))[x])
  })
}

results <- expand_grid(explained = explained, explaining = explaining) %>%
  mutate(
    estimate = map2_dbl(explained, explaining, ~ std_coef(light_effect, .x, .y)),
    draws    = map2(explained, explaining, ~ boot_coef(light_effect, .x, .y, B = 1000))
  )


boot_long <- results %>%
  dplyr::select(explained, explaining, draws) %>%
  unnest_longer(draws) %>%
  rename(coef = draws)

boot_stats <- boot_long %>%
  group_by(explained, explaining) %>%
  summarise(
    lo   = unname(quantile(coef, 0.10)),
    lo5  = unname(quantile(coef, 0.25)),
    med  = unname(quantile(coef, 0.50)),
    hi5  = unname(quantile(coef, 0.75)),
    hi   = unname(quantile(coef, 0.90)),
    nrep = dplyr::n(),
    # test bilatéral : 2 x la proportion de tirages du "mauvais" côté de 0,
    # bornée par la résolution du bootstrap 1/(B+1)
    p_boot = max(2 * min(mean(coef <= 0), mean(coef >= 0)), 1 / (nrep + 1)),
    .groups = "drop"
  )



boot_stats %>% filter(explaining !="bl_diff_abs") %>% 
  mutate(
    p_class = cut(p_boot,
                  breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                  labels = c("< 0.001", "< 0.01", "< 0.05", "< 0.1", "ns"),
                  right = FALSE)
  ) %>%
  pivot_longer(cols = c("explained", "explaining")) %>%
  left_join(var_meta, by = c("value" = "variable")) %>% 
  dplyr::select(-value) %>%
  pivot_wider(names_from = "name", values_from = c("label", "type")) %>%
  mutate(
    type_explained   = factor(type_explained,
                              levels = c("Composition", "Quantity", "Complexity")),
    label_explained  = factor(label_explained,  levels = rev(order)),
    label_explaining = factor(label_explaining, levels = order)
  ) %>%
  ggplot(aes(x = med, y = label_explained)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = .3, colour = "grey70") +
  geom_linerange(aes(xmin = lo,  xmax = hi,  colour = p_class), linewidth = .9) +
  geom_linerange(aes(xmin = lo5, xmax = hi5, colour = p_class), linewidth = 1.5) +
  geom_point(aes(colour = p_class), size = 1.3) +
  scale_colour_manual(
    values = c("< 0.001" = "#08306b", "< 0.01" = "#2171b5",
               "< 0.05"  = "#6baed6", "< 0.1"  = "#c6dbef",
               "ns"      = "grey78"),
    drop = FALSE, name = "p (bootstrap)"
  ) +
  facet_grid(type_explained ~ label_explaining, scales = "free_y", space = "free_y") +
  labs(x = "Standardized coefficient", y = NULL) +
  theme_bw(base_size = 9) +
  theme(
    panel.spacing.x  = unit(.6, "lines"),
    strip.text.y     = element_text(angle = 0),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )

```

# From plot to subplot scale

## PCA of light distribution at subplot scale

```{r}
var_pca_light=c("lrb_mean","ext_ref_agg","rb_cv","gini_h_rb","bl_diff_abs","slope_ref_agg","reg_category")

# create dataset
light_pca_s <- prep_data(tls_metrics,forest_data_subplot,var_pca_light,plot_level = "subplot",t=12,s=0)  

# compute pca
df_pca_s=light_pca_s[,var_pca_light] %>% select(-reg_category)#"rb_ref_agg","padAbove_cv","rb_cv",
pca_light_s <- prcomp(df_pca_s,center = TRUE, scale. = TRUE)

light_pca_pred_s = cbind(light_pca_s,predict(pca_light_s,df_pca_s)[,c(1,2)])


# get arrows
var <- get_pca_var(pca_light_s)
var_coord <- as.data.frame(var$coord[, 1:2])
colnames(var_coord) <- c("PC1", "PC2")
var_coord$variable <- rownames(var_coord)
var_coord$contrib <- var$contrib[, 1] + var$contrib[, 2]
var_coord$cos2 <- var$cos2[, 1] + var$cos2[, 2]

var_coord=var_coord %>%
  left_join(var_meta,by="variable") %>% 
  mutate(label=case_when(variable=="slope_cross_agg"~paste0(label," 0.9"),
                         variable=="z_cross_agg"~paste0(label," 0.9"),
                         TRUE~label),
         type=if_else(is.na(type),"Other",type))

# plot
ggplot() +
  geom_point(
    data = light_pca_pred_s,
    aes(PC1, PC2, color = reg_category),
    size=2.5 #,show.legend = FALSE
  ) +
  scale_color_brewer(name = "Regeneration cover", palette = "YlGn",na.value = "grey") +
  
  new_scale_color() +
  
  geom_segment(
    data = var_coord,
    aes(x = 0, y = 0, xend = PC1*5, yend = PC2*5,  color = type), #alpha = contrib,
    arrow = arrow(length = unit(0.20, "cm")),
    lineend = "round"
  ) +
  geom_text_repel(
    data = var_coord,
    aes(x = PC1*5, y = PC2*5, label = label,  color = type), #alpha = contrib,
    size = 3, show.legend = FALSE
  ) +
  # geom_xsideboxplot(
  #   data = light_pca_pred_s %>% filter(!is.na(reg_category)),
  #   aes(x = PC1, y = reg_category, fill = reg_category),
  #   orientation = "y", outliers = FALSE
  # ) +
  # geom_ysideboxplot(
  #   data = light_pca_pred_s %>% filter(!is.na(reg_category)),
  #   aes(y = PC2, x = reg_category, fill = reg_category),
  #   orientation = "x", outliers = FALSE
  # )+
  scale_fill_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  scale_color_brewer(name = "Variable type", palette = "Dark2") +
  guides(color="none")+
  
  coord_equal() +
  theme_minimal() +
  guides(
    alpha = "none",
    color = guide_legend(override.aes = list(alpha = 1))
  ) +
  labs(
    x = paste0("PCA axis 1 (", round(summary(pca_light_s)$importance["Proportion of Variance","PC1"]*100, 1), "% of variation)"),
    y = paste0("PCA axis 2 (", round(summary(pca_light_s)$importance["Proportion of Variance","PC2"]*100, 1), "% of variation)")
  )+
  scale_y_continuous(breaks = seq(-10, 10, by = 2))+
  scale_x_continuous(breaks = seq(-10, 10, by = 2))+
  theme(
    legend.position = "bottom",
    legend.direction = "vertical",
    # --- main panel: black frame ---
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    
    # --- side panels: small ---
    ggside.panel.scale = 0.15,          # smaller = thinner strips (try 0.10–0.20)
    
    # --- side panels: no grid ---
    ggside.panel.grid = element_blank(),
    
    # --- side panels: no frame/border of their own, hide their axes ---
    ggside.panel.border = element_blank(),
    ggside.panel.background = element_blank(),
    ggside.axis.text  = element_blank(),
    ggside.axis.ticks = element_blank()
  )-> pca_light_s

# pca_light

light_pca_comp_s <- prep_data(tls_metrics,forest_data_subplot,
                            c(var_pca_light,"padAbove_agg","padAbove_cv"),
                            plot_level = "subplot",t=12,s=0)  %>% 
  pivot_longer(cols = -c("plot","subplot","reg_category")) %>%
  filter(!is.na(reg_category)) %>%
  mutate(name=factor(name,levels = c(var_pca_light,"padAbove_agg","padAbove_cv")),
         reg_category = factor(reg_category)) %>%
  group_by(name) %>% mutate(value = as.numeric(scale(value))) %>% ungroup() %>%
  left_join(var_meta, by = c("name" = "variable")) %>% 
  arrange(name) %>%        # type's level order drives it
  mutate(label=case_when(name=="slope_cross_agg"~paste0(label," 0.9"),
                         name=="z_cross_agg"~paste0(label," 0.9"),
                         TRUE~label),
         label = fct_inorder(label))


cld_one <- function(df) {
  lev <- levels(droplevels(df$reg_category))
  if (length(lev) < 2) return(tibble(reg_category = lev, letter = ""))
  
  kw <- kruskal.test(value ~ droplevels(reg_category), data = df)
  if (is.na(kw$p.value) || kw$p.value > 0.05)
    return(tibble(reg_category = lev, letter = "a"))   # no differences
  
  dt <- dunn_test(df, value ~ reg_category, p.adjust.method = "BH")
  pv <- setNames(dt$p.adj, paste(dt$group1, dt$group2, sep = "-"))
  l  <- multcompLetters(pv)$Letters
  tibble(reg_category = names(l), letter = unname(l))
}

cld_df <- light_pca_comp_s %>%
  group_by(type, label) %>%
  group_modify(~ cld_one(.x)) %>%
  ungroup() %>%
  mutate(reg_category = factor(reg_category, levels = levels(light_pca_comp_s$reg_category))) %>%
  # assign y coordinates
  left_join(
    light_pca_comp_s %>% group_by(type, label, reg_category) %>%
      summarise(y = max(value, na.rm = TRUE), .groups = "drop"),
    by = c("type", "label", "reg_category")
  )


ggplot(light_pca_comp_s, aes(label, value, fill = reg_category)) +
  geom_boxplot(position = position_dodge(width = 0.8, preserve = "single"),
               width=0.5,outlier.shape = NA) +
  geom_point(aes(color=reg_category),
             position = position_jitterdodge(jitter.width = 0.15,
                                             dodge.width = 0.8,
                                             seed = 1),
             size = 1, alpha = 0.8, shape = 16,
             show.legend = FALSE) +
  geom_text(data = cld_df,
            aes(x = label, y = y + 0.25, label = letter, group = reg_category),
            position = position_dodge(width = 0.8),
            inherit.aes = FALSE, size = 3, vjust = 0) +
  scale_fill_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  scale_color_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  facet_grid(~type, scales = "free_x") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.12)))+
  labs(y="Scaled values",x="")+
  theme_minimal(base_size = 9)+
  theme(legend.position = "none")->a_s
# a

cowplot::plot_grid(pca_light_s,a_s,ncol=2,align = "h", axis = "tb",
                   rel_widths = c(1, 1.5))


```


## Light distribution effect on regeneration structure - subplot level

```{r}
explained <- c("richness","richness_sapling","H","cwm_shade","abundance","n_sapling","Height_increment_mn","empty_gf","H_height","fhd","ri","pad_cv")
explaining <- c("lrb_mean","gini_h_rb","slope_ref_agg","bl_diff_abs")

light_effect_s <- prep_data(tls_metrics,forest_data_subplot,c(explained,explaining),
                            plot_level="subplot",
                            t=12,s=0,tg=0.9) %>% 
  mutate(slope_ref_agg=-slope_ref_agg)

order=c(var_meta$label[match(explaining,var_meta$variable)],
        var_meta$label[match(explained,var_meta$variable)])


std_coef <- function(d, y, x) {#, covar = "browsing"
  d <- d %>%
    dplyr::select(all_of(c(y, x, "plot"))) %>%#, covar
    drop_na() %>%
    mutate(plot = factor(plot)) %>%                              # grouping factor, kept out of scaling
    mutate(across(where(is.numeric), ~ as.numeric(scale(.))))    # factors (browsing, plot) left alone
  f <- as.formula(paste0(y, " ~ ",  " + ", x, " + (1 | plot)"))#covar,
  m <- lmer(f, data = d, control = lmerControl(calc.derivs = FALSE))
  unname(fixef(m)[x])
}

# --- bootstrap distribution of that coef ---
boot_coef <- function(d, y, x, B = 1000) {# covar = "browsing",
  d <- d %>% dplyr::select(all_of(c(y, x, "plot"))) %>% drop_na() %>%#covar, 
    mutate(plot = factor(plot)) %>%
    mutate(across(where(is.numeric), ~ as.numeric(scale(.))))
  n <- nrow(d)
  f <- as.formula(paste0(y, " ~ ", " + ", x, " + (1 | plot)"))# covar,
  map_dbl(seq_len(B), function(b) {
    db <- d[sample(n, n, replace = TRUE), ]
    m  <- suppressMessages(suppressWarnings(
      lmer(f, data = db, control = lmerControl(calc.derivs = FALSE))))
    unname(fixef(m)[x])
  })
}

results <- expand_grid(explained = explained, explaining = explaining) %>%
  mutate(
    estimate = map2_dbl(explained, explaining, ~ std_coef(light_effect_s, .x, .y)),
    draws    = map2(explained, explaining, ~ boot_coef(light_effect_s, .x, .y, B = 1000))
  )


boot_long <- results %>%
  dplyr::select(explained, explaining, draws) %>%
  unnest_longer(draws) %>%
  rename(coef = draws)

boot_stats <- boot_long %>%
  group_by(explained, explaining) %>%
  summarise(
    lo   = unname(quantile(coef, 0.10)),
    lo5  = unname(quantile(coef, 0.25)),
    med  = unname(quantile(coef, 0.50)),
    hi5  = unname(quantile(coef, 0.75)),
    hi   = unname(quantile(coef, 0.90)),
    nrep = dplyr::n(),
    # test bilatéral : 2 x la proportion de tirages du "mauvais" côté de 0,
    # bornée par la résolution du bootstrap 1/(B+1)
    p_boot = max(2 * min(mean(coef <= 0), mean(coef >= 0)), 1 / (nrep + 1)),
    .groups = "drop"
  )


boot_stats %>% filter(explaining !="bl_diff_abs") %>% 
  mutate(
    p_class = cut(p_boot,
                  breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                  labels = c("< 0.001", "< 0.01", "< 0.05", "< 0.1", "ns"),
                  right = FALSE)
  ) %>%
  pivot_longer(cols = c("explained", "explaining")) %>%
  left_join(var_meta, by = c("value" = "variable")) %>% 
  dplyr::select(-value) %>%
  pivot_wider(names_from = "name", values_from = c("label", "type")) %>%
  mutate(
    type_explained   = factor(type_explained,
                              levels = c("Composition", "Quantity", "Complexity")),
    label_explained  = factor(label_explained,  levels = rev(order)),
    label_explaining = factor(label_explaining, levels = order)
  ) %>%
  ggplot(aes(x = med, y = label_explained)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = .3, colour = "grey70") +
  geom_linerange(aes(xmin = lo,  xmax = hi,  colour = p_class), linewidth = .9) +
  geom_linerange(aes(xmin = lo5, xmax = hi5, colour = p_class), linewidth = 1.5) +
  geom_point(aes(colour = p_class), size = 1.3) +
  scale_colour_manual(
    values = c("< 0.001" = "#08306b", "< 0.01" = "#2171b5",
               "< 0.05"  = "#6baed6", "< 0.1"  = "#c6dbef",
               "ns"      = "grey78"),
    drop = FALSE, name = "p (bootstrap)"
  ) +
  facet_grid(type_explained ~ label_explaining, scales = "free_y", space = "free_y") +
  labs(x = "Standardized coefficient", y = NULL) +
  theme_bw(base_size = 9) +
  theme(
    panel.spacing.x  = unit(.6, "lines"),
    strip.text.y     = element_text(angle = 0),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )

```

 # Temporal variability and spatial variability

```{r}
# create data
var_metrics <- c('ext_ref_agg', "slope_ref_agg", "gini_h_rb")
light_variability =tls_metrics %>% 
  filter(slice_num==0,subplot!="plot") %>% 
  dplyr::select(any_of(c( c("plot","subplot","time"),var_metrics))) %>% 
  unique() %>%
  mutate(across(all_of(var_metrics), as.numeric)) %>%
  pivot_longer(all_of(var_metrics), names_to = "metric", values_to = "value")


# compute temporal and spatial variability
sd_temporal <- light_variability %>%
  group_by(plot, metric, subplot) %>%
  summarise(sd_t = sd(value, na.rm = TRUE),
            m_sub = mean(value, na.rm = TRUE), .groups = "drop") %>%
  group_by(plot, metric) %>%
  summarise(temporal_within = mean(sd_t, na.rm = TRUE),
            spatial_marg    = sd(m_sub, na.rm = TRUE), .groups = "drop")

sd_spatial <- light_variability %>%
  group_by(plot, metric, time) %>%
  summarise(sd_s = sd(value, na.rm = TRUE),
            m_time = mean(value, na.rm = TRUE), .groups = "drop") %>%
  group_by(plot, metric) %>%
  summarise(spatial_within  = mean(sd_s, na.rm = TRUE),
            temporal_marg   = sd(m_time, na.rm = TRUE), .groups = "drop")

res <- light_variability %>%
  group_by(plot, metric) %>%
  summarise(mean = mean(value, na.rm = TRUE),
            n_sub = n_distinct(subplot), n_date = n_distinct(time),
            .groups = "drop") %>%
  left_join(sd_temporal, by = c("plot", "metric")) %>%
  left_join(sd_spatial,  by = c("plot", "metric")) %>%
  mutate(
    # part de la variabilité totale expliquée par le temps (0 = tout spatial,
    # 1 = tout temporel) -> utile pour colorer / trier les plots
    temporal_part = temporal_within / (temporal_within + spatial_within),
    # versions normalisées (CV) pour comparer des métriques d'échelles
    # différentes ; NA si la moyenne est proche de 0 (cas de slope_ref_agg)
    cv_temporal = ifelse(abs(mean) > 1e-8, temporal_within / abs(mean), NA),
    cv_spatial  = ifelse(abs(mean) > 1e-8, spatial_within  / abs(mean), NA)
  )
res %>% 
  left_join(var_meta,by=c("metric"="variable")) %>% 
  mutate(label=factor(label,levels=c("Mean extinction \n at 4.5m","Gini index","Light profile slope \nat 4.5m"))) %>% 
  ggplot(aes(x = spatial_within, y = temporal_within)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point( size = 2.5) +
  # geom_text(aes(label = plot), vjust = -0.9, size = 3, show.legend = FALSE) +
  scale_colour_gradient2(low = "#2c7bb6", mid = "grey80", high = "#d7191c",
                         midpoint = 0.5, limits = c(0, 1),
                         name = "Part\ntemporelle") +
  facet_wrap(~ label) +
  expand_limits(x = 0, y = 0) +
  # coord_equal() +          # retirer si les 2 axes ont des ordres de grandeur très différents
  labs(
    # title = "Variabilité temporelle vs spatiale, par plot",
    # subtitle = "Au-dessus de la diagonale : la variabilité temporelle domine",
    x = "Spatial variability (between subplots)",
    y = "Temporal variability (between simulation times)"
  ) +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank(), strip.background = element_blank())
```

# Confounding effect of adult structure

## Effect size

```{r}
explained <- c("richness","richness_sapling","H","cwm_shade","abundance","n_sapling","Height_increment_mn","empty_gf","H_height","fhd","ri","pad_cv") 
explaining <- c("n_trees_alive","species_richness_alive","mean_dbh_alive_cm","basal_area_alive_m2","mean_height_alive_m","browsing")

adult_effect<- prep_data(tls_metrics,forest_data_plot,id_cols = c(explained,explaining),
                         plot_level = "plot")


order=c(var_meta$label[match(explaining,var_meta$variable)],
        var_meta$label[match(explained,var_meta$variable)])



std_coef <- function(d, y, x) {
  d <- d %>%
    dplyr::select(all_of(c(y, x))) %>%
    drop_na() %>%
    mutate(across(where(is.numeric), ~ as.numeric(scale(.))))  # factors (e.g. browsing) left alone
  f <- reformulate(c( x), response = y)
  unname(coef(lm(f, data = d))[x])
}

# --- bootstrap distribution of that coef ---
boot_coef <- function(d, y, x,  B = 1000) {
  d <- d %>% dplyr::select(all_of(c(y, x))) %>% drop_na() %>%
      mutate(across(where(is.numeric), ~ as.numeric(scale(.)))) 
  n <- nrow(d)
  f <- reformulate(c( x), response = y)
  map_dbl(seq_len(B), function(b) {
    db <- d[sample(n, n, replace = TRUE), ]  # <- move this line out of the loop to scale once
    unname(coef(lm(f, data = db))[x])
  })
}

results <- expand_grid(explained = explained, explaining = explaining) %>%
  mutate(
    estimate = map2_dbl(explained, explaining, ~ std_coef(adult_effect, .x, .y)),
    draws    = map2(explained, explaining, ~ boot_coef(adult_effect, .x, .y, B = 1000))
  )


boot_long <- results %>%
  dplyr::select(explained, explaining, draws) %>%
  unnest_longer(draws) %>%
  rename(coef = draws)

boot_stats <- boot_long %>%
  group_by(explained, explaining) %>%
  summarise(
    lo   = unname(quantile(coef, 0.10)),
    lo5  = unname(quantile(coef, 0.25)),
    med  = unname(quantile(coef, 0.50)),
    hi5  = unname(quantile(coef, 0.75)),
    hi   = unname(quantile(coef, 0.90)),
    nrep = dplyr::n(),
    # test bilatéral : 2 x la proportion de tirages du "mauvais" côté de 0,
    # bornée par la résolution du bootstrap 1/(B+1)
    p_boot = max(2 * min(mean(coef <= 0), mean(coef >= 0)), 1 / (nrep + 1)),
    .groups = "drop"
  )



boot_stats %>% 
  mutate(
    p_class = cut(p_boot,
                  breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                  labels = c("< 0.001", "< 0.01", "< 0.05", "< 0.1", "ns"),
                  right = FALSE)
  ) %>%
  pivot_longer(cols = c("explained", "explaining")) %>%
  left_join(var_meta, by = c("value" = "variable")) %>% 
  dplyr::select(-value) %>%
  pivot_wider(names_from = "name", values_from = c("label", "type")) %>%
  mutate(
    type_explained   = factor(type_explained,
                              levels = c("Composition", "Quantity", "Complexity")),
    label_explained  = factor(label_explained,  levels = rev(order)),
    label_explaining = factor(label_explaining, levels = order)
  ) %>%
  ggplot(aes(x = med, y = label_explained)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = .3, colour = "grey70") +
  geom_linerange(aes(xmin = lo,  xmax = hi,  colour = p_class), linewidth = .9) +
  geom_linerange(aes(xmin = lo5, xmax = hi5, colour = p_class), linewidth = 1.5) +
  geom_point(aes(colour = p_class), size = 1.3) +
  scale_colour_manual(
    values = c("< 0.001" = "#08306b", "< 0.01" = "#2171b5",
               "< 0.05"  = "#6baed6", "< 0.1"  = "#c6dbef",
               "ns"      = "grey78"),
    drop = FALSE, name = "p (bootstrap)"
  ) +
  facet_grid(type_explained ~ label_explaining, scales = "free_y", space = "free_y") +
  labs(x = "Standardized coefficient", y = NULL) +
  theme_bw(base_size = 9) +
  theme(
    panel.spacing.x  = unit(.6, "lines"),
    strip.text.y     = element_text(angle = 0),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )

```
## Variance partition analysis


```{r}

```

# Species analysis

```{r eval=FALSE, include=FALSE}
data.frame(plot=unique(regen_metrics_subplot$species$plot)) %>% 
  expand_grid(subplot=unique(regen_metrics_subplot$species$subplot)) %>% 
  expand_grid(Species=unique(regen_metrics_subplot$species$Species)) %>% 
  filter(Species!="NA") %>% 
  left_join(regen_metrics_subplot$species) %>% 
  dplyr::select(plot,subplot,species=Species,abundance) %>% 
  mutate(abundance=if_else(is.na(abundance),0,abundance)) %>% 
  left_join(metric3d_all %>% 
              filter(time==12,subplot!="plot") %>% 
              dplyr::select(plot,subplot,rb_mean,rb_cv,occupied_gf)) %>% 
  left_join(H_ext_all %>% 
              filter(time==12,extinction==0.5) %>% 
              dplyr::select(plot,subplot,mean,mean_slope)) %>% 
  group_by(species) %>% 
  mutate(n=n()) %>% 
  filter(n>30) %>% 
  dplyr::select(-n) %>% 
  pivot_longer(cols=-c("plot","subplot","species","abundance")) %>% 
  ggplot(aes(abundance,value))+
  geom_point()+
  facet_wrap(species~name,scales="free", ncol=5)



regen_metrics_plot$species %>% 
  dplyr::select(plot,species=Species,abundance) %>% 
  left_join(metric3d_all %>% 
              filter(time==12,subplot=="plot") %>% 
              dplyr::select(plot,subplot,slice_num,rb_mean,rb_cv,occupied_gf)) %>% 
  left_join(H_ext_all %>% 
              filter(time==12,extinction==0.5) %>% 
              dplyr::select(plot,subplot,mean,mean_slope)) %>% View()
  filter(plot!="GER10") %>% 
  group_by(species) %>% 
  mutate(n=n()) %>% 
  filter(n>30) %>% 
  dplyr::select(-n) %>% 
  ggplot(aes(log(abundance),log(rb_mean),color=species))+
  geom_point()+
  geom_smooth(method="gam",se = FALSE)+
  facet_grid(species~slice_num)
  
  
data.frame(plot=unique(regen_metrics_subplot$species$plot)) %>% 
  # expand_grid(subplot=unique(regen_metrics_subplot$species$subplot)) %>% 
  expand_grid(Species=unique(regen_metrics_subplot$species$Species)) %>% 
  filter(Species!="NA") %>% 
  left_join(regen_metrics_subplot$species) %>% 
  dplyr::select(plot,species=Species,abundance) %>% 
  mutate(abundance=if_else(is.na(abundance),0,abundance)) %>% 
  left_join(metric3d_all %>% 
              filter(time==12,subplot=="plot") %>% 
              dplyr::select(plot,subplot,slice_num,rb_mean,rb_cv,occupied_gf)) %>% 
  left_join(H_ext_all %>% 
              filter(time==12,extinction==0.5) %>% 
              dplyr::select(plot,subplot,mean,mean_slope)) %>%
  filter(plot!="GER10") %>% 
  group_by(species) %>% 
  mutate(n=n()) %>% 
  filter(n>30) %>% 
  dplyr::select(-n) %>% 
  ggplot(aes(abundance,log(rb_mean),color=species))+
  geom_point()+
  geom_smooth(method="gam",se = FALSE)+
  facet_grid(species~slice_num)
  



df <- regen_metrics_subplot$all %>% dplyr::select(plot,subplot) %>% unique() %>% 
  expand_grid(Species=unique(regen_metrics_subplot$species$Species))%>%
  filter(Species != "NA") %>%
  left_join(regen_metrics_subplot$species) %>%
  dplyr::select(plot, subplot, species = Species, abundance) %>%
  mutate(abundance=if_else(is.na(abundance),0,abundance),
         presence = if_else(is.na(abundance), "absent", "present")) %>%
  left_join(metric3d_all %>%
              filter(time == 12, subplot != "plot", slice_num == 1) %>%
              dplyr::select(plot, subplot, rb_mean, rb_cv, occupied_gf)) %>%
  mutate(log_rb_mean = log(rb_mean)) %>%
  filter(is.finite(log_rb_mean))



df %>%
  group_by(species) %>%
  filter(n_distinct(presence) == 2, all(table(presence) >= 1)) %>%  # need both groups, a few obs each
  wilcox_test(log_rb_mean ~ presence) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

regen_metrics_subplot$all %>% dplyr::select(plot,subplot) %>% unique() %>% 
  expand_grid(Species=unique(regen_metrics_subplot$species$Species)) %>% 
  filter(Species!="NA") %>% 
  left_join(regen_metrics_subplot$species) %>% 
  dplyr::select(plot,subplot,species=Species,abundance) %>% 
  mutate(presence=if_else(is.na(abundance),"absent","present")) %>% 
  left_join(metric3d_all %>% 
              filter(time==12,subplot!="plot",slice_num==1) %>% 
              dplyr::select(plot,subplot,rb_mean,rb_cv,occupied_gf)) %>% 
  ggplot(aes(species,log(rb_mean),color=presence))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter()+
  ylim(c(5.5,9))
  



df <- regen_metrics_plot$all %>% dplyr::select(plot) %>% unique() %>% 
  expand_grid(Species=unique(regen_metrics_plot$species$Species)) %>% 
  filter(Species!="NA") %>% 
  left_join(regen_metrics_plot$species) %>% 
  dplyr::select(plot,species=Species,abundance) %>% 
  mutate(presence=if_else(is.na(abundance),"absent","present")) %>% 
  left_join(metric3d_all %>% 
              filter(time==12,subplot=="plot",slice_num==1) %>% 
              dplyr::select(plot,rb_mean,rb_cv,occupied_gf)) %>% 
  mutate(log_rb_mean = log(rb_mean)) %>%
  filter(is.finite(log_rb_mean))

regen_metrics_plot$all %>% dplyr::select(plot) %>% unique() %>% 
  expand_grid(Species=unique(regen_metrics_plot$species$Species)) %>% 
  filter(Species!="NA") %>% 
  left_join(regen_metrics_plot$species) %>% 
  dplyr::select(plot,species=Species,abundance) %>% 
  mutate(presence=if_else(is.na(abundance),"absent","present")) %>% 
  left_join(metric3d_all %>% 
              filter(time==12,subplot=="plot",slice_num==1) %>% 
              dplyr::select(plot,rb_mean,rb_cv,occupied_gf)) %>% 
  ggplot(aes(species,log(rb_mean),color=presence))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter()+
  ylim(c(5.5,9))
  
  
library(lme4); library(lmerTest)

species_list <- df %>%
  group_by(species) %>%
  filter(n_distinct(presence) == 2, n_distinct(plot) >= 3) %>%
  pull(species) %>% unique()

results <- lapply(species_list, function(sp) {
  d <- df[df$species == sp, ]
  tryCatch({
    m  <- lm(log_rb_mean ~ presence , data = d)
    co <- summary(m)$coefficients["presencepresent", ]
    data.frame(species = sp,
               estimate  = co["Estimate"],
               std.error = co["Std. Error"],
               statistic = co["t value"],
               p.value   = co["Pr(>|t|)"])
  }, error = function(e)
    data.frame(species = sp, estimate = NA, std.error = NA,
               statistic = NA, p.value = NA))
})

res <- do.call(rbind, results)
res$p.adj <- p.adjust(res$p.value, method = "BH")
res




df$rb_c <- df$log_rb_mean - mean(df$log_rb_mean, na.rm = TRUE)   # centred predictor
df$species <- factor(df$species)

library(glmmTMB)

mnb <- glmmTMB(abundance ~ species * rb_c + (1 | plot),
               data = df, family = nbinom2)

# if there are more zeros than even NB expects (common), add a zero-inflation part:
mzi <- glmmTMB(abundance ~ species * rb_c + (1 | plot),
               ziformula = ~1, data = df, family = nbinom2)

AIC(mnb, mzi)  


library(emmeans)

et <- emtrends(mnb, ~ species, var = "rb_c")   # or mp / mzi
et                                             # per-species slope on log scale

# as rate ratios (abundance multiplies by this per 1-unit rb_mean):
summary(et, infer = TRUE) |>
  transform(RR = exp(rb_c.trend))
```

