---
title: "Figures exploration"
author: "Anne Baranger"
date: "2026-05-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE,message=FALSE)
library(targets)
lapply(c("stringr","tidyr","dplyr","tibble", # data analysis
         "ggplot2","viridis", "ggridges","wesanderson",# plot
         "factoextra", "nlme","modi","rstan","shinystan",# stat & models
         "rworldmap" # geographic data #terre, sf
         ),require,character.only=TRUE)
theme_set(theme_bw()) # ggplot theme
common_theme <- theme(
  # legend.position = "none",
  strip.background = element_rect(fill = "white",
                                        colour = "gray12"),
  panel.background = element_rect(fill = "white",
                                  colour = "gray12",
                                  size = 0.5, linetype = "solid"),
  plot.title = element_text(size = 12),
  axis.title = element_text(size = 10),
  axis.text = element_text(size = 10),
  strip.text = element_text(size =10),
  plot.margin = unit(c(0, 0.2, 0, 0.2), "cm")  # Example margin; adjust as needed
)

loadRData <- function(fileName){
  #loads an RData file, and returns it
  load(fileName)
  get(ls()[ls() != "fileName"])
}

```

# Data preparation 

## Loading classic data. 

```{r}
tar_load(metric3d_all)
tar_load(regen_metrics_plot)
tar_load(regen_metrics_subplot)


metric3d_all %>% 
  select(plot,time) %>% 
  unique() %>% group_by(plot) %>% 
  summarise(`Number of simulated time slots`=n())

metric3d_all %>% 
  select(plot,subplot) %>% 
  unique() %>% group_by(plot) %>% 
  summarise(`Number of subplots`=n()-1)

```

## Get metrics at the plot level 

```{r}
metric3d_all_plot=metric3d_all %>% 
  filter(subplot=="plot") %>% 
  pivot_longer(cols =  c(matches("h_"),"coef_lin")) 

```

 

# Exploration of complexity measures 

```{r,fig.height=10,fig.width=10}
metric3d_all %>% 
  filter(subplot!="plot") %>% 
  pivot_longer(cols = c(matches("h_"),"coef_lin")) %>% 
  ggplot(aes(slice_num,value,color=plot))+
  geom_line(aes(group=interaction(plot,subplot)),alpha = 0.4)+
  geom_point(data=metric3d_all_plot,aes(slice_num,value,color=plot))+
  geom_line(data=metric3d_all_plot,aes(slice_num,value,color=plot,group=interaction(plot,subplot)))+
  facet_grid(name~plot,scale="free_y")+
  theme_bw()


metric3d_all_plot
metric3d_all %>% 
  filter(subplot!="plot") %>% 
  filter(time==12) %>% 
  ggplot(aes(h_mean,h_sd,color=time))+
  geom_point(aes(group=interaction(time,subplot)),alpha = 0.4)+
  geom_line(aes(group=interaction(time,subplot)),alpha = 0.4)+
  geom_point(data=metric3d_all %>% filter(subplot=="plot") %>% filter(time==12),
             aes(h_mean,h_sd,color=time))+
  geom_line(data=metric3d_all %>% filter(subplot=="plot") %>% filter(time==12),
            aes(h_mean,h_sd,color=time,group=interaction(time,subplot)))+
  facet_wrap(~plot,scale="free_y")+
  theme_bw()
```

Main outcomes :
* measure height without accounting for empty areas doesn't make sense. Drop variables "_na"
* median doesn't make sense when accounting for empty areas, drop "h_med"
* "coef_lin" is not informative
* The shape of h_mean=f(slice_num) seems to provide insight on the understory structure. Linear form are more common for plots with less regeneration, while saturating form could be more common for dense understory

# Exploration of field metrics

## relevance of seedling number
```{r}
regen_metrics_subplot %>% 
  ggplot(aes(abundance,n_seedling))+
  geom_point()+
  geom_abline(slope=1)+
  labs(x="Total abundance",
       y="Seedlings number (class 1 and 2)",
       title="Subplot level")

regen_metrics_plot %>% 
  ggplot(aes(abundance,n_seedling))+
  geom_point()+
  geom_abline(slope=1)+
  labs(x="Total abundance",
       y="Seedlings number (class 1 and 2)",
       title="Plot level")
```
Seedling number is highly correlated to total abundance at both subplot and plot level.

## relevance of sapling number
```{r}
regen_metrics_subplot %>% 
  ggplot(aes(abundance,n_sapling))+
  geom_point()+
  geom_smooth(method="lm")+
  labs(x="Total abundance",
       y="Saplings number (class 1 and 2)",
       title="Subplot level")

regen_metrics_plot %>% 
  ggplot(aes(abundance,n_sapling))+
  geom_point()+
  geom_smooth(method="lm")+
  labs(x="Total abundance",
       y="Saplings number (class 1 and 2)",
       title="Plot level")
```


## Browsing metrics
```{r}
regen_metrics_plot %>% 
  select(-Species) %>% 
  pivot_longer(cols=-c("plot","browsing")) %>% 
  mutate(name=factor(name,levels = c("abundance","abundance_cv","n_seedling","richness","richness_sd",
                                     "Height_increment_mn_hclass2","Height_increment_mn_hclass3",
                                     "Height_increment_cv_hclass2","Height_increment_cv_hclass3",
                                     "H",
                                     "H_hclass1","H_hclass2","H_hclass3","H_hclass4","H_hclass5"))) %>% 
  ggplot(aes(browsing,value))+
  geom_point()+
  geom_smooth(method="lm")+
  facet_wrap(~name,scale="free_y",ncol=5)+
  theme_bw()+
  labs(title = "At the plot scale")



regen_metrics_subplot %>% 
  select(-Species) %>% 
  pivot_longer(cols=-c("plot","subplot","browsing")) %>% 
  mutate(name=factor(name,levels = c("abundance","abundance_cv","n_seedling","richness","richness_sd",
                                     "Height_increment_mn_hclass2","Height_increment_mn_hclass3",
                                     "Height_increment_cv_hclass2","Height_increment_cv_hclass3",
                                     "H",
                                     "H_hclass1","H_hclass2","H_hclass3","H_hclass4","H_hclass5"))) %>% 
  ggplot(aes(browsing,value))+
  geom_point()+
  geom_smooth(method="lm")+
  facet_wrap(~name,scale="free_y",ncol=5)+
  theme_bw()+
  labs(title = "At the subplot scale")
```
Browsing reduces height increment, probably differs between species -> test it
It probably reduces abundance, but this is uncertain. 
Linked to higher richness and to higher sd in richness, at the plot scale.

### by species
```{r}
regen_metrics_subplot %>% 
  select(plot,browsing,Species,matches("Height")) %>% 
  pivot_longer(cols=-c("plot","browsing","Species")) %>% 
  mutate(name=factor(name,levels = c("Height_increment_mn_hclass2","Height_increment_mn_hclass3",
                                     "Height_increment_cv_hclass2","Height_increment_cv_hclass3"))) %>% 
  ggplot(aes(browsing,value,color=Species))+
  geom_point()+
  geom_smooth(method="gam",se=FALSE)+
  facet_wrap(~name,scale="free_y",ncol=5)+
  theme_bw()
```
Not clear. Plot not relevant at the plot scale.


## Difference between plot scale and subplot scale

```{r}


regen_metrics_subplot %>%
  left_join(
    regen_metrics_plot %>% select(plot, H_plot = H, n_seedling_plot = n_seedling),
    by = "plot"
  ) %>%
  ggplot(aes(color = plot)) +
  # Lines connecting each subplot point to its plot-level point
  geom_segment(aes(
    x = n_seedling, y = H,
    xend = n_seedling_plot, yend = H_plot
  ), alpha = 0.4, linewidth = 0.4) +
  # Subplot points (filled circles)
  geom_point(aes(n_seedling, H)) +
  # Plot-level points (triangles)
  geom_point(aes(n_seedling, H_plot), size = 3)

```
# Exploration of radiative budget 

## Radiative budget throughout the day
```{r}
nb_time=metric3d_all %>% select(plot,time) %>% group_by(plot) %>% summarise(ntime=n_distinct(time)) %>% 
  filter(ntime>2) %>% pull(plot)

metric3d_all %>% 
  filter(plot %in% nb_time) %>% 
  ggplot(aes(time,rb_mean,color=as.factor(slice_num)))+
  geom_point()+
  geom_line()+
  facet_grid(plot~subplot)+
  scale_x_continuous(breaks=seq(8, 18, 1))
```
12 or 13 is probably the maximum for rb at the plot scale.
Huge variability though at the subplot scale. For GER11 two peaks corresponding to gaps of light coming into the subplot.


# Link between light distribution and structure

## Between heterogeneity of light and abundance/richness/H/brwosing

```{r}
reg_ext=regen_metrics_plot %>% 
  select(plot,richness,abundance,H,browsing) %>% unique() %>% 
  pivot_longer(cols=-c(plot),names_to = "y",values_to = "y_val")


reg_ext %>% 
  left_join(metric3d_all %>% 
              select(plot,subplot,slice_num,time,rb_mean) %>% 
              filter(slice_num==1) %>% filter(time==12) %>% 
              filter(subplot=="plot"),by=c("plot")) %>% 
  ggplot(aes(rb_mean,y_val,col=plot))+
  geom_point()+geom_smooth(aes(color=NULL),method="lm")+
  scale_x_log10()+
  facet_wrap(~y,scale="free_y",ncol=4)
```

```{r}
regen_metrics_subplot %>% 
  select(plot,subplot,richness,abundance,n_sapling,H,Height_increment_mn_hclass2) %>% 
  group_by(plot,subplot) %>% 
  summarise(across(everything(), \(x) mean(x, na.rm = TRUE)),
            .groups = "drop") %>% 
  pivot_longer(cols=-c("plot","subplot")) %>% 
  mutate(slice_num=NA) %>% 
  filter(plot%in%metric3d_all$plot)->regen_ext
metric3d_all %>% 
  select(plot,subplot,slice_num,rb_mean,rb_sd,rb_cv) %>% 
  filter(slice_num%in%c(1,4)) %>% 
  pivot_longer(cols=c("rb_mean","rb_sd","rb_cv")) %>% 
  bind_rows(regen_ext) %>% 
  mutate(name=factor(name,level=c("rb_mean","rb_sd","rb_cv",
                                  "abundance","n_sapling","richness","H","Height_increment_mn_hclass2"))) %>% 
  ggplot(aes(value,plot))+
  geom_boxplot(aes(color=as.factor(slice_num)))+
  facet_wrap(~name,scale="free_x",ncol=3)+
  theme_bw()+
  labs(color="Height slice",x="Radiative budget metric or Structural metric",y="")



metric3d_all %>% 
  select(plot,subplot,time,slice_num,rb_mean,rb_sd,rb_cv) %>% 
  filter(slice_num%in%c(1,4)) %>% 
  filter(!time%in%c(11,13,15)) %>% 
  left_join(regen_ext %>% 
              pivot_wider(names_from = name,values_from = value) %>% 
              select(-slice_num),by=c("plot","subplot")) %>% 
  ggplot(aes(rb_mean,richness))+
  geom_point(aes(color=plot))+
  geom_smooth(method="gam")+
  facet_grid(time~slice_num,scale="free")+theme_bw()->p
p
```

# ACP regen light - subplot

```{r}
regen_metrics_subplot %>% 
  select(-matches("Height"),-H_hclass3,-H_hclass4,-H_hclass5,-Species) %>% 
  unique() %>% 
  mutate(n_sapling=case_when(is.na(n_sapling)~0,
                             TRUE~n_sapling),
         H_hclass1=case_when(is.na(H_hclass1)~0,
                             TRUE~H_hclass1)) %>% 
  left_join(metric3d_all %>% 
              filter(subplot!="plot")%>% 
              filter(time==12&slice_num==1) %>% 
              select(plot,subplot,h_mean,h_sd,h_cv,rb_mean,rb_sd,rb_cv),
            by=c("plot","subplot")) %>%
  filter(plot%in%unique(metric3d_all$plot)) %>% 
  select(-plot,-subplot)->pca_df

library(factoextra)

pca_df <- pca_df[sapply(pca_df, is.numeric)]
pca_df <- na.omit(pca_df)

# Run PCA (center and scale recommended)
pca_res <- prcomp(pca_df, center = TRUE, scale. = TRUE)

# Summary of variance explained
summary(pca_res)

# Correlation circle (variables factor map)
fviz_pca_var(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE                   # avoid label overlap
)


fviz_pca_var(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE,                   # avoid label overlap
  axes = c(1,3)
)

fviz_eig(pca_res, addlabels = TRUE, ylim = c(0, 50))

```

