---
title: "Fig_paper"
author: "Anne Baranger"
date: "2026-06-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(targets)
# packages needed to *print* the stored ggplot objects
library(ggplot2); library(ggside); library(ggnewscale); library(ggrepel); library(patchwork)

tar_load(c(fig_pca_regen, cor_regen,
           fig_light_plot, cor_light, fig_light_eff,
           fig_light_subplot, fig_light_eff_subplot,
           fig_variability, fig_adult_eff,
           varpart, fig_varpart_venn, fig_varpart_uni, fig_rda_triplot))
```

# Regeneration structure pattern

```{r, fig.width=4, fig.height=5}
fig_pca_regen
```

```{r}
knitr::kable(cor_regen, digits = 3)
```

- hard to distinguish different regeneration clusters
- variable types are not necessarily collinear, can even be uncorrelated
- field-derived and lidar-derived variables roughly uncorrelated, catching different dimensions
- prediction at the subplot scale = adds variability

# Light distribution descriptors - plot level

```{r, fig.width=10, fig.height=5}
fig_light_plot
```

```{r}
knitr::kable(cor_light, digits = 3)
```

# Light distribution effect on regeneration structure - plot level

```{r, fig.width=7.5, fig.height=4.5}
fig_light_eff
```

# From plot to subplot scale

```{r, fig.width=10, fig.height=5}
fig_light_subplot
```

```{r, fig.width=7.5, fig.height=4.5}
fig_light_eff_subplot
```

# Temporal and spatial variability

```{r, fig.width=9, fig.height=3.5}
fig_variability
```

# Confounding effect of adult structure

## Effect size

```{r, fig.width=10, fig.height=5}
fig_adult_eff
```

## Variance partitioning

Plots analysed: n = `r varpart$n`. Excluded (NA): `r if (length(varpart$excluded)) paste(varpart$excluded, collapse = ", ") else "none"`.

```{r}
knitr::kable(round(varpart$vif, 2), col.names = "VIF")
varpart$global_test
knitr::kable(varpart$multi_fractions)
knitr::kable(varpart$uni_fractions, digits = 3)
knitr::kable(varpart$uni_importance, digits = 3)
```

```{r, fig.width=7, fig.height=5}
fig_varpart_venn
```

```{r, fig.width=12, fig.height=5}
fig_varpart_uni
```

```{r, fig.width=7, fig.height=6}
fig_rda_triplot
```
