---
title: "Figures exploration"
author: "Anne Baranger"
date: "2026-05-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE,message=FALSE)
library(targets)
lapply(c("stringr","tidyr","dplyr","tibble","tidyverse", # data analysis
         "ggplot2","viridis", "ggridges","wesanderson","ggridges",# plot
         "factoextra", "nlme","modi","rstan","shinystan","boot",# stat & models
         "rworldmap" # geographic data #terre, sf
         ),require,character.only=TRUE)
theme_set(theme_bw()) # ggplot theme
common_theme <- theme(
  # legend.position = "none",
  strip.background = element_rect(fill = "white",
                                        colour = "gray12"),
  panel.background = element_rect(fill = "white",
                                  colour = "gray12",
                                  size = 0.5, linetype = "solid"),
  plot.title = element_text(size = 12),
  axis.title = element_text(size = 10),
  axis.text = element_text(size = 10),
  strip.text = element_text(size =10),
  plot.margin = unit(c(0, 0.2, 0, 0.2), "cm")  # Example margin; adjust as needed
)

loadRData <- function(fileName){
  #loads an RData file, and returns it
  load(fileName)
  get(ls()[ls() != "fileName"])
}

```

# Data preparation 

## Loading classic data. 

```{r}
tar_load(metric3d_all)
tar_load(regen_metrics_plot)
tar_load(regen_metrics_subplot)
tar_load(plot_df)


metric3d_all %>% 
  select(plot,time) %>% 
  unique() %>% group_by(plot) %>% 
  summarise(`Number of simulated time slots`=n())

metric3d_all %>% 
  select(plot,subplot) %>% 
  unique() %>% group_by(plot) %>% 
  summarise(`Number of subplots`=n()-1)

```


# Exploration of complexity measures 

```{r,fig.height=10,fig.width=10}

metric3d_all_plot=metric3d_all %>% 
  filter(subplot=="plot") %>% 
  pivot_longer(cols =  c(matches("h_"),"occupied_gf",matches("pad")))

metric3d_all %>% 
  filter(subplot!="plot") %>% 
  pivot_longer(cols = c(matches("h_"),"occupied_gf",matches("pad"))) %>% 
  filter(!grepl("_na",name)) %>% 
  ggplot(aes(slice_num,value,color=plot))+
  geom_line(aes(group=interaction(plot,subplot)),alpha = 0.4)+
  geom_point(data=metric3d_all_plot,aes(slice_num,value,color=plot))+
  geom_line(data=metric3d_all_plot,aes(slice_num,value,color=plot,group=interaction(plot,subplot)))+
  facet_grid(name~plot,scale="free_y")+
  theme_bw()


# metric3d_all_plot
metric3d_all %>% 
  filter(subplot!="plot") %>% 
  filter(time==12) %>% 
  ggplot(aes(h_mean,h_sd,color=time))+
  geom_point(aes(group=interaction(time,subplot)),alpha = 0.4)+
  geom_line(aes(group=interaction(time,subplot)),alpha = 0.4)+
  geom_point(data=metric3d_all %>% filter(subplot=="plot") %>% filter(time==12),
             aes(h_mean,h_sd,color=time))+
  geom_line(data=metric3d_all %>% filter(subplot=="plot") %>% filter(time==12),
            aes(h_mean,h_sd,color=time,group=interaction(time,subplot)))+
  facet_wrap(~plot,scale="free_y")+
  theme_bw()
```

Main outcomes :
* measure height without accounting for empty areas doesn't make sense. Drop variables "_na"
* median doesn't make sense when accounting for empty areas, drop "h_med"
* "coef_lin" is not informative
* The shape of h_mean=f(slice_num) seems to provide insight on the understory structure. Linear form are more common for plots with less regeneration, while saturating form could be more common for dense understory

## Rank according to abondance

```{r}
metric3d_all_plot=metric3d_all %>% 
  filter(subplot=="plot")

regen_metrics_plot$all %>% 
  select(plot,n_sapling) %>% 
  mutate(n_sapling=replace_values(n_sapling, NA ~ 0)) %>% 
  unique() %>% 
  arrange(n_sapling) %>% 
  pull(plot)->plot_arrange
  
metric3d_all %>% 
  filter(subplot != "plot") %>% 
  mutate(plot = factor(plot, levels = plot_arrange)) %>% 
  ggplot(aes(slice_num, occupied_gf, color = plot)) +
  geom_line(aes(group = interaction(plot, subplot)), alpha = 0.4) +
  geom_point(
    data = metric3d_all_plot %>% 
      mutate(plot = factor(plot, levels = plot_arrange)),
    aes(slice_num, occupied_gf, color = plot)
  ) +
  geom_line(
    data = metric3d_all_plot %>% 
      mutate(plot = factor(plot, levels = plot_arrange)),
    aes(slice_num, occupied_gf, color = plot, group = interaction(plot, subplot))
  ) +
  facet_grid( ~ plot, scales = "free_y") +
  theme_bw()





regen_metrics_subplot$all %>% 
  mutate(plot_sub=paste0(plot,"_",substr(subplot,9,9))) %>% 
  select(plot_sub,n_sapling) %>%
  mutate(n_sapling=replace_values(n_sapling, NA ~ 0)) %>% 
  unique() %>% 
  arrange(n_sapling) ->subplot_arrange


metric3d_all %>% 
  filter(subplot != "plot") %>% 
  mutate(plot_sub = paste0(plot, "_", substr(subplot, 9, 9))) %>% 
  mutate(plot_sub = factor(plot_sub, levels = subplot_arrange$plot_sub)) %>% 
  left_join(subplot_arrange) %>% 
  ggplot(aes(slice_num, occupied_gf,
             group = plot_sub,
             color = n_sapling)) +
  geom_line(size = 0.9, alpha = 0.9) +
  scale_color_viridis_c(
    option = "plasma",
    trans = "log10"
  ) +
  theme_bw()


df_plot <- metric3d_all %>% 
  filter(subplot != "plot") %>% 
  mutate(
    plot_sub = paste0(plot, "_", substr(subplot, 9, 9)),
    plot_sub = factor(plot_sub, levels = subplot_arrange$plot_sub)
  ) %>% 
  left_join(subplot_arrange) %>% 
  mutate(
    abundance_cat = ntile(n_sapling, 5),
    abundance_cat = factor(
      abundance_cat,
      levels = 1:5
    )
  )

ggplot(df_plot, aes(slice_num, occupied_gf, group = plot_sub, color = n_sapling)) +
  geom_line(size = 1.1, alpha = 0.9) +
  scale_color_viridis_c(
    option = "plasma",
    trans = "log10"
  ) +
  facet_wrap(~ abundance_cat,ncol=5) +
  theme_bw()
```

# Exploration of field metrics

## relevance of seedling number
```{r}
regen_metrics_subplot$all %>% 
  ggplot(aes(abundance,n_seedling))+
  geom_point()+
  geom_abline(slope=1)+
  labs(x="Total abundance",
       y="Seedlings number (class 1 and 2)",
       title="Subplot level")

regen_metrics_plot$all %>% 
  ggplot(aes(abundance,n_seedling))+
  geom_point()+
  geom_abline(slope=1)+
  labs(x="Total abundance",
       y="Seedlings number (class 1 and 2)",
       title="Plot level")
```
Seedling number is highly correlated to total abundance at both subplot and plot level.

## relevance of sapling number
```{r}
regen_metrics_subplot$all %>% 
  ggplot(aes(abundance,n_sapling))+
  geom_point()+
  geom_smooth(method="lm")+
  labs(x="Total abundance",
       y="Saplings number (class 1 and 2)",
       title="Subplot level")

regen_metrics_plot$all %>% 
  ggplot(aes(abundance,n_sapling))+
  geom_point()+
  geom_smooth(method="lm")+
  labs(x="Total abundance",
       y="Saplings number (class 1 and 2)",
       title="Plot level")
```


## Browsing metrics
```{r}
regen_metrics_plot$all %>% 
  pivot_longer(cols=-c("plot","browsing")) %>% 
  mutate(name=factor(name,levels = c("abundance","abundance_cv","n_seedling","richness","richness_sd",
                                     "Height_increment_mn_hclass2","Height_increment_mn_hclass3",
                                     "Height_increment_cv_hclass2","Height_increment_cv_hclass3",
                                     "H",
                                     "H_hclass1","H_hclass2","H_hclass3","H_hclass4","H_hclass5"))) %>% 
  ggplot(aes(browsing,value))+
  geom_point()+
  geom_smooth(method="lm")+
  facet_wrap(~name,scale="free_y",ncol=5)+
  theme_bw()+
  labs(title = "At the plot scale")



regen_metrics_subplot$all %>% 
  pivot_longer(cols=-c("plot","subplot","browsing")) %>% 
  mutate(name=factor(name,levels = c("abundance","abundance_cv","n_seedling","richness","richness_sd",
                                     "Height_increment_mn_hclass2","Height_increment_mn_hclass3",
                                     "Height_increment_cv_hclass2","Height_increment_cv_hclass3",
                                     "H",
                                     "H_hclass1","H_hclass2","H_hclass3","H_hclass4","H_hclass5"))) %>% 
  ggplot(aes(browsing,value))+
  geom_point()+
  geom_smooth(method="lm")+
  facet_wrap(~name,scale="free_y",ncol=5)+
  theme_bw()+
  labs(title = "At the subplot scale")
```
Browsing reduces height increment, probably differs between species -> test it
It probably reduces abundance, but this is uncertain. 
Linked to higher richness and to higher sd in richness, at the plot scale.

### by species
```{r}
regen_metrics_subplot$species %>% 
  select(plot,browsing,Species,matches("Height")) %>% 
  pivot_longer(cols=-c("plot","browsing","Species")) %>% 
  mutate(name=factor(name,levels = c("Height_increment_mn","Height_increment_cv",
                                     "Height_increment_mn_hclass2","Height_increment_mn_hclass3",
                                     "Height_increment_cv_hclass2","Height_increment_cv_hclass3"))) %>% 
  ggplot(aes(browsing,value,color=Species))+
  geom_point()+
  geom_smooth(method="gam",se=FALSE)+
  facet_wrap(~name,scale="free_y",ncol=2)+
  theme_bw()
```
Not clear. Plot not relevant at the plot scale.


## Difference between plot scale and subplot scale

```{r}


regen_metrics_subplot$all %>%
  left_join(
    regen_metrics_plot$all %>% select(plot, H_plot = H, n_seedling_plot = n_seedling),
    by = "plot"
  ) %>%
  ggplot(aes(color = plot)) +
  # Lines connecting each subplot point to its plot-level point
  geom_segment(aes(
    x = n_seedling, y = H,
    xend = n_seedling_plot, yend = H_plot
  ), alpha = 0.4, linewidth = 0.4) +
  # Subplot points (filled circles)
  geom_point(aes(n_seedling, H)) +
  # Plot-level points (triangles)
  geom_point(aes(n_seedling, H_plot), size = 3)

```

## Link between richness and abundance

```{r}
regen_metrics_plot$all %>% 
  ggplot(aes(n_sapling,richness))+
  geom_point()+
  geom_smooth(method="gam")
```


# Exploration of radiative budget 

## Radiative budget throughout the day
```{r}
nb_time=metric3d_all %>% select(plot,time) %>% group_by(plot) %>% summarise(ntime=n_distinct(time)) %>% 
  filter(ntime>2) %>% pull(plot)

metric3d_all %>% 
  filter(plot %in% nb_time) %>% 
  ggplot(aes(time,rb_mean,color=as.factor(slice_num)))+
  geom_point()+
  geom_line()+
  facet_grid(plot~subplot)+
  scale_x_continuous(breaks=seq(8, 18, 1))
```
12 or 13 is probably the maximum for rb at the plot scale.
Huge variability though at the subplot scale. For GER11 two peaks corresponding to gaps of light coming into the subplot.


# Link between light distribution and structure

## Between heterogeneity of light and abundance/richness/H/brwosing

```{r}
reg_ext=regen_metrics_plot %>% 
  select(plot,richness,abundance,H,browsing) %>% unique() %>% 
  pivot_longer(cols=-c(plot),names_to = "y",values_to = "y_val")


reg_ext %>% 
  left_join(metric3d_all %>% 
              select(plot,subplot,slice_num,time,rb_mean) %>% 
              filter(slice_num==1) %>% filter(time==12) %>% 
              filter(subplot=="plot"),by=c("plot")) %>% 
  ggplot(aes(rb_mean,y_val,col=plot))+
  geom_point()+geom_smooth(aes(color=NULL),method="lm")+
  scale_x_log10()+
  facet_wrap(~y,scale="free_y",ncol=4)
```

```{r}
regen_metrics_subplot$all %>% 
  select(plot,subplot,richness,abundance,n_sapling,H,Height_increment_mn_hclass2) %>% 
  group_by(plot,subplot) %>% 
  summarise(across(everything(), \(x) mean(x, na.rm = TRUE)),
            .groups = "drop") %>% 
  pivot_longer(cols=-c("plot","subplot")) %>% 
  mutate(slice_num=NA) %>% 
  filter(plot%in%metric3d_all$plot)->regen_ext
metric3d_all %>% 
  select(plot,subplot,slice_num,rb_mean,rb_sd,rb_cv) %>% 
  filter(slice_num%in%c(1,4)) %>% 
  pivot_longer(cols=c("rb_mean","rb_sd","rb_cv")) %>% 
  bind_rows(regen_ext) %>% 
  mutate(name=factor(name,level=c("rb_mean","rb_sd","rb_cv",
                                  "abundance","n_sapling","richness","H","Height_increment_mn_hclass2"))) %>% 
  ggplot(aes(value,plot))+
  geom_boxplot(aes(color=as.factor(slice_num)))+
  facet_wrap(~name,scale="free_x",ncol=3)+
  theme_bw()+
  labs(color="Height slice",x="Radiative budget metric or Structural metric",y="")



metric3d_all %>% 
  select(plot,subplot,time,slice_num,rb_mean,rb_sd,rb_cv) %>% 
  filter(slice_num%in%c(1,4)) %>% 
  filter(!time%in%c(11,13,15)) %>% 
  left_join(regen_ext %>% 
              pivot_wider(names_from = name,values_from = value) %>% 
              select(-slice_num),by=c("plot","subplot")) %>% 
  ggplot(aes(rb_mean,richness))+
  geom_point(aes(color=plot))+
  geom_smooth(method="gam")+
  facet_grid(time~slice_num,scale="free")+theme_bw()->p
p
```

# ACP regen light

## subplot

```{r}
regen_metrics_subplot %>% 
  select(-matches("Height"),-H_hclass3,-H_hclass4,-H_hclass5,-Species) %>% 
  unique() %>% 
  mutate(n_sapling=case_when(is.na(n_sapling)~0,
                             TRUE~n_sapling),
         H_hclass1=case_when(is.na(H_hclass1)~0,
                             TRUE~H_hclass1)) %>% 
  left_join(metric3d_all %>% 
              filter(subplot!="plot")%>% 
              filter(time==12&slice_num==1) %>% 
              select(plot,subplot,h_mean,h_sd,h_cv,rb_mean,rb_sd,rb_cv),
            by=c("plot","subplot")) %>%
  filter(plot%in%unique(metric3d_all$plot)) %>% 
  select(-plot,-subplot)->pca_df

library(factoextra)

pca_df <- pca_df[sapply(pca_df, is.numeric)]
pca_df <- na.omit(pca_df)

# Run PCA (center and scale recommended)
pca_res <- prcomp(pca_df[,1:10], center = TRUE, scale. = TRUE)

# Summary of variance explained
summary(pca_res)

# Correlation circle (variables factor map)
fviz_pca_var(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE                   # avoid label overlap
)

fviz_pca_biplot(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE                   # avoid label overlap
)


fviz_pca_var(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE,                   # avoid label overlap
  axes = c(1,3)
)

fviz_eig(pca_res, addlabels = TRUE, ylim = c(0, 50))

```
## Plot
```{r}
regen_metrics_plot %>% 
  select(-matches("Height"),-H_hclass3,-H_hclass4,-H_hclass5,-Species) %>% 
  unique() %>% 
  mutate(n_sapling=case_when(is.na(n_sapling)~0,
                             TRUE~n_sapling),
         H_hclass1=case_when(is.na(H_hclass1)~0,
                             TRUE~H_hclass1)) %>% 
  left_join(metric3d_all %>% 
              filter(subplot=="plot")%>% 
              filter(time==12&slice_num==1) %>% 
              select(plot,subplot,h_mean,h_sd,h_cv,rb_mean,rb_sd,rb_cv),
            by=c("plot")) %>%
  filter(plot%in%unique(metric3d_all$plot)) %>% 
  select(-plot)->pca_df

library(factoextra)

pca_df <- pca_df[sapply(pca_df, is.numeric)]
pca_df <- na.omit(pca_df)

# Run PCA (center and scale recommended)
pca_res <- prcomp(pca_df, center = TRUE, scale. = TRUE)

# Summary of variance explained
summary(pca_res)

# Correlation circle (variables factor map)
fviz_pca_var(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE                   # avoid label overlap
)

fviz_pca_biplot(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE                   # avoid label overlap
)


fviz_pca_var(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE,                   # avoid label overlap
  axes = c(1,3)
)

fviz_eig(pca_res, addlabels = TRUE, ylim = c(0, 50))

```


# Canopy species

```{r}

plot_df %>% 
  left_join(regen_metrics_plot %>% 
              select(plot,richness,abundance,H) %>% 
              unique())%>% 
  dplyr::select(-cutting_rate,-recruitment_rate,-n_cut,-cutting_rate,-n_recruits)->a


pca_df <-sapply(a[,2:22], as.numeric)
pca_df <- na.omit(pca_df) 

# Run PCA (center and scale recommended)
pca_res <- prcomp(pca_df, center = TRUE, scale. = TRUE)

# Summary of variance explained
summary(pca_res)

# Correlation circle (variables factor map)
fviz_pca_var(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE                   # avoid label overlap
)


fviz_pca_var(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE,                   # avoid label overlap
  axes = c(1,3)
)

fviz_eig(pca_res, addlabels = TRUE, ylim = c(0, 50))
```


## Effect of adults diversity on light heterogeneity / quantity

```{r}
plot_df %>% 
  select(plot,n_trees_alive,species_richness_alive) %>% 
  left_join(metric3d_all_plot %>% 
              filter(time==12,slice_num==1)%>% 
              select(plot,rb_mean,rb_sd,rb_cv) ) %>% 
  ggplot(aes(species_richness_alive,rb_sd))+
  geom_point()+
  geom_smooth(method="lm")


plot_df %>% 
  select(plot,n_trees_alive,species_richness_alive) %>% 
  left_join(regen_metrics_plot$all %>% 
              select(plot,richness,richness_seedling,richness_sapling) ) %>% 
  ggplot(aes(species_richness_alive,richness_sapling))+
  geom_point()+
  geom_smooth(method="lm")


plot_df %>% 
  select(plot,species_richness_alive) %>% 
  left_join(regen_metrics_plot$all %>% 
              select(plot,richness)) %>% 
  left_join(metric3d_all_plot %>%
              filter(time==12,slice_num==1)%>% 
              select(plot,rb_sd,rb_mean,rb_cv))%>%
  mutate(rb_sd=log(rb_sd),
         rb_mean=log(rb_mean)) %>%
  mutate(across(c(species_richness_alive, richness, rb_sd, rb_mean), scale))->sem_df

library(piecewiseSEM)
library(semPlot)
library(lavaan)

lavaan_model <- '
  richness ~ species_richness_alive + rb_sd + rb_mean
  rb_sd    ~ species_richness_alive
  rb_mean  ~ species_richness_alive
'

fit_lav <- sem(lavaan_model, data = sem_df)
summary(fit_lav)
# Plot
semPaths(
  object         = fit_lav,
  what           = "std",
  whatLabels     = "std",
  layout         = "tree2",
  edge.label.cex = 1.2,
  node.label.cex = 1.2,
  residuals      = FALSE,
  intercepts     = FALSE
)




# 1. Extract coefficients from piecewiseSEM
coef_table <- coefs(fit_lav, standardize = "scale")

# 2. Add significance stars
coef_table$label <- with(coef_table, ifelse(P.Value < 0.001, paste0(round(Std.Estimate, 2), "***"),
                                    ifelse(P.Value < 0.01,  paste0(round(Std.Estimate, 2), "**"),
                                    ifelse(P.Value < 0.05,  paste0(round(Std.Estimate, 2), "*"),
                                    ifelse(P.Value < 0.1,   paste0(round(Std.Estimate, 2), "."),
                                                             round(Std.Estimate, 2))))))

# 3. Plot with custom labels
semPaths(
  object         = fit_lav,
  what           = "std",
  whatLabels     = "std",
  layout         = "tree2",
  edge.label.cex = 1.2,
  node.label.cex = 1.2,
  residuals      = FALSE,
  intercepts     = FALSE,
  edgeLabels     = coef_table$label  # inject piecewiseSEM labels with stars
)

regen_metrics_plot$all %>%
  select(plot,plot,richness) %>% 
  left_join(plot_df %>% 
              select(plot,species_richness_alive)) %>% 
  left_join(metric3d_all %>%
              filter(subplot=="plot") %>% 
              filter(time==12,slice_num==1)%>% 
              select(plot,subplot,rb_sd,rb_mean,rb_cv))%>%
  mutate(rb_sd=log(rb_sd),
         rb_mean=log(rb_mean)) %>%
  mutate(across(c(species_richness_alive, richness, rb_sd, rb_mean), scale))->sem_df
summary(lm(richness~rb_sd,data=sem_df))

regen_metrics_plot$all %>% 
  select(plot,plot,
         richness,richness_sd,richness_seedling,richness_sapling,
         abundance,abundance_sd,abundance_cv,
         Height_increment_mn,Height_increment_cv,
         H,H_height,browsing) %>% unique() %>% 
  pivot_longer(cols=-c(plot),names_to = "y",values_to = "y_val") %>% 
  left_join(metric3d_all %>% 
              select(plot,subplot,slice_num,time,rb_sd,rb_mean,occupied_gf) %>% 
              # filter(slice_num==1) %>% 
              filter(time==12) %>% 
              filter(subplot=="plot"),by=c("plot")) %>% 
  filter(plot%in%metric3d_all$plot) %>% 
  ggplot(aes(rb_mean,y_val,col=plot))+
  geom_point()+geom_smooth(aes(color=NULL),method="lm")+
  scale_x_log10()+
  facet_wrap(slice_num~y,scale="free_y",nrow=4)


```



# First hypothesis 
We hypothesised that greater spatial heterogeneity of light is associated with greater species diversity of regeneration (species richness, Shannon index), independently of mean light quantity.

## check collinearity

```{r}
metric3d_all %>% 
              select(plot,subplot,slice_num,time,rb_cv,rb_mean,rb_sd) %>% 
              # filter(slice_num==1) %>% 
              filter(time==12) %>% 
              filter(subplot=="plot")->pca_df

# Run PCA (center and scale recommended)
pca_res <- prcomp(pca_df[,5:7], center = TRUE, scale. = TRUE)

# Summary of variance explained
summary(pca_res)

# Correlation circle (variables factor map)
fviz_pca_var(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE                   # avoid label overlap
)
```
## coefficient of rb_cv slope
```{r}
# ── 1. Données longues ────────────────────────────────────────────────────────
data <- regen_metrics_plot$all %>%
  select(plot,
         richness, richness_sd, richness_seedling, richness_sapling,
         H, H_hclass1, H_hclass2, H_hclass3, H_hclass4, H_hclass5,
         H_height) %>%
  left_join(
    metric3d_all %>%
      select(plot, subplot, slice_num, time, rb_sd, rb_mean, rb_cv, pad_sd) %>%
      filter(time == 12, subplot == "plot"),
    by = "plot"
  ) %>%
  filter(plot %in% metric3d_all$plot) %>%
  pivot_longer(
    cols = -c(plot, subplot, slice_num, time, rb_sd, rb_mean, rb_cv),
    names_to  = "y",
    values_to = "y_val"
  ) %>% 
  mutate(slice_num=as.factor(slice_num)) %>% 
  group_by(y) %>% 
  mutate(y_val=scale(y_val))

vars <- c("richness", "richness_sd", "richness_seedling", "richness_sapling",
          "H", "H_hclass1", "H_hclass2", "H_hclass3", "H_hclass4", "H_hclass5",
          "H_height", "pad_sd")

# ── 2. Fonction bootstrap ─────────────────────────────────────────────────────
# Modèle : y_val ~ rb_cv * slice_num
# coef(fit) retourne :
#   [1] intercept
#   [2] rb_cv              ← pente "principale" de rb_cv (à slice_num = 0)
#   [3] slice_num
#   [4] rb_cv:slice_num    ← comment la pente de rb_cv change avec slice_num

boot_fn <- function(data, indices) {
  d   <- data[indices, ]
  fit <- lm(y_val ~ slice_num + slice_num:rb_cv, data = d)
  coef(fit)
}

# ── 3. Bootstrap par variable ────────────────────────────────────────────────
boot_results <- map_dfr(vars, function(v) {

  data_lm <- data %>%
    filter(y == v) %>%
    drop_na(y_val, rb_cv, slice_num) %>%
    droplevels()

  if (nrow(data_lm) < 10) return(NULL)

  # strata = slice_num : chaque rééchantillon garde tous les niveaux
  # -> nombre de coefficients stable d'un tirage à l'autre
  boot_res <- boot(data = data_lm, statistic = boot_fn, R = 1000,
                   strata = data_lm$slice_num)

  cn        <- names(boot_res$t0)
  slope_idx <- which(grepl(":rb_cv$", cn))                 # pentes par niveau
  lvls      <- sub("^slice_num", "", sub(":rb_cv$", "", cn[slope_idx]))

  map_dfr(seq_along(slope_idx), function(i) {
    tibble(
      var       = v,
      slice_num = lvls[i],
      slope     = boot_res$t[, slope_idx[i]]
    )
  })
})

# ── 4. Plot ridgeline ─────────────────────────────────────────────────────────
slice_levels <- levels(data$slice_num)
n_vars       <- length(vars)
n_slices     <- length(slice_levels)

boot_results %>%
  mutate(
    slice_num  = factor(slice_num, levels = slice_levels),
    var_idx    = as.numeric(factor(var, levels = vars)),
    slice_code = as.numeric(slice_num),                    # 1..n_slices
    slice_pos  = var_idx +
      (slice_code - mean(seq_len(n_slices))) / max(n_slices - 1, 1) * 0.8
  ) %>%
  ggplot(aes(x = slope, y = slice_pos, group = interaction(var, slice_num),
             fill = slice_num)) +
  stat_density_ridges(
    quantile_lines = TRUE, quantiles = c(0.05, 0.5, 0.95),
    alpha = 0.6, color = "white", linewidth = 0.3,
    scale = 2
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_y_continuous(breaks = 1:n_vars, labels = vars) +
  scale_fill_viridis_d(option = "mako", name = "Slice height") +  # _d : discret
  labs(x = "Slope of rb_cv (bootstrap)", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.y      = element_text(face = "bold")
  )
```
-> comm : run the analysis for group of variables (richness, shannon index, pad_sd) to avoid scaling 


## subplpot scale


```{r}
# ── 1. Données longues ────────────────────────────────────────────────────────
data <- regen_metrics_subplot$all %>%
  select(plot,subplot,
         richness, richness_sd, richness_seedling, richness_sapling,
         H, H_hclass1, H_hclass2, H_hclass3, H_hclass4, H_hclass5,
         H_height) %>%
  left_join(
    metric3d_all %>%
      select(plot, subplot, slice_num, time, rb_sd, rb_mean, rb_cv, pad_sd) %>%
      filter(time == 12, subplot != "plot"),
    by = c("plot","subplot")
  ) %>%
  filter(plot %in% metric3d_all$plot) %>%
  mutate(slice_num=as.factor(slice_num),
         plot=as.factor(plot)) %>% 
  pivot_longer(
    cols = -c(plot, subplot, slice_num, time, rb_sd, rb_mean, rb_cv),
    names_to  = "y",
    values_to = "y_val"
  ) %>% 
  group_by(y) %>% 
  mutate(y_val=scale(y_val)) %>% ungroup()

vars <- c("richness", "richness_sd", "richness_seedling", "richness_sapling",
          "H", "H_hclass1", "H_hclass2", "H_hclass3", "H_hclass4", "H_hclass5",
          "H_height", "pad_sd")

# ── 2. Fonction bootstrap ─────────────────────────────────────────────────────
# Modèle : y_val ~ rb_cv * slice_num
# coef(fit) retourne :
#   [1] intercept
#   [2] rb_cv              ← pente "principale" de rb_cv (à slice_num = 0)
#   [3] slice_num
#   [4] rb_cv:slice_num    ← comment la pente de rb_cv change avec slice_num

boot_fn <- function(data, indices) {
  d   <- data[indices, ]
  fit <- lm(y_val ~ rb_cv * slice_num, data = d)
  coef(fit)
}

# ── 3. Bootstrap par variable × slice_num ────────────────────────────────────
# On bootstrappe une seule fois par variable (slice_num est continu dans le modèle).
# On récupère ensuite la pente effective de rb_cv à chaque valeur de slice_num :
#   slope(rb_cv | slice_num = s) = beta_rb_cv + beta_rb_cv:slice_num * s

slices <- sort(unique(data$slice_num))   # valeurs observées de slice_num

boot_results <- map_dfr(vars, function(v) {
  
  data_lm <- data %>%
    filter(y == v) %>%          # pad_sd est dans y comme les autres
    drop_na(y_val, rb_cv, slice_num)
  
  if (nrow(data_lm) < 10) return(NULL)
  
  boot_res <- boot(data = data_lm, statistic = boot_fn, R = 1000)
  
  beta_rb_cv       <- boot_res$t[, 2]
  beta_rb_cv_slice <- boot_res$t[, 4]
  
  map_dfr(slices, function(s) {
    tibble(
      var       = v,
      slice_num = s,
      slope     = beta_rb_cv + beta_rb_cv_slice * s
    )
  })
})

# ── 4. Plot ridgeline ─────────────────────────────────────────────────────────
# Créer des positions numériques
n_vars   <- length(vars)
n_slices <- length(slices)

boot_results %>%
  mutate(
    var_idx   = as.numeric(factor(var, levels = vars)),
    slice_pos = var_idx + (slice_num - mean(slices)) /
                  (max(slices) - min(slices)) * 0.8
  ) %>%
  ggplot(aes(x = slope, y = slice_pos, group = interaction(var, slice_num),
             fill = slice_num)) +
  stat_density_ridges(
    quantile_lines = TRUE, quantiles = c(0.05, 0.5, 0.95),
    alpha = 0.6, color = "white", size = 0.3,
    scale = 2
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_y_continuous(
    breaks = 1:n_vars,
    labels = vars
  ) +
  scale_fill_viridis_c(option = "mako", name = "Slice height") +
  labs(x = "Slope of rb_cv (bootstrap)", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.y      = element_text(face = "bold")
  )
```


# Second hypothesis :  comparison of coefficients between sapling and seedling
This heterogeneity–diversity relationship was expected to be stronger for younger regeneration stages, both because older individuals have likely experienced different light conditions over time, and because they are subject to additional, potentially stronger filters such as competition and browsing 


```{r}
# Les 3 variables à comparer
vars_compare <- c("richness", "richness_seedling", "richness_sapling")

# Extraire les distributions bootstrap pour ces 3 variables
boot_compare <- boot_results %>%
  filter(var %in% vars_compare)

# ── Toutes les paires ─────────────────────────────────────────────────────────
pairs <- combn(vars_compare, 2, simplify = FALSE)

diff_results <- map_dfr(pairs, function(pair) {
  map_dfr(slices, function(s) {
    
    slopes_a <- boot_compare %>% filter(var == pair[1], slice_num == s) %>% pull(slope)
    slopes_b <- boot_compare %>% filter(var == pair[2], slice_num == s) %>% pull(slope)
    
    # Différence des distributions bootstrap
    diff <- slopes_a - slopes_b
    
    tibble(
      var_a     = pair[1],
      var_b     = pair[2],
      pair      = paste(pair[1], "−", pair[2]),
      slice_num = s,
      diff_mean = mean(diff),
      diff_lo   = quantile(diff, 0.025),
      diff_hi   = quantile(diff, 0.975),
      diff_p    = mean(diff > 0),          # P(diff > 0)
      sig       = diff_lo > 0 | diff_hi < 0  # IC ne contient pas 0
    )
  })
})

# ── Plot ──────────────────────────────────────────────────────────────────────
diff_results %>%
  ggplot(aes(x = slice_num, y = diff_mean, color = sig)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_ribbon(aes(ymin = diff_lo, ymax = diff_hi, fill = sig), 
              alpha = 0.2, color = NA) +
  geom_line(size = 0.8) +
  geom_point(size = 2) +
  scale_color_manual(values = c("FALSE" = "grey50", "TRUE" = "#E84855"),
                     name = "IC 95% ≠ 0") +
  scale_fill_manual(values  = c("FALSE" = "grey50", "TRUE" = "#E84855"),
                    name = "IC 95% ≠ 0") +
  facet_wrap(~ pair, ncol = 1) +
  labs(
    x     = "slice_num (hauteur LiDAR)",
    y     = "Différence de pente rb_cv (A − B)",
    title = "Comparaison des pentes rb_cv entre variables de richesse",
    subtitle = "Rouge = IC 95% bootstrap ne contient pas 0"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text       = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# ── Table résumé ──────────────────────────────────────────────────────────────
diff_results %>%
  select(pair, slice_num, diff_mean, diff_lo, diff_hi, diff_p, sig) %>%
  mutate(across(where(is.numeric), round, 3)) %>%
  print(n = Inf)
```


# Third hypothesis 

We also hypothesized that, independently of heterogeneity, mean light quantity is positively associated with regeneration abundance and biomass 


```{r}
data <- regen_metrics_plot$all %>%
  select(plot,
         abundance, abundance_sd,n_sapling,
         Height_increment_mn_hclass2,Height_increment_mn_hclass3) %>%
  left_join(
    metric3d_all %>%
      select(plot, subplot, slice_num, time, rb_sd, rb_mean, rb_cv,
             pad_tot,pad_mean,occupied_gf) %>%
      filter(time == 12, subplot == "plot"),
    by = "plot"
  ) %>%
  filter(plot %in% metric3d_all$plot) %>%
  mutate(rb_mean=log(rb_mean)) %>% 
  pivot_longer(
    cols = -c(plot, subplot, slice_num, time, rb_sd, rb_mean, rb_cv),
    names_to  = "y",
    values_to = "y_val"
  ) %>% 
  group_by(y) %>% 
  mutate(y_val=scale(y_val))

# data %>% 
#   ggplot(aes(rb_mean,y_val,col=plot))+
#   geom_point()+geom_smooth(aes(color=NULL),method="lm")+
#   scale_x_log10()+
#   facet_wrap(slice_num~y,scale="free_y",nrow=4)


vars <- c("abundance", "abundance_sd", "n_sapling", "Height_increment_mn_hclass2",
          "Height_increment_mn_hclass3", "pad_tot", "pad_mean", "occupied_gf")

# ── 2. Fonction bootstrap ─────────────────────────────────────────────────────
# Modèle : y_val ~ rb_cv * slice_num
# coef(fit) retourne :
#   [1] intercept
#   [2] rb_cv              ← pente "principale" de rb_cv (à slice_num = 0)
#   [3] slice_num
#   [4] rb_cv:slice_num    ← comment la pente de rb_cv change avec slice_num

boot_fn <- function(data, indices) {
  d   <- data[indices, ]
  fit <- lm(y_val ~ rb_mean * slice_num, data = d)
  coef(fit)
}

# ── 3. Bootstrap par variable × slice_num ────────────────────────────────────
# On bootstrappe une seule fois par variable (slice_num est continu dans le modèle).
# On récupère ensuite la pente effective de rb_cv à chaque valeur de slice_num :
#   slope(rb_cv | slice_num = s) = beta_rb_cv + beta_rb_cv:slice_num * s

slices <- sort(unique(data$slice_num))   # valeurs observées de slice_num

boot_results <- map_dfr(vars, function(v) {
  
  data_lm <- data %>%
    filter(y == v) %>%          # pad_sd est dans y comme les autres
    drop_na(y_val, rb_mean, slice_num)
  
  if (nrow(data_lm) < 10) return(NULL)
  
  boot_res <- boot(data = data_lm, statistic = boot_fn, R = 1000)
  
  beta_rb_mean       <- boot_res$t[, 2]
  beta_rb_mean_slice <- boot_res$t[, 4]
  
  map_dfr(slices, function(s) {
    tibble(
      var       = v,
      slice_num = s,
      slope     = beta_rb_mean + beta_rb_mean * s
    )
  })
})

# ── 4. Plot ridgeline ─────────────────────────────────────────────────────────
# Créer des positions numériques
n_vars   <- length(vars)
n_slices <- length(slices)


boot_results %>%
  mutate(
    var_idx   = as.numeric(factor(var, levels = vars)),
    slice_pos = var_idx + (slice_num - mean(slices)) /
                  (max(slices) - min(slices)) * 0.8
  ) %>%
  ggplot(aes(x = slope, y = slice_pos, group = interaction(var, slice_num),
             fill = slice_num)) +
  stat_density_ridges(
    quantile_lines = TRUE, quantiles = c(0.05, 0.5, 0.95),
    alpha = 0.6, color = "white", size = 0.3,
    scale = 2
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_y_continuous(
    breaks = 1:n_vars,
    labels = vars
  ) +
  scale_fill_viridis_c(option = "mako", name = "Hauteur\n(slice_num)") +
  labs(x = "Pente de rb_cv (bootstrap)", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.y      = element_text(face = "bold")
  )

```

effect on abundance and and gap_fraction

# Fourth hypothesis
we tested the interaction between light heterogeneity and quantity, predicting that at low mean light levels, heterogeneity has a relatively stronger effect on regeneration diversity and structure
```{r}
data <- regen_metrics_plot$all %>%
  select(plot,
         abundance,richness,richness_seedling) %>%
  left_join(
    metric3d_all %>%
      select(plot, subplot, slice_num, time, rb_sd, rb_mean, rb_cv,
             pad_tot,pad_mean,occupied_gf,pad_sd) %>%
      filter(time == 12, subplot == "plot"),
    by = "plot"
  ) %>%
  filter(plot %in% metric3d_all$plot) %>%
  mutate(rb_mean=log(rb_mean)) %>% 
  pivot_longer(
    cols = -c(plot, subplot, slice_num, time, rb_sd, rb_mean, rb_cv),
    names_to  = "y",
    values_to = "y_val"
  ) %>% 
  group_by(y) %>% 
  mutate(y_val=scale(y_val)) %>% 
  filter(slice_num==1)


data_model <- data %>% ungroup() %>% 
  distinct(plot, rb_cv, rb_mean, rb_sd) %>%
  mutate(
    rb_mean_c = scale(rb_mean, center = TRUE, scale = TRUE)[,1],  # centré réduit
    rb_cv_c   = scale(rb_cv,   center = TRUE, scale = TRUE)[,1]
  )

cat("Corrélation rb_cv ~ rb_mean :", cor(data_model$rb_cv, data_model$rb_mean), "\n")

# ── 2. Joindre aux variables de diversité ─────────────────────────────────────
data_int <- data %>%
  filter(slice_num == 1) %>%    # un slice pour simplifier
  select(plot, y, y_val) %>%
  left_join(data_model, by = "plot")



vars <- c("abundance","richness", "richness_seedling", "pad_sd","occupied_gf")

# ── 3. Bootstrap avec interaction rb_cv × rb_mean ─────────────────────────────
# Modèle : y ~ rb_cv_c * rb_mean_c
# coef :
#   [1] intercept
#   [2] rb_cv_c           ← effet de l'hétérogénéité à rb_mean moyen
#   [3] rb_mean_c         ← effet de la quantité à rb_cv moyen
#   [4] rb_cv_c:rb_mean_c ← interaction : est-ce que l'effet de rb_cv 
#                           change selon rb_mean ? (hypothèse ici)

boot_fn_int <- function(data, indices) {
  d   <- data[indices, ]
  fit <- lm(y_val ~ rb_cv_c * rb_mean_c, data = d)
  coef(fit)
}

boot_int_results <- map_dfr(vars, function(v) {
  
  data_lm <- data_int %>%
    filter(y == v) %>%
    drop_na(y_val, rb_cv_c, rb_mean_c)
  
  if (nrow(data_lm) < 10) return(NULL)
  
  boot_res <- boot(data = data_lm, statistic = boot_fn_int, R = 1000)
  
  tibble(
    var          = v,
    # Effet rb_cv à rb_mean moyen
    slope_rb_cv  = boot_res$t[, 2],
    # Interaction : signe négatif = rb_cv plus fort quand rb_mean faible ✓ hypothèse
    slope_int    = boot_res$t[, 4]
  )
})

# ── 4. Résumé statistique ─────────────────────────────────────────────────────
summary_int <- boot_int_results %>%
  group_by(var) %>%
  summarise(
    # Effet principal rb_cv
    rb_cv_mean = mean(slope_rb_cv),
    rb_cv_lo   = quantile(slope_rb_cv, 0.025),
    rb_cv_hi   = quantile(slope_rb_cv, 0.975),
    rb_cv_sig  = rb_cv_lo > 0 | rb_cv_hi < 0,
    # Interaction
    int_mean   = mean(slope_int),
    int_lo     = quantile(slope_int, 0.025),
    int_hi     = quantile(slope_int, 0.975),
    int_sig    = int_lo > 0 | int_hi < 0,
    # P(interaction dans le sens de l'hypothèse) = P(coef < 0)
    # rb_mean faible + rb_cv élevé → interaction négative attendue
    p_hyp      = mean(slope_int < 0)
  ) %>%
  mutate(across(where(is.numeric), round, 3))

print(summary_int)

# ── 5. Plot : distribution de l'interaction par variable ──────────────────────
# Calculer le sens de l'interaction par variable pour le fill
fill_colors <- boot_int_results %>%
  group_by(var) %>%
  summarise(median_int = median(slope_int)) %>%
  mutate(hyp_confirmed = median_int < 0)

boot_int_results %>%
  mutate(var = factor(var, levels = vars)) %>%
  left_join(fill_colors, by = "var") %>%
  ggplot(aes(x = slope_int, y = fct_rev(var), fill = hyp_confirmed)) +
  stat_density_ridges(
    quantile_lines = TRUE, quantiles = c(0.025, 0.5, 0.975),
    alpha = 0.75, color = "white", size = 0.3,
    scale = 1.2
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey30") +
  scale_fill_manual(
    values = c("TRUE" = "#4A90D9", "FALSE" = "#E84855"),
    labels = c("TRUE" = "Hypothèse confirmée\n(rb_cv + fort si rb_mean faible)",
               "FALSE" = "Sens inverse"),
    name = NULL
  ) +
  labs(
    x        = "Coefficient d'interaction rb_cv × rb_mean",
    y        = NULL,
    title    = "Test de l'interaction lumière hétérogénéité × quantité",
    subtitle = "Bleu = interaction négative (confirme l'hypothèse)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.y      = element_text(face = "bold"),
    legend.position  = "bottom"
  )
```
# Fifth hypothesis
We hypothesised that canopy diversity influences regeneration diversity through two independent, quantifiable pathways: (i) by modulating the spatial heterogeneity of the light environment (H5.1), and (ii) by modulating seed availability through the presence of different seed producing trees (H5.2).

```{r}
data <- regen_metrics_plot$all %>%
  select(plot,
         richness,browsing) %>%
  left_join(
    metric3d_all %>%
      select(plot, subplot, slice_num, time, rb_sd, rb_mean, rb_cv) %>%
      filter(time == 12, subplot == "plot"),
    by = "plot"
  ) %>%
  left_join(plot_df %>% 
              select(plot,n_trees_alive,species_richness_alive,species_richness_dominant)) %>% 
  filter(plot %in% metric3d_all$plot) %>% 
  filter(slice_num==1) %>% 
  mutate(rb_sd=log(rb_sd))

summary(lm(richness~rb_sd+species_richness_alive+browsing,data=data))
anova(lm(richness~rb_sd+species_richness_alive+browsing,data=data))
Anova(lm(richness~rb_sd+species_richness_alive+browsing,data=data))

metric3d_all %>%
  select(plot, subplot, slice_num, time, rb_sd, rb_mean, rb_cv) %>%
  filter(time == 12, subplot == "plot",slice_num==1)%>%
  left_join(plot_df ) %>% 
  pivot_longer(cols=-c(plot,subplot,slice_num,time,rb_sd,rb_mean,rb_cv)) %>% 
  ggplot(aes(log(rb_mean),value))+
  geom_point()+
  geom_smooth(method="gam")+
  facet_wrap(~name,scale="free_y")
  
```
 #todo

-> posthoc analysis of the structure of seedling/sapling community - regarding their shade tolerance = compute community wieghted mean?
-> correlation with seed mass ? heavy seeds likely more present under the cover of their parents

Plot effect in subplot analysis
Compute survival between height class!
test rb_sd ~ abundance in different height class
compute mortality rate of parents
