---
title: "Fig_paper"
author: "Anne Baranger"
date: "2026-06-24"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE,message=FALSE)
library(targets)
lapply(c("stringr","tidyr","dplyr","tibble","tidyverse", # data analysis
         "ggplot2","viridis", "ggridges","wesanderson","ggridges","ggrepel",# plot
         "factoextra", "nlme","modi","rstan","shinystan","boot",# stat & models
         "rworldmap" # geographic data #terre, sf
         ),require,character.only=TRUE)
theme_set(theme_bw()) # ggplot theme
common_theme <- theme(
  # legend.position = "none",
  strip.background = element_rect(fill = "white",
                                        colour = "gray12"),
  panel.background = element_rect(fill = "white",
                                  colour = "gray12",
                                  size = 0.5, linetype = "solid"),
  plot.title = element_text(size = 12),
  axis.title = element_text(size = 10),
  axis.text = element_text(size = 10),
  strip.text = element_text(size =10),
  plot.margin = unit(c(0, 0.2, 0, 0.2), "cm")  # Example margin; adjust as needed
)

loadRData <- function(fileName){
  #loads an RData file, and returns it
  load(fileName)
  get(ls()[ls() != "fileName"])
}

```

# Data preparation 

```{r}
tar_load(metric3d_all)
tar_load(regen_metrics_plot)
tar_load(regen_metrics_subplot)
tar_load(plot_df)
tar_load(H_ext_all)

```

# Regeneration structure pattern


```{r}

### Plot scale ###
##################
regen_metrics_plot$all %>% 
  mutate(richness_cv=richness_sd/richness,
         richness_sapling=if_else(is.na(richness_sapling),0,richness_sapling)) %>% 
  select(plot,richness, H, richness_sapling, # diversity // richness_cv,
         abundance, # quantity // abundance_cv,
         Height_increment_mn, # dynamics // Height_increment_cv,
         H_height,
         browsing) %>% # complexity 
  left_join(metric3d_all %>% 
              filter(subplot=="plot",time==12) %>% 
              select(plot,slice_num,
                     h_mean,h_cv, # quantity and complexity 
                     empty_gf,pad_tot,pad_cv)) %>% 
  filter(slice_num==1) %>% 
  select(-slice_num)-> pca_data_plot


pca_df <-sapply(pca_data_plot[,2:dim(pca_data_plot)[2]], as.numeric)
pca_df <- na.omit(pca_df) 

# Run PCA (center and scale recommended)
pca_res <- prcomp(pca_df, center = TRUE, scale. = TRUE)

# Summary of variance explained
# summary(pca_res)

# Correlation circle (variables factor map)
# fviz_pca_biplot(
#   pca_res,
#   # col.var = "contrib",            # color by contribution
#   # gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
#   repel = TRUE               
# ) -> a 

### SUBplot scale ###
##################
regen_metrics_subplot$all %>%
  mutate(richness_cv=richness_sd/richness,
         richness_sapling=if_else(is.na(richness_sapling),0,richness_sapling)) %>%
  select(plot,subplot,richness, H, richness_sapling, # diversity // richness_cv,
         abundance, # quantity // abundance_cv,
         Height_increment_mn, # dynamics // Height_increment_cv,
         H_height,
         browsing) %>% # complexity
  left_join(metric3d_all %>%
              filter(subplot!="plot",time==12) %>%
              select(plot,subplot,slice_num,
                     h_mean,h_cv, # quantity and complexity
                     empty_gf,pad_tot,pad_cv)) %>%
  filter(slice_num==1) %>%
  select(-slice_num)-> pca_data_subplot
# 
# 
# pca_df <-sapply(pca_data_subplot[,3:dim(pca_data_subplot)[2]], as.numeric)
# pca_df <- na.omit(pca_df) 
# 
# # Run PCA (center and scale recommended)
# pca_res <- prcomp(pca_df, center = TRUE, scale. = TRUE)
# 
# # Summary of variance explained
# summary(pca_res)
# 
# # Correlation circle (variables factor map)
# fviz_pca_var(
#   pca_res,
#   col.var = "contrib",            # color by contribution
#   gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
#   repel = TRUE               
# ) -> b
# 
# 
# cowplot::plot_grid(a,b)


bind_rows(pca_data_plot,pca_data_subplot)-> pca_tot
pca_tot = cbind(pca_tot,predict(pca_res,pca_tot[,2:13])[,c(1,2)])


var <- get_pca_var(pca_res)

var_coord <- as.data.frame(var$coord[, 1:2])
colnames(var_coord) <- c("PC1", "PC2")

var_coord$variable <- rownames(var_coord)
var_coord$contrib <- var$contrib[, 1] + var$contrib[, 2]
var_coord$cos2 <- var$cos2[, 1] + var$cos2[, 2]

var_coord=var_coord %>% 
  mutate(type=case_when(variable%in%c("richness","richness_sapling","H")~"Composition",
                        variable%in%c("abundance","pad_tot","h_mean","Height_increment_mn","empty_gf")~"Quantity",
                        variable%in%c("H_height","h_cv","pad_cv")~"Complexity",
                        TRUE~"Other"),
         variable=case_when(variable=="richness"~"Richness \ntotal",
                            variable=="richness_sapling"~"Richness \nsaplings",
                            variable=="H"~"Shanon \nindex",
                            variable=="abundance"~"Abundance",
                            variable=="Height_increment_mn"~"Height \nincrement",
                            variable=="H_height"~"Vertical \ndistribution",
                            variable=="Browsing"~"Browsing",
                            variable=="h_cv"~"Height \nvariability",
                            variable=="h_mean"~"Mean \nheight",
                            variable=="empty_gf"~"Gap fraction",
                            variable=="pad_tot"~"PAD \ntotal",
                            variable=="pad_cv"~"PAD \nvariability",))


ggplot() +
  geom_point(
    data = pca_tot %>% 
      mutate(
        plot_level = case_when(
          is.na(subplot) ~ "Plot",
          !is.na(subplot) ~ "Subplot"
        )
      ),
    aes(
      PC1, PC2,
      shape = plot_level,
      size = plot_level
    )
  ) +
  scale_shape_manual(
    name = "Spatial scale",
    values = c(
      "Plot" = 19,
      "Subplot" = 1
    )
  ) +
  scale_size_manual(
    name = "Spatial scale",
    values = c(
      "Plot" = 2.5,
      "Subplot" = 1.5
    )
  ) +
  geom_segment(
    data = var_coord,
    aes(
      x = 0, y = 0,
      xend = PC1*5, yend = PC2*5,
      alpha = contrib,
      color=type
    ),
    arrow = arrow(length = unit(0.20, "cm")),
    lineend = "round"
  ) +
  geom_text_repel(
    data = var_coord,
    aes(
      x = PC1*5,
      y = PC2*5,
      label = variable,
      alpha = contrib,
      color=type
    ),
    size = 3,
    show.legend = FALSE
  ) +
  coord_equal() +
  theme_minimal()+
  guides(alpha = "none")+
  labs(x=paste0("PCA axis 1 (",
                round(summary(pca_res)$importance["Proportion of Variance","PC1"]*100,
                      digits = 1),
                "% of variation)"),
       y=paste0("PCA axis 1 (",
                round(summary(pca_res)$importance["Proportion of Variance","PC2"]*100,
                      digits = 1),
                "% of variation)"),
       color="Variable type")
  


```

- hard to distinguish different regeneration clusters
- variable type are not necessarly colinear, can even be non correlated
- field-derived and lidar-derived variables roughly non correlated, catching different dimensions
- prediction at the subplot scale = add variability

# Dimensions of light distribution controlling regeneration structure and composition


```{r}
metric3d_all %>% 
  filter(subplot=="plot",time==12) %>% 
  filter(slice_num==1) %>% 
  select(plot,
         rb_mean,rb_cv) %>% 
  left_join(H_ext_all %>% 
              filter(extinction==0.5,subplot=="plot",time==12)) %>% 
  rename("Radiative\nbudget mean"=rb_mean, "Radiative\nbudget var."=rb_cv,
      "Half-ext. \nheight"=mean, "Half-ext.\nheight (SD)"=sd,
      "Extinction\nspeed"=mean_slope, "Extinction\nspeed (SD)"=sd_slope)->light_pca




light_pca <-sapply(light_pca[,c("Radiative\nbudget mean", "Radiative\nbudget var.",
      "Half-ext. \nheight", "Half-ext.\nheight (SD)",
      "Extinction\nspeed", "Extinction\nspeed (SD)")], as.numeric)
light_pca <- na.omit(light_pca) 

# Run PCA (center and scale recommended)
pca_res <- prcomp(light_pca, center = TRUE, scale. = TRUE)

fviz_pca_biplot(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE
) 
```


## plot

```{r}
pca_data_plot %>% 
  left_join(metric3d_all %>% 
              filter(subplot=="plot",time==12) %>% 
              filter(slice_num==1) %>% 
              select(plot,
                     rb_mean,rb_cv)) %>% 
  left_join(H_ext_all %>% 
              filter(extinction==0.5,subplot=="plot",time==12)) %>% 
  mutate(rb_mean=log(rb_mean))->light_test


explained <- c("richness","richness_sapling","H","abundance","pad_tot","h_mean","Height_increment_mn","empty_gf","H_height","h_cv","pad_cv")
explaining <- c("rb_mean","rb_cv","mean","sd","mean_slope","sd_slope")



std_coef <- function(d, y, x, covar = "browsing") {
  d <- d %>%
    select(all_of(c(y, x, covar))) %>%
    drop_na() %>%
    mutate(across(where(is.numeric), ~ as.numeric(scale(.))))  # factors (e.g. browsing) left alone
  f <- reformulate(c(covar, x), response = y)
  unname(coef(lm(f, data = d))[x])
}

# --- bootstrap distribution of that coef ---
boot_coef <- function(d, y, x, covar = "browsing", B = 1000) {
  d <- d %>% select(all_of(c(y, x, covar))) %>% drop_na() %>%
      mutate(across(where(is.numeric), ~ as.numeric(scale(.)))) 
  n <- nrow(d)
  f <- reformulate(c(covar, x), response = y)
  map_dbl(seq_len(B), function(b) {
    db <- d[sample(n, n, replace = TRUE), ]  # <- move this line out of the loop to scale once
    unname(coef(lm(f, data = db))[x])
  })
}

results <- expand_grid(explained = explained, explaining = explaining) %>%
  mutate(
    estimate = map2_dbl(explained, explaining, ~ std_coef(light_test, .x, .y)),
    draws    = map2(explained, explaining, ~ boot_coef(light_test, .x, .y, B = 1000))
  )


boot_long <- results %>%
  select(explained, explaining, draws) %>%
  unnest_longer(draws) %>%
  rename(coef = draws)

boot_long %>%
  group_by(explained, explaining) %>%
  summarise(
    lo  = unname(quantile(coef, 0.1)),
    med = unname(quantile(coef, 0.5)),
    hi  = unname(quantile(coef, 0.9)),
    .groups = "drop"
  ) %>%
  mutate(
    type = factor(case_when(
      explained %in% c("richness","richness_sapling","H") ~ "Composition",
      explained %in% c("abundance","pad_tot","h_mean","Height_increment_mn","empty_gf") ~ "Quantity",
      explained %in% c("H_height","h_cv","pad_cv") ~ "Complexity"
    ), levels = c("Composition","Quantity","Complexity")),
    excl0 = lo > 0 | hi < 0,                         # interval excludes zero
    explained = recode(explained,
      richness="Richness\ntotal", richness_sapling="Richness\nsaplings",
      H="Shannon\nindex", abundance="Abundance", pad_tot="PAD\ntotal",
      h_mean="Mean\nheight", Height_increment_mn="Height\nincrement",
      empty_gf="Gap fraction", H_height="Vertical\ndistribution",
      h_cv="Height\nvariability", pad_cv="PAD\nvariability"),
    # explained = factor(explained, levels = rev(explained_order)),
    explaining = recode(explaining,
      rb_mean="Radiative\nbudget mean", rb_cv="Radiative\nbudget var.",
      mean="Half-ext. \nheight", sd="Half-ext.\nheight (SD)",
      mean_slope="Extinction\nspeed", sd_slope="Extinction\nspeed (SD)")
  ) %>% 
  ggplot( aes(x = med, y = explained)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = .3, colour = "grey70") +
  geom_linerange(aes(xmin = lo, xmax = hi, colour = excl0), linewidth = .9) +
  geom_point(aes(colour = excl0), size = 1.3) +
  scale_colour_manual(
    values = c(`TRUE` = "#1f5fa8", `FALSE` = "grey78")
  ) +
  guides(colour="none")+
  facet_grid(type ~ explaining, scales = "free_y", space = "free_y") +
  labs(x = "Standardized coefficient", y = NULL) +
  theme_bw(base_size = 9) +
  theme(
    panel.spacing.x = unit(.6, "lines"),
    strip.text.y    = element_text(angle = 0),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )

```
##subplot

```{r}
pca_data_subplot %>% 
  left_join(metric3d_all %>% 
              filter(subplot!="plot",time==12) %>% 
              filter(slice_num==1) %>% 
              select(plot,subplot,
                     rb_mean,rb_cv)) %>% 
  left_join(H_ext_all %>% 
              filter(extinction==0.5,subplot!="plot",time==12))->light_test_sub


explained <- c("richness","richness_sapling","H","abundance","pad_tot","h_mean","Height_increment_mn","empty_gf","H_height","h_cv","pad_cv")
explaining <- c("rb_mean","rb_cv","mean","sd","mean_slope","sd_slope")


std_coef <- function(d, y, x, covar = "browsing") {
  d <- d %>%
    select(all_of(c(y, x, covar, "plot"))) %>%
    drop_na() %>%
    mutate(plot = factor(plot)) %>%                              # grouping factor, kept out of scaling
    mutate(across(where(is.numeric), ~ as.numeric(scale(.))))    # factors (browsing, plot) left alone
  f <- as.formula(paste0(y, " ~ ", covar, " + ", x, " + (1 | plot)"))
  m <- lmer(f, data = d, control = lmerControl(calc.derivs = FALSE))
  unname(fixef(m)[x])
}

# --- bootstrap distribution of that coef ---
boot_coef <- function(d, y, x, covar = "browsing", B = 1000) {
  d <- d %>% select(all_of(c(y, x, covar, "plot"))) %>% drop_na() %>%
      mutate(plot = factor(plot)) %>%
      mutate(across(where(is.numeric), ~ as.numeric(scale(.))))
  n <- nrow(d)
  f <- as.formula(paste0(y, " ~ ", covar, " + ", x, " + (1 | plot)"))
  map_dbl(seq_len(B), function(b) {
    db <- d[sample(n, n, replace = TRUE), ]
    m  <- suppressMessages(suppressWarnings(
            lmer(f, data = db, control = lmerControl(calc.derivs = FALSE))))
    unname(fixef(m)[x])
  })
}


results <- expand_grid(explained = explained, explaining = explaining) %>%
  mutate(
    estimate = map2_dbl(explained, explaining, ~ std_coef(light_test_sub, .x, .y)),
    draws    = map2(explained, explaining, ~ boot_coef(light_test_sub, .x, .y, B = 1000))
  )


boot_long <- results %>%
  select(explained, explaining, draws) %>%
  unnest_longer(draws) %>%
  rename(coef = draws)

boot_long %>%
  group_by(explained, explaining) %>%
  summarise(
    lo  = unname(quantile(coef, 0.05)),
    med = unname(quantile(coef, 0.5)),
    hi  = unname(quantile(coef, 0.95)),
    .groups = "drop"
  ) %>%
  mutate(
    type = factor(case_when(
      explained %in% c("richness","richness_sapling","H") ~ "Composition",
      explained %in% c("abundance","pad_tot","h_mean","Height_increment_mn","empty_gf") ~ "Quantity",
      explained %in% c("H_height","h_cv","pad_cv") ~ "Complexity"
    ), levels = c("Composition","Quantity","Complexity")),
    excl0 = lo > 0 | hi < 0,                         # interval excludes zero
    explained = recode(explained,
      richness="Richness\ntotal", richness_sapling="Richness\nsaplings",
      H="Shannon\nindex", abundance="Abundance", pad_tot="PAD\ntotal",
      h_mean="Mean\nheight", Height_increment_mn="Height\nincrement",
      empty_gf="Gap fraction", H_height="Vertical\ndistribution",
      h_cv="Height\nvariability", pad_cv="PAD\nvariability"),
    # explained = factor(explained, levels = rev(explained_order)),
    explaining = recode(explaining,
      rb_mean="Radiative\nbudget mean", rb_cv="Radiative\nbudget var.",
      mean="Half-ext. \nheight", sd="Half-ext.\nheight (SD)",
      mean_slope="Extinction\nspeed", sd_slope="Extinction\nspeed (SD)")
  ) %>% 
  ggplot( aes(x = med, y = explained)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = .3, colour = "grey70") +
  geom_linerange(aes(xmin = lo, xmax = hi, colour = excl0), linewidth = .9) +
  geom_point(aes(colour = excl0), size = 1.3) +
  scale_colour_manual(
    values = c(`TRUE` = "#1f5fa8", `FALSE` = "grey78")
  ) +
  guides(colour="none")+
  facet_grid(type ~ explaining, scales = "free_y", space = "free_y") +
  labs(x = "Standardized coefficient", y = NULL) +
  theme_bw(base_size = 9) +
  theme(
    panel.spacing.x = unit(.6, "lines"),
    strip.text.y    = element_text(angle = 0),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )
```

# effect of adults on light distribution and regeneration structure


```{r}
sem = plot_df %>% 
  select(plot,n_trees_alive,species_richness_alive,mean_dbh_alive_cm,
         mean_height_alive_m,mortality_rate) %>% 
  left_join(light_test)


explained <- c("richness","richness_sapling","H","abundance","pad_tot","h_mean","Height_increment_mn","empty_gf","H_height","h_cv","pad_cv")
explaining <- c("n_trees_alive","species_richness_alive","mean_dbh_alive_cm","mean_height_alive_m","mortality_rate")



std_coef <- function(d, y, x, covar = "browsing") {
  d <- d %>%
    select(all_of(c(y, x, covar))) %>%
    drop_na() %>%
    mutate(across(where(is.numeric), ~ as.numeric(scale(.))))  # factors (e.g. browsing) left alone
  f <- reformulate(c(covar, x), response = y)
  unname(coef(lm(f, data = d))[x])
}

# --- bootstrap distribution of that coef ---
boot_coef <- function(d, y, x, covar = "browsing", B = 1000) {
  d <- d %>% select(all_of(c(y, x, covar))) %>% drop_na() %>%
      mutate(across(where(is.numeric), ~ as.numeric(scale(.)))) 
  n <- nrow(d)
  f <- reformulate(c(covar, x), response = y)
  map_dbl(seq_len(B), function(b) {
    db <- d[sample(n, n, replace = TRUE), ]  # <- move this line out of the loop to scale once
    unname(coef(lm(f, data = db))[x])
  })
}

results <- expand_grid(explained = explained, explaining = explaining) %>%
  mutate(
    estimate = map2_dbl(explained, explaining, ~ std_coef(sem, .x, .y)),
    draws    = map2(explained, explaining, ~ boot_coef(sem, .x, .y, B = 1000))
  )


boot_long <- results %>%
  select(explained, explaining, draws) %>%
  unnest_longer(draws) %>%
  rename(coef = draws)

boot_long %>%
  group_by(explained, explaining) %>%
  summarise(
    lo  = unname(quantile(coef, 0.1)),
    med = unname(quantile(coef, 0.5)),
    hi  = unname(quantile(coef, 0.9)),
    .groups = "drop"
  ) %>%
  mutate(
    type = factor(case_when(
      explained %in% c("richness","richness_sapling","H") ~ "Composition",
      explained %in% c("abundance","pad_tot","h_mean","Height_increment_mn","empty_gf") ~ "Quantity",
      explained %in% c("H_height","h_cv","pad_cv") ~ "Complexity"
    ), levels = c("Composition","Quantity","Complexity")),
    excl0 = lo > 0 | hi < 0,                         # interval excludes zero
    explained = recode(explained,
      richness="Richness\ntotal", richness_sapling="Richness\nsaplings",
      H="Shannon\nindex", abundance="Abundance", pad_tot="PAD\ntotal",
      h_mean="Mean\nheight", Height_increment_mn="Height\nincrement",
      empty_gf="Gap fraction", H_height="Vertical\ndistribution",
      h_cv="Height\nvariability", pad_cv="PAD\nvariability"),
    # explained = factor(explained, levels = rev(explained_order)),
    explaining = recode(explaining,
      rb_mean="Radiative\nbudget mean", rb_cv="Radiative\nbudget var.",
      mean="Half-ext. \nheight", sd="Half-ext.\nheight (SD)",
      mean_slope="Extinction\nspeed", sd_slope="Extinction\nspeed (SD)")
  ) %>% 
  ggplot( aes(x = med, y = explained)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = .3, colour = "grey70") +
  geom_linerange(aes(xmin = lo, xmax = hi, colour = excl0), linewidth = .9) +
  geom_point(aes(colour = excl0), size = 1.3) +
  scale_colour_manual(
    values = c(`TRUE` = "#1f5fa8", `FALSE` = "grey78")
  ) +
  guides(colour="none")+
  facet_grid(type ~ explaining, scales = "free_y", space = "free_y") +
  labs(x = "Standardized coefficient", y = NULL) +
  theme_bw(base_size = 9) +
  theme(
    panel.spacing.x = unit(.6, "lines"),
    strip.text.y    = element_text(angle = 0),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )

```

```{r}
sem = plot_df %>% 
  select(plot,n_trees_alive,species_richness_alive,mean_dbh_alive_cm,
         mean_height_alive_m,mortality_rate) %>% 
  left_join(light_test)



library(piecewiseSEM)
library(semPlot)
library(lavaan)

lavaan_model <- '
  richness ~ mean + rb_cv + n_trees_alive+ species_richness_alive+ mean_dbh_alive_cm+
         mean_height_alive_m+mortality_rate
  rb_cv    ~ n_trees_alive+ species_richness_alive+ mean_dbh_alive_cm+
         mean_height_alive_m+mortality_rate
  rb_mean  ~ n_trees_alive+ species_richness_alive+ mean_dbh_alive_cm+
         mean_height_alive_m+mortality_rate
'

fit_lav <- sem(lavaan_model, data = sem)
summary(fit_lav)
# Plot
semPaths(
  object         = fit_lav,
  what           = "std",
  whatLabels     = "std",
  layout         = "tree2",
  edge.label.cex = 1.2,
  node.label.cex = 1.2,
  residuals      = FALSE,
  intercepts     = FALSE
)




# 1. Extract coefficients from piecewiseSEM
coef_table <- coefs(fit_lav, standardize = "scale")

# 2. Add significance stars
coef_table$label <- with(coef_table, ifelse(P.Value < 0.001, paste0(round(Std.Estimate, 2), "***"),
                                    ifelse(P.Value < 0.01,  paste0(round(Std.Estimate, 2), "**"),
                                    ifelse(P.Value < 0.05,  paste0(round(Std.Estimate, 2), "*"),
                                    ifelse(P.Value < 0.1,   paste0(round(Std.Estimate, 2), "."),
                                                             round(Std.Estimate, 2))))))

# 3. Plot with custom labels
semPaths(
  object         = fit_lav,
  what           = "std",
  whatLabels     = "std",
  layout         = "tree2",
  edge.label.cex = 1.2,
  node.label.cex = 1.2,
  residuals      = FALSE,
  intercepts     = FALSE,
  edgeLabels     = coef_table$label  # inject piecewiseSEM labels with stars
)

```

