---
title: "Fig_paper"
author: "Anne Baranger"
date: "2026-06-24"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE,message=FALSE)
library(targets)
lapply(c("stringr","tidyr","dplyr","tibble","tidyverse", # data analysis
         "ggplot2","viridis", "ggridges","wesanderson","ggridges","ggrepel","ggnewscale","ggside", # plot
         "factoextra", "nlme","modi","rstan","shinystan","boot", "rstatix","lme4",# stat & models
         "rworldmap" # geographic data #terre, sf
         ),require,character.only=TRUE)
theme_set(theme_bw()) # ggplot theme
common_theme <- theme(
  # legend.position = "none",
  strip.background = element_rect(fill = "white",
                                        colour = "gray12"),
  panel.background = element_rect(fill = "white",
                                  colour = "gray12",
                                  size = 0.5, linetype = "solid"),
  plot.title = element_text(size = 12),
  axis.title = element_text(size = 10),
  axis.text = element_text(size = 10),
  strip.text = element_text(size =10),
  plot.margin = unit(c(0, 0.2, 0, 0.2), "cm")  # Example margin; adjust as needed
)

loadRData <- function(fileName){
  #loads an RData file, and returns it
  load(fileName)
  get(ls()[ls() != "fileName"])
}

```

# Data preparation 

```{r}
tar_load(metric3d_all)
tar_load(metric3d_all_unsliced)
tar_load(metric3d_all_seg)
tar_load(regen_metrics_plot)
tar_load(regen_metrics_subplot)
tar_load(plot_df)
tar_load(H_ext_all)
tar_load(regen_cat)
tar_load(regen_cat_subplot)
```

# Regeneration structure pattern


```{r, fig.width=10, fig.height=6}
prep_pca_data <- function(regen, id_cols, plot_level) {
  regen %>%
    mutate(across(c(richness_sapling, n_sapling), ~ replace_na(.x, 0))) %>%
    select(all_of(id_cols),
           richness, H, richness_sapling, cwm_shade,   # composition
           abundance, n_sapling,                       # quantity
           Height_increment_mn,abundance_cv,                        # dynamics 
           H_height, browsing) %>%                     # complexity
    inner_join(
      metric3d_all_unsliced %>%
        filter(if (plot_level) subplot == "plot" else subplot != "plot",time == 12) %>% #,
               # , slice_num == 1) %>%
        select(all_of(id_cols), empty_gf, pad_tot, pad_cv, ri,fhd), #h_mean, h_cv,
      by = id_cols
    )
}




pca_data_plot    <- prep_pca_data(regen_metrics_plot$all,    "plot",              TRUE)
pca_data_subplot <- prep_pca_data(regen_metrics_subplot$all, c("plot","subplot"), FALSE)


pca_df   <- pca_data_plot %>% select(-plot) %>% mutate(across(everything(), as.numeric)) %>% drop_na()
pca_res  <- prcomp(pca_df, center = TRUE, scale. = TRUE)

pca_tot <- bind_rows(pca_data_plot, pca_data_subplot) %>%
  left_join(regen_cat) %>%
  bind_cols(as_tibble(predict(pca_res, .[, colnames(pca_df)])[, 1:2]))


v <- get_pca_var(pca_res)
var_coord <- tibble(
  variable = rownames(v$coord),
  PC1      = v$coord[, 1],
  PC2      = v$coord[, 2],
  contrib  = rowSums(v$contrib[, 1:2]),
  cos2     = rowSums(v$cos2[, 1:2])
) %>%
  left_join(var_meta, by = "variable") %>%
  mutate(type = replace_na(type, "Other"))

pts <- pca_tot %>%
  filter(is.na(subplot)) %>%
  mutate(plot_level = "Plot")

ggplot() +
  geom_point(
    data = pts,
    aes(PC1, PC2, shape = plot_level, size = plot_level, color = reg_category)
  ) +
  geom_xsideboxplot(
    data = pts %>% filter(!is.na(reg_category)),
    aes(x = PC1, y = reg_category, fill = reg_category),
    orientation = "y", outliers = FALSE
  ) +
  geom_ysideboxplot(
    data = pts %>% filter(!is.na(reg_category)),
    aes(y = PC2, x = reg_category, fill = reg_category),
    orientation = "x", outliers = FALSE
  )+
  scale_fill_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  scale_shape_manual(name = "Spatial scale", values = c("Plot" = 19, "Subplot" = 1)) +
  scale_size_manual(name = "Spatial scale", values = c("Plot" = 2.5, "Subplot" = 1.5)) +
  scale_color_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  guides(color="none",size="none",shape="none")+
  new_scale_color() +
  geom_segment(
    data = var_coord,
    aes(x = 0, y = 0, xend = PC1*5, yend = PC2*5, color = type),
    arrow = arrow(length = unit(0.20, "cm")), lineend = "round"
  ) +
  geom_text_repel(
    data = var_coord,
    aes(x = PC1*5, y = PC2*5, label = label, color = type),
    size = 3, show.legend = FALSE
  ) +
  scale_color_brewer(name = "Variable type", palette = "Dark2") +
  theme_minimal() +
  # shrink the side panels + hide their axes
  theme(
    # --- main panel: black frame ---
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    
    # --- side panels: small ---
    ggside.panel.scale = 0.15,          # smaller = thinner strips (try 0.10–0.20)
    
    # --- side panels: no grid ---
    ggside.panel.grid = element_blank(),
    
    # --- side panels: no frame/border of their own, hide their axes ---
    ggside.panel.border = element_blank(),
    ggside.panel.background = element_blank(),
    ggside.axis.text  = element_blank(),
    ggside.axis.ticks = element_blank(),
    
  legend.position = "bottom"
  )+
  labs(
    x = paste0("PCA axis 1 (", round(summary(pca_res)$importance["Proportion of Variance","PC1"]*100, 1), "% of variation)"),
    y = paste0("PCA axis 2 (", round(summary(pca_res)$importance["Proportion of Variance","PC2"]*100, 1), "% of variation)")
  ) + 
  coord_cartesian()->c
c
```


- hard to distinguish different regeneration clusters
- variable type are not necessarly colinear, can even be non correlated
- field-derived and lidar-derived variables roughly non correlated, catching different dimensions
- prediction at the subplot scale = add variability

# Dimensions of light distribution controlling regeneration structure and composition

## plot 

### Between plots trend

```{r}
metric3d_all_unsliced %>% 
  filter(subplot=="plot",time==12) %>% 
  select(plot,
         rb_mean,rb_cv) %>% 
  left_join(H_ext_all %>% 
              filter(target==0.9,subplot=="plot",time==12)) %>% 
  mutate(rb_mean=log(rb_mean)) %>% 
  left_join(regen_cat)->light_pca

df_pca=light_pca[,c("rb_mean","z_cross_agg","ext_ref_agg","padTot_agg",
                  "padAbove_agg","zPAD50_agg")]#"rb_ref_agg","padAbove_cv","rb_cv",

pca_res <- prcomp(df_pca,
                  center = TRUE, scale. = TRUE)

light_pca_pred = cbind(light_pca,predict(pca_res,df_pca)[,c(1,2)])


var <- get_pca_var(pca_res)

var_coord <- as.data.frame(var$coord[, 1:2])
colnames(var_coord) <- c("PC1", "PC2")

var_coord$variable <- rownames(var_coord)
var_coord$contrib <- var$contrib[, 1] + var$contrib[, 2]
var_coord$cos2 <- var$cos2[, 1] + var$cos2[, 2]


var_coord=var_coord %>%
  left_join(var_meta,by="variable") %>% 
  mutate(label=case_when(variable=="slope_cross_agg"~paste0(label," 0.9"),
                         variable=="z_cross_agg"~paste0(label," 0.9"),
                         TRUE~label))

ggplot() +
  geom_point(
    data = light_pca_pred,
    aes(PC1, PC2, color = reg_category),
    size=2.5,show.legend = FALSE
  ) +
  scale_color_brewer(name = "Regeneration cover", palette = "YlGn",na.value = "grey") +
  
  new_scale_color() +
  
  geom_segment(
    data = var_coord,
    aes(x = 0, y = 0, xend = PC1*5, yend = PC2*5,  color = type), #alpha = contrib,
    arrow = arrow(length = unit(0.20, "cm")),
    lineend = "round"
  ) +
  geom_text_repel(
    data = var_coord,
    aes(x = PC1*5, y = PC2*5, label = label,  color = type), #alpha = contrib,
    size = 3, show.legend = FALSE
  ) +
  geom_xsideboxplot(
    data = light_pca_pred %>% filter(!is.na(reg_category)),
    aes(x = PC1, y = reg_category, fill = reg_category),
    orientation = "y", outliers = FALSE
  ) +
  geom_ysideboxplot(
    data = light_pca_pred %>% filter(!is.na(reg_category)),
    aes(y = PC2, x = reg_category, fill = reg_category),
    orientation = "x", outliers = FALSE
  )+
  scale_fill_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  scale_color_brewer(name = "Variable type", palette = "Dark2") +
  guides(color="none")+
  
  coord_equal() +
  theme_minimal() +
  guides(
    alpha = "none",
    color = guide_legend(override.aes = list(alpha = 1))
  ) +
  labs(
    x = paste0("PCA axis 1 (", round(summary(pca_res)$importance["Proportion of Variance","PC1"]*100, 1), "% of variation)"),
    y = paste0("PCA axis 2 (", round(summary(pca_res)$importance["Proportion of Variance","PC2"]*100, 1), "% of variation)")
  )+
  theme(
    # --- main panel: black frame ---
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    
    # --- side panels: small ---
    ggside.panel.scale = 0.15,          # smaller = thinner strips (try 0.10–0.20)
    
    # --- side panels: no grid ---
    ggside.panel.grid = element_blank(),
    
    # --- side panels: no frame/border of their own, hide their axes ---
    ggside.panel.border = element_blank(),
    ggside.panel.background = element_blank(),
    ggside.axis.text  = element_blank(),
    ggside.axis.ticks = element_blank(),
    
  legend.position = "bottom"
  )


# light_pca %>% 
#   pivot_longer(cols=-c("plot","time","subplot","target","reg_category","mean_reg")) %>% 
#   filter(is.finite(value)) %>%    
#   group_by(name) %>% 
#   mutate(value=scale(value)) %>% 
#   ungroup() %>% 
#   select(plot,subplot,reg_category,name,value) %>% 
#   left_join(var_meta,
#             by=c("name"="variable")) %>% 
#   ggplot(aes(label,value,fill=reg_category))+
#   geom_boxplot()+
#   scale_fill_brewer(name = "Regeneration cover", palette = "YlGn",na.value = "grey")

```

### comparative effects of plant material and light distribution on the structure of regeneration

```{r}
d <- metric3d_all_unsliced %>% 
  filter(subplot=="plot",time==12) %>% 
  select(plot,
         rb_mean,rb_cv) %>% 
  left_join(H_ext_all %>% 
              filter(target==0.9,subplot=="plot",time==12)) %>% 
  mutate(rb_mean=log(rb_mean)) %>% 
  left_join(regen_cat) %>% 
  select(any_of(c("plot","reg_category","rb_mean","z_cross_agg","ext_ref_agg","padTot_agg",
                  "padAbove_agg","zPAD50_agg"))) %>%
  pivot_longer(cols = -c("plot","reg_category")) %>%
  filter(!is.na(reg_category)) %>%
  mutate(name=factor(name,levels =c("rb_mean","ext_ref_agg","z_cross_agg","padTot_agg",
                  "padAbove_agg","zPAD50_agg") )) %>% 
  group_by(name) %>% mutate(value = as.numeric(scale(value))) %>%
  ungroup() %>%
  left_join(var_meta, by = c("name" = "variable")) %>% 
  arrange(name) %>%        # type's level order drives it
  mutate(label=case_when(name=="slope_cross_agg"~paste0(label," 0.9"),
                         name=="z_cross_agg"~paste0(label," 0.9"),
                         TRUE~label),
         label = fct_inorder(label))


d <- d %>% mutate(reg_category = factor(reg_category))

cld_one <- function(df) {
  lev <- levels(droplevels(df$reg_category))
  if (length(lev) < 2) return(tibble(reg_category = lev, letter = ""))
  
  kw <- kruskal.test(value ~ droplevels(reg_category), data = df)
  if (is.na(kw$p.value) || kw$p.value > 0.05)
    return(tibble(reg_category = lev, letter = "a"))   # no differences
  
  dt <- dunn_test(df, value ~ reg_category, p.adjust.method = "BH")
  pv <- setNames(dt$p.adj, paste(dt$group1, dt$group2, sep = "-"))
  l  <- multcompLetters(pv)$Letters
  tibble(reg_category = names(l), letter = unname(l))
}

cld_df <- d %>%
  group_by(type, label) %>%
  group_modify(~ cld_one(.x)) %>%
  ungroup() %>%
  mutate(reg_category = factor(reg_category, levels = levels(d$reg_category))) %>%
  left_join(
    d %>% group_by(type, label, reg_category) %>%
      summarise(y = max(value, na.rm = TRUE), .groups = "drop"),
    by = c("type", "label", "reg_category")
  )


ggplot(d, aes(label, value, fill = reg_category)) +
  geom_boxplot(position = position_dodge(width = 0.8, preserve = "single"),
               width=0.5,outlier.shape = NA) +
  geom_point(aes(color=reg_category),
             position = position_jitterdodge(jitter.width = 0.15,
                                             dodge.width = 0.8,
                                             seed = 1),
             size = 1, alpha = 0.8, shape = 16,
             show.legend = FALSE) +
  geom_text(data = cld_df,
            aes(x = label, y = y + 0.25, label = letter, group = reg_category),
            position = position_dodge(width = 0.8),
            inherit.aes = FALSE, size = 3, vjust = 0) +
  scale_fill_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  scale_color_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  facet_grid(~type, scales = "free_x") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.12)))+
  labs(y="Scaled values",x="")+
  theme_minimal(base_size = 9)+
  theme(legend.position = "none")->a


ggplot() +
  geom_point(
    data = light_pca_pred,
    aes(PC1, PC2, color = reg_category),
    size=2.5,show.legend = TRUE
  ) +
  scale_color_brewer(name = "Regeneration cover", palette = "YlGn",na.value = "grey") +
  
  new_scale_color() +
  
  geom_segment(
    data = var_coord,
    aes(x = 0, y = 0, xend = PC1*5, yend = PC2*5,  color = type), #alpha = contrib,
    arrow = arrow(length = unit(0.20, "cm")),
    lineend = "round"
  ) +
  geom_text_repel(
    data = var_coord,
    aes(x = PC1*5, y = PC2*5, label = label,  color = type), #alpha = contrib,
    size = 3, show.legend = FALSE
  ) +
  scale_color_brewer(name = "Variable type", palette = "Dark2") +
  guides(color="none")+
  geom_hline(yintercept = 0, colour = "grey85", linewidth = 0.3) +
  geom_vline(xintercept = 0, colour = "grey85", linewidth = 0.3) +  
  coord_cartesian() +
  theme_minimal(base_size = 9) +

  guides(
    alpha = "none",
    color = guide_legend(override.aes = list(alpha = 1))
  ) +
  labs(
    x = paste0("PCA axis 1 (", round(summary(pca_res)$importance["Proportion of Variance","PC1"]*100, 1), "% of variation)"),
    y = paste0("PCA axis 2 (", round(summary(pca_res)$importance["Proportion of Variance","PC2"]*100, 1), "% of variation)")
  )+
  theme(
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    legend.position = "left"
  )->b


cowplot::plot_grid(b+theme(legend.position="none"),a,ncol=2,align = "h", axis = "tb",
                   rel_widths = c(1, 1.5))


```



### effect size

```{r}
pca_data_plot %>% 
  left_join(metric3d_all_unsliced %>% 
              filter(subplot=="plot",time==12) %>% 
              # filter(slice_num==1) %>% 
              select(plot,
                     rb_mean,rb_cv)) %>% 
  left_join(H_ext_all %>% 
              filter(target==0.5,subplot=="plot",time==12)) %>% 
  mutate(rb_mean=log(rb_mean),
         rb_cv=log(rb_cv))->light_test


explained <- c("richness","richness_sapling","H","cwm_shade","abundance","pad_tot","Height_increment_mn","empty_gf","H_height","fhd","ri","pad_cv","abundance_cv")
explaining <- c("rb_mean","z_cross_agg","ext_ref_agg","slope_cross_agg","rb_cv","z_cross_cv","ext_ref_cv","slope_cross_cv")



std_coef <- function(d, y, x) { #, covar = "browsing"
  d <- d %>%
    select(all_of(c(y, x))) %>% #, covar
    drop_na() %>%
    mutate(across(where(is.numeric), ~ as.numeric(scale(.))))  # factors (e.g. browsing) left alone
  f <- reformulate(c( x), response = y) #covar,
  unname(coef(lm(f, data = d))[x])
}

# --- bootstrap distribution of that coef ---
boot_coef <- function(d, y, x, B = 1000) { #, covar = "browsing"
  d <- d %>% select(all_of(c(y, x))) %>% drop_na() %>% # , covar
      mutate(across(where(is.numeric), ~ as.numeric(scale(.)))) 
  n <- nrow(d)
  f <- reformulate(c( x), response = y) # covar,
  map_dbl(seq_len(B), function(b) {
    db <- d[sample(n, n, replace = TRUE), ]  # <- move this line out of the loop to scale once
    unname(coef(lm(f, data = db))[x])
  })
}

results <- expand_grid(explained = explained, explaining = explaining) %>%
  mutate(
    estimate = map2_dbl(explained, explaining, ~ std_coef(light_test, .x, .y)),
    draws    = map2(explained, explaining, ~ boot_coef(light_test, .x, .y, B = 1000))
  )


boot_long <- results %>%
  select(explained, explaining, draws) %>%
  unnest_longer(draws) %>%
  rename(coef = draws)

boot_long %>%
  group_by(explained, explaining) %>%
  summarise(
    lo  = unname(quantile(coef, 0.05)),
    med = unname(quantile(coef, 0.5)),
    hi  = unname(quantile(coef, 0.95)),
    .groups = "drop"
  ) %>%
  pivot_longer(cols=c("explained","explaining")) %>% 
  left_join(var_meta,by=c("value"="variable")) %>% 
  select(-value) %>% 
  pivot_wider(names_from="name",values_from = c("label","type")) %>% 
  mutate(
    type_explained = factor(type_explained, levels = c("Composition","Quantity","Complexity")),
    excl0 = lo > 0 | hi < 0                         # interval excludes zero
  ) %>% 
  ggplot( aes(x = med, y = label_explained)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = .3, colour = "grey70") +
  geom_linerange(aes(xmin = lo, xmax = hi, colour = excl0), linewidth = .9) +
  geom_point(aes(colour = excl0), size = 1.3) +
  scale_colour_manual(
    values = c(`TRUE` = "#1f5fa8", `FALSE` = "grey78")
  ) +
  guides(colour="none")+
  facet_grid(type_explained ~ label_explaining, scales = "free_y", space = "free_y") +
  labs(x = "Standardized coefficient", y = NULL) +
  theme_bw(base_size = 9) +
  theme(
    panel.spacing.x = unit(.6, "lines"),
    strip.text.y    = element_text(angle = 0),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )

# 
#  anova(lm(richness~rb_sd+rb_mean,data=light_test))
# anova(lm(richness~rb_cv+rb_mean,data=light_test))
# summary(lm(richness~rb_mean,data=light_test))
# summary(lm(richness~rb_cv,data=light_test))
```
## subplot

### between subplot

```{r}
metric3d_all_unsliced %>%
  filter(subplot!="plot",time==12) %>%
  # filter(slice_num==1) %>%
  select(plot,subplot,
         rb_mean,rb_cv) %>%
  left_join(H_ext_all %>%
              filter(extinction==0.88,subplot!="plot",time==12)) %>%
  mutate(rb_mean=log(rb_mean)) %>%
  rename("Radiative\nbudget mean"=rb_mean, "Radiative\nbudget var."=rb_cv,
      "Half-ext. \nheight"=mean, "Half-ext.\nheight (SD)"=sd,
      "Extinction\nspeed"=mean_slope, "Extinction\nspeed (SD)"=sd_slope) %>%
  left_join(regen_cat_subplot)->light_pca



# Run PCA (center and scale recommended)
pca_res <- prcomp(light_pca[,c("Radiative\nbudget mean", "Radiative\nbudget var.",
      "Half-ext. \nheight", "Half-ext.\nheight (SD)",
      "Extinction\nspeed", "Extinction\nspeed (SD)")], center = TRUE, scale. = TRUE)

# fviz_pca_biplot(
#   pca_res,
#   col.var = "contrib",            # color by contribution
#   gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
#   repel = TRUE
# ) 
# 
# GGally::ggpairs(light_pca)


light_pca_pred = cbind(light_pca,predict(pca_res,
                                         light_pca[,c("Radiative\nbudget mean", "Radiative\nbudget var.",
                                                      "Half-ext. \nheight", "Half-ext.\nheight (SD)",
                                                      "Extinction\nspeed", "Extinction\nspeed (SD)")])[,c(1,2)])


var <- get_pca_var(pca_res)

var_coord <- as.data.frame(var$coord[, 1:2])
colnames(var_coord) <- c("PC1", "PC2")

var_coord$variable <- rownames(var_coord)
var_coord$contrib <- var$contrib[, 1] + var$contrib[, 2]
var_coord$cos2 <- var$cos2[, 1] + var$cos2[, 2]


var_coord=var_coord %>% 
  mutate(type=case_when(variable%in%c("Radiative\nbudget mean", "Half-ext. \nheight", "Extinction\nspeed")~"Mean",
                        variable%in%c("Radiative\nbudget var.","Half-ext.\nheight (SD)","Extinction\nspeed (SD)")~"Variability"))


ggplot() +
  geom_point(
    data = light_pca_pred,
    aes(PC1, PC2, color = reg_category),
    size=2.5
  ) +
  scale_color_brewer(name = "Regeneration cover", palette = "YlGn",na.value = "grey") +

  new_scale_color() +

  geom_segment(
    data = var_coord,
    aes(x = 0, y = 0, xend = PC1*5, yend = PC2*5,  color = type), #alpha = contrib,
    arrow = arrow(length = unit(0.20, "cm")),
    lineend = "round"
  ) +
  geom_text_repel(
    data = var_coord,
    aes(x = PC1*5, y = PC2*5, label = variable,  color = type), #alpha = contrib,
    size = 3, show.legend = FALSE
  ) +
  scale_color_brewer(name = "Variable type", palette = "Dark2") +

  coord_equal() +
  theme_minimal() +
  guides(
    alpha = "none",
    color = guide_legend(override.aes = list(alpha = 1))
  ) +
  labs(
    x = paste0("PCA axis 1 (", round(summary(pca_res)$importance["Proportion of Variance","PC1"]*100, 1), "% of variation)"),
    y = paste0("PCA axis 2 (", round(summary(pca_res)$importance["Proportion of Variance","PC2"]*100, 1), "% of variation)")
  )



metric3d_all_unsliced %>% 
  filter(subplot!="plot",time==12,plot!="GER18") %>% 
  select(plot,
         rb_mean,rb_cv) %>% 
  left_join(H_ext_all %>% 
              filter(extinction==0.88,subplot!="plot",time==12)) %>% 
  rename("Radiative\nbudget mean"=rb_mean, "Radiative\nbudget var."=rb_cv,
      "Half-ext. \nheight"=mean, "Half-ext.\nheight (SD)"=sd,
      "Extinction\nspeed"=mean_slope, "Extinction\nspeed (SD)"=sd_slope) %>% 
  pivot_longer(cols=-c("plot","time","subplot","extinction")) %>% 
  filter(is.finite(value)) %>% 
  group_by(name) %>% 
  mutate(value=scale(value)) %>% 
  ungroup() %>% 
  select(plot,subplot,name,value) %>% 
  left_join(regen_cat_subplot) %>%
  filter(!is.na(reg_category)) %>% 
  ggplot(aes(name,value,fill=reg_category))+
  geom_boxplot()+
  scale_fill_brewer(name = "Regeneration cover", palette = "YlGn",na.value = "grey")

```


### effect size

```{r}
pca_data_subplot %>% 
  left_join(metric3d_all %>% 
              filter(subplot!="plot",time==12) %>% 
              filter(slice_num==1) %>% 
              select(plot,subplot,
                     rb_mean,rb_cv)) %>% 
  left_join(H_ext_all %>% 
              filter(extinction==0.5,subplot!="plot",time==12))%>% 
  mutate(rb_mean=log(rb_mean))->light_test_sub


explained <- c("richness","richness_sapling","H","cwm_shade","abundance","pad_tot","Height_increment_mn","empty_gf","H_height","fhd","ri","pad_cv")
explaining <- c("rb_mean","rb_cv","mean","sd","mean_slope","sd_slope")

std_coef <- function(d, y, x) {#, covar = "browsing"
  d <- d %>%
    select(all_of(c(y, x, "plot"))) %>%#, covar
    drop_na() %>%
    mutate(plot = factor(plot)) %>%                              # grouping factor, kept out of scaling
    mutate(across(where(is.numeric), ~ as.numeric(scale(.))))    # factors (browsing, plot) left alone
  f <- as.formula(paste0(y, " ~ ",  " + ", x, " + (1 | plot)"))#covar,
  m <- lmer(f, data = d, control = lmerControl(calc.derivs = FALSE))
  unname(fixef(m)[x])
}

# --- bootstrap distribution of that coef ---
boot_coef <- function(d, y, x, B = 1000) {# covar = "browsing",
  d <- d %>% select(all_of(c(y, x, "plot"))) %>% drop_na() %>%#covar, 
      mutate(plot = factor(plot)) %>%
      mutate(across(where(is.numeric), ~ as.numeric(scale(.))))
  n <- nrow(d)
  f <- as.formula(paste0(y, " ~ ", " + ", x, " + (1 | plot)"))# covar,
  map_dbl(seq_len(B), function(b) {
    db <- d[sample(n, n, replace = TRUE), ]
    m  <- suppressMessages(suppressWarnings(
            lmer(f, data = db, control = lmerControl(calc.derivs = FALSE))))
    unname(fixef(m)[x])
  })
}


results <- expand_grid(explained = explained, explaining = explaining) %>%
  mutate(
    estimate = map2_dbl(explained, explaining, ~ std_coef(light_test_sub, .x, .y)),
    draws    = map2(explained, explaining, ~ boot_coef(light_test_sub, .x, .y, B = 1000))
  )


boot_long <- results %>%
  select(explained, explaining, draws) %>%
  unnest_longer(draws) %>%
  rename(coef = draws)

boot_long %>%
  group_by(explained, explaining) %>%
  summarise(
    lo  = unname(quantile(coef, 0.05)),
    med = unname(quantile(coef, 0.5)),
    hi  = unname(quantile(coef, 0.95)),
    .groups = "drop"
  ) %>%
  mutate(
    type = factor(case_when(
      explained %in% c("richness","richness_sapling","H","cwm_shade") ~ "Composition",
      explained %in% c("abundance","pad_tot","h_mean","Height_increment_mn","empty_gf") ~ "Quantity",
      explained %in% c("H_height","h_cv","pad_cv","ri","fhd") ~ "Complexity"
    ), levels = c("Composition","Quantity","Complexity")),
    excl0 = lo > 0 | hi < 0,                         # interval excludes zero
    explained = recode(explained,
      richness="Richness\ntotal", richness_sapling="Richness\nsaplings",
      H="Shannon\nindex", cwm_shade="Shade tolerance",
      abundance="Abundance", pad_tot="PAD\ntotal",
      h_mean="Mean\nheight", Height_increment_mn="Height\nincrement",
      empty_gf="Gap fraction", H_height="Vertical\ndistribution",
      h_cv="Height\nvariability", pad_cv="PAD\nvariability",
      ri="Rumple index",fhd="Foliage height \ndiversity"),
    # explained = factor(explained, levels = rev(explained_order)),
    explaining = recode(explaining,
      rb_mean="Radiative\nbudget mean", rb_cv="Radiative\nbudget var.",
      mean="Half-ext. \nheight", sd="Half-ext.\nheight (SD)",
      mean_slope="Extinction\nspeed", sd_slope="Extinction\nspeed (SD)")
  ) %>% 
  ggplot( aes(x = med, y = explained)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = .3, colour = "grey70") +
  geom_linerange(aes(xmin = lo, xmax = hi, colour = excl0), linewidth = .9) +
  geom_point(aes(colour = excl0), size = 1.3) +
  scale_colour_manual(
    values = c(`TRUE` = "#1f5fa8", `FALSE` = "grey78")
  ) +
  guides(colour="none")+
  facet_grid(type ~ explaining, scales = "free_y", space = "free_y") +
  labs(x = "Standardized coefficient", y = NULL) +
  theme_bw(base_size = 9) +
  theme(
    panel.spacing.x = unit(.6, "lines"),
    strip.text.y    = element_text(angle = 0),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )

```

# effect of adults on light distribution and regeneration structure


```{r}
sem = plot_df %>% 
  select(plot,n_trees_alive,species_richness_alive,mean_dbh_alive_cm,
         mean_height_alive_m,mortality_rate) %>% 
  left_join(light_test)



explained <- c("richness","richness_sapling","H","cwm_shade","abundance","pad_tot","Height_increment_mn","empty_gf","H_height","fhd","ri","pad_cv")
explaining <- c("n_trees_alive","species_richness_alive","mean_dbh_alive_cm","mean_height_alive_m","mortality_rate","browsing")



std_coef <- function(d, y, x) {
  d <- d %>%
    select(all_of(c(y, x))) %>%
    drop_na() %>%
    mutate(across(where(is.numeric), ~ as.numeric(scale(.))))  # factors (e.g. browsing) left alone
  f <- reformulate(c( x), response = y)
  unname(coef(lm(f, data = d))[x])
}

# --- bootstrap distribution of that coef ---
boot_coef <- function(d, y, x,  B = 1000) {
  d <- d %>% select(all_of(c(y, x))) %>% drop_na() %>%
      mutate(across(where(is.numeric), ~ as.numeric(scale(.)))) 
  n <- nrow(d)
  f <- reformulate(c( x), response = y)
  map_dbl(seq_len(B), function(b) {
    db <- d[sample(n, n, replace = TRUE), ]  # <- move this line out of the loop to scale once
    unname(coef(lm(f, data = db))[x])
  })
}

results <- expand_grid(explained = explained, explaining = explaining) %>%
  mutate(
    estimate = map2_dbl(explained, explaining, ~ std_coef(sem, .x, .y)),
    draws    = map2(explained, explaining, ~ boot_coef(sem, .x, .y, B = 1000))
  )


boot_long <- results %>%
  select(explained, explaining, draws) %>%
  unnest_longer(draws) %>%
  rename(coef = draws)

boot_long %>%
  group_by(explained, explaining) %>%
  summarise(
    lo  = unname(quantile(coef, 0.1)),
    med = unname(quantile(coef, 0.5)),
    hi  = unname(quantile(coef, 0.9)),
    .groups = "drop"
  ) %>%
  mutate(
    type = factor(case_when(
      explained %in% c("richness","richness_sapling","H","cwm_shade") ~ "Composition",
      explained %in% c("abundance","pad_tot","h_mean","Height_increment_mn","empty_gf") ~ "Quantity",
      explained %in% c("H_height","h_cv","pad_cv","ri","fhd") ~ "Complexity"
    ), levels = c("Composition","Quantity","Complexity")),
    excl0 = lo > 0 | hi < 0,                         # interval excludes zero
    explained = recode(explained,
      richness="Richness\ntotal", richness_sapling="Richness\nsaplings",
      H="Shannon\nindex", cwm_shade="Shade tolerance",
      abundance="Abundance", pad_tot="PAD\ntotal",
      h_mean="Mean\nheight", Height_increment_mn="Height\nincrement",
      empty_gf="Gap fraction", H_height="Vertical\ndistribution",
      h_cv="Height\nvariability", pad_cv="PAD\nvariability",
      ri="Rumple index",fhd="Foliage height \ndiversity"),
    # explained = factor(explained, levels = rev(explained_order)),
    explaining = recode(explaining,
      n_trees_alive="Number of\n living trees", species_richness_alive="Adult richness",
      mean_dbh_alive_cm="Mean diameter", mean_height_alive_m="Mean height",
      mortality_rate="Mortality"),
    
  ) %>% 
  ggplot( aes(x = med, y = explained)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = .3, colour = "grey70") +
  geom_linerange(aes(xmin = lo, xmax = hi, colour = excl0), linewidth = .9) +
  geom_point(aes(colour = excl0), size = 1.3) +
  scale_colour_manual(
    values = c(`TRUE` = "#1f5fa8", `FALSE` = "grey78")
  ) +
  guides(colour="none")+
  facet_grid(type ~ explaining, scales = "free_y", space = "free_y") +
  labs(x = "Standardized coefficient", y = NULL) +
  theme_bw(base_size = 9) +
  theme(
    panel.spacing.x = unit(.6, "lines"),
    strip.text.y    = element_text(angle = 0),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )

```

# Disentangling indirect effect of adults

```{r}
sem = plot_df %>% 
  select(plot,n_trees_alive,species_richness_alive,mean_dbh_alive_cm,
         mean_height_alive_m,mortality_rate) %>% 
  left_join(light_test) %>% 
  mutate(rb_mean=log(rb_mean)) %>% 
  mutate(across(where(is.numeric), ~ as.numeric(scale(.)))) #%>% 
  # rename("Number of\n living trees"=n_trees_alive, "Adult richness"=species_richness_alive,
  #     "Mean diameter"=z, "Mean height"=mean_height_alive_m,
  #     mortality_rate="Mortality")



library(piecewiseSEM)
library(semPlot)
library(lavaan)

lavaan_model <- '
  richness ~ mean_slope + rb_cv + n_trees_alive+ species_richness_alive+ mean_dbh_alive_cm+
         mean_height_alive_m
  mean_slope ~  n_trees_alive+ species_richness_alive+ mean_dbh_alive_cm+
         mean_height_alive_m
  rb_cv    ~  n_trees_alive+ species_richness_alive+ mean_dbh_alive_cm+
         mean_height_alive_m
  
'


lavaan_model <- '
  richness ~ mean + rb_cv + n_trees_alive+ species_richness_alive+ mean_dbh_alive_cm+
         mean_height_alive_m
  h_mean ~   mean + rb_cv + n_trees_alive+ species_richness_alive+ mean_dbh_alive_cm+
         mean_height_alive_m
  empty_gf~  mean + rb_mean + sd + n_trees_alive+ species_richness_alive+ mean_dbh_alive_cm+
         mean_height_alive_m
  abundance ~ mean_slope + mean + rb_mean + sd + n_trees_alive+ species_richness_alive+ mean_dbh_alive_cm+ mean_height_alive_m
  pad_cv ~ mean + rb_mean + n_trees_alive+ species_richness_alive+ mean_dbh_alive_cm+ mean_height_alive_m
  mean_slope ~  n_trees_alive+ species_richness_alive+ mean_dbh_alive_cm+
         mean_height_alive_m
  sd ~ n_trees_alive+ species_richness_alive+ mean_dbh_alive_cm+
         mean_height_alive_m
  rb_cv    ~  n_trees_alive+ species_richness_alive+ mean_dbh_alive_cm+
         mean_height_alive_m
  rb_mean  ~ n_trees_alive+ species_richness_alive+ mean_dbh_alive_cm+
         mean_height_alive_m
  
'

fit_lav <- sem(lavaan_model, data = sem)
summary(fit_lav)
# Plot
semPaths(
  object         = fit_lav,
  what           = "std",
  whatLabels     = "std",
  layout         = "tree2",
  edge.label.cex = 1.2,
  node.label.cex = 1.2,
  residuals      = FALSE,
  intercepts     = FALSE
)




# 1. Extract coefficients from piecewiseSEM
coef_table <- coefs(fit_lav, standardize = "scale")

# 2. Add significance stars
coef_table$label <- with(coef_table, ifelse(P.Value < 0.001, paste0(round(Std.Estimate, 2), "***"),
                                    ifelse(P.Value < 0.01,  paste0(round(Std.Estimate, 2), "**"),
                                    ifelse(P.Value < 0.05,  paste0(round(Std.Estimate, 2), "*"),
                                    ifelse(P.Value < 0.1,   paste0(round(Std.Estimate, 2), "."),
                                                             round(Std.Estimate, 2))))))

# 3. Plot with custom labels
semPaths(
  object         = fit_lav,
  what           = "std",
  whatLabels     = "std",
  layout         = "tree2",
  edge.label.cex = 1.2,
  node.label.cex = 1.2,
  residuals      = FALSE,
  intercepts     = FALSE,
  edgeLabels     = coef_table$label  # inject piecewiseSEM labels with stars
)



mod_sem <- list(
  rb_mean     = lm(rb_mean     ~ species_richness_alive,        data = sem),
  richness_sapling = lm(richness_sapling ~ rb_mean + species_richness_alive, data = sem)
)
boot_sem <- semEff::bootEff(mod_sem, R = 500, seed = 13, parallel = "no")



sem_eff <- semEff(boot_sem)
summary(sem_eff, response = "richness_sapling")

S <- sem_eff$Summary$richness


eff_tab <- getEffTable(sem_eff)          # clean numeric table, all responses

plot_df <- subset(
  eff_tab,
  response == "richness.sapling" & effect_type %in% c("direct", "indirect", "total")
)
plot_df$effect_type <- factor(plot_df$effect_type,
                              levels = c("direct", "indirect", "total"),
                              labels = c("Direct", "Indirect", "Total"))
plot_df$sig <- with(plot_df, lower_ci > 0 | upper_ci < 0)

ggplot(plot_df, aes(effect, predictor, colour = effect_type)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey70") +
  geom_pointrange(
    aes(xmin = lower_ci, xmax = upper_ci, alpha = sig),
    position  = position_dodge(width = 0.6),
    linewidth = 0.7, fatten = 3
  ) +
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.35), guide = "none") +
  scale_colour_manual(values = c(Direct = "#2c7fb8",
                                 Indirect = "#d95f0e",
                                 Total = "#31a354")) +
  labs(x = "Standardised effect on richness", y = NULL, colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank(),
        legend.position = "top")
```


```{r}
library(semEff)
library(DiagrammeR)

et <- getEffTable(sem_eff)   # the tidy table from before

# one direct path: its coefficient + whether the CI excludes zero
path <- function(resp, pred) {
  x <- et[et$effect_type == "direct" & et$response == resp & et$predictor == pred, ]
  list(coef = x$effect, sig = x$lower_ci > 0 | x$upper_ci < 0)
}

sra_mean  <- path("rb.mean",     "species.richness.alive")  # indirect path, step 1
mean_rich <- path("richness.sapling", "rb.mean")                    # indirect path, step 2
sra_rich  <- path("richness.sapling", "species.richness.alive")  # direct path

# indirect effect (sra -> mean -> richness), straight from the table
ind <- et[et$effect_type == "indirect" &
          et$response == "richness.sapling" &
          et$predictor == "species.richness.alive", "effect"]

# style each arrow: colour = sign, solid = CI excludes 0, thickness = |coef|
edge <- function(from, to, p) {
  col <- if (p$coef >= 0) "#2166ac" else "#b2182b"
  sprintf('%s -> %s [label = " %.2f", color = "%s", fontcolor = "%s", style = "%s", penwidth = %.2f]',
          from, to, p$coef, col, col,
          if (isTRUE(p$sig)) "solid" else "dashed",
          1 + 4 * abs(p$coef))
}

dot <- sprintf('
digraph sem {
  graph [rankdir = LR, ranksep = 1.2, nodesep = 0.9,
         label = "Indirect effect (species richness -> mean -> richness): %.2f",
         labelloc = "b", fontname = "Helvetica", fontsize = 10, fontcolor = "#777777"]
  node  [shape = box, style = "rounded,filled", fillcolor = "#f7f7f7",
         color = "#666666", fontname = "Helvetica", fontsize = 12, margin = "0.22,0.13"]
  edge  [fontname = "Helvetica", fontsize = 11, arrowsize = 0.8]

  sra  [label = "species richness\\n(alive)"]
  mean [label = "mean"]
  rich [label = "richness"]

  %s
  %s
  %s
}',
  ind,
  edge("sra",  "mean", sra_mean),
  edge("mean", "rich", mean_rich),
  edge("sra",  "rich", sra_rich))

grViz(dot)
```

```{r}
library(piecewiseSEM)
psem_mod <- psem(
  lm(richness ~ mean + species_richness_alive, data = sem),
  lm(mean ~ species_richness_alive,            data = sem),
  data = sem
)

plot(
  psem_mod,
  show       = "std",     # standardized coefficients on the arrows
  digits     = 2,
  ns_dashed  = TRUE,      # non-significant paths dashed
  node_attrs = data.frame(shape = "rectangle", fillcolor = "#f7f7f7",
                          color = "#666666", fontname = "Helvetica"),
  edge_attrs = data.frame(fontname = "Helvetica")
)
```

# species analysis

```{r}
data.frame(plot=unique(regen_metrics_subplot$species$plot)) %>% 
  expand_grid(subplot=unique(regen_metrics_subplot$species$subplot)) %>% 
  expand_grid(Species=unique(regen_metrics_subplot$species$Species)) %>% 
  filter(Species!="NA") %>% 
  left_join(regen_metrics_subplot$species) %>% 
  select(plot,subplot,species=Species,abundance) %>% 
  mutate(abundance=if_else(is.na(abundance),0,abundance)) %>% 
  left_join(metric3d_all %>% 
              filter(time==12,subplot!="plot") %>% 
              select(plot,subplot,rb_mean,rb_cv,occupied_gf)) %>% 
  left_join(H_ext_all %>% 
              filter(time==12,extinction==0.5) %>% 
              select(plot,subplot,mean,mean_slope)) %>% 
  group_by(species) %>% 
  mutate(n=n()) %>% 
  filter(n>30) %>% 
  select(-n) %>% 
  pivot_longer(cols=-c("plot","subplot","species","abundance")) %>% 
  ggplot(aes(abundance,value))+
  geom_point()+
  facet_wrap(species~name,scales="free", ncol=5)



regen_metrics_plot$species %>% 
  select(plot,species=Species,abundance) %>% 
  left_join(metric3d_all %>% 
              filter(time==12,subplot=="plot") %>% 
              select(plot,subplot,slice_num,rb_mean,rb_cv,occupied_gf)) %>% 
  left_join(H_ext_all %>% 
              filter(time==12,extinction==0.5) %>% 
              select(plot,subplot,mean,mean_slope)) %>% View()
  filter(plot!="GER10") %>% 
  group_by(species) %>% 
  mutate(n=n()) %>% 
  filter(n>30) %>% 
  select(-n) %>% 
  ggplot(aes(log(abundance),log(rb_mean),color=species))+
  geom_point()+
  geom_smooth(method="gam",se = FALSE)+
  facet_grid(species~slice_num)
  
  
data.frame(plot=unique(regen_metrics_subplot$species$plot)) %>% 
  # expand_grid(subplot=unique(regen_metrics_subplot$species$subplot)) %>% 
  expand_grid(Species=unique(regen_metrics_subplot$species$Species)) %>% 
  filter(Species!="NA") %>% 
  left_join(regen_metrics_subplot$species) %>% 
  select(plot,species=Species,abundance) %>% 
  mutate(abundance=if_else(is.na(abundance),0,abundance)) %>% 
  left_join(metric3d_all %>% 
              filter(time==12,subplot=="plot") %>% 
              select(plot,subplot,slice_num,rb_mean,rb_cv,occupied_gf)) %>% 
  left_join(H_ext_all %>% 
              filter(time==12,extinction==0.5) %>% 
              select(plot,subplot,mean,mean_slope)) %>%
  filter(plot!="GER10") %>% 
  group_by(species) %>% 
  mutate(n=n()) %>% 
  filter(n>30) %>% 
  select(-n) %>% 
  ggplot(aes(abundance,log(rb_mean),color=species))+
  geom_point()+
  geom_smooth(method="gam",se = FALSE)+
  facet_grid(species~slice_num)
  



df <- regen_metrics_subplot$all %>% select(plot,subplot) %>% unique() %>% 
  expand_grid(Species=unique(regen_metrics_subplot$species$Species))%>%
  filter(Species != "NA") %>%
  left_join(regen_metrics_subplot$species) %>%
  select(plot, subplot, species = Species, abundance) %>%
  mutate(abundance=if_else(is.na(abundance),0,abundance),
         presence = if_else(is.na(abundance), "absent", "present")) %>%
  left_join(metric3d_all %>%
              filter(time == 12, subplot != "plot", slice_num == 1) %>%
              select(plot, subplot, rb_mean, rb_cv, occupied_gf)) %>%
  mutate(log_rb_mean = log(rb_mean)) %>%
  filter(is.finite(log_rb_mean))



df %>%
  group_by(species) %>%
  filter(n_distinct(presence) == 2, all(table(presence) >= 1)) %>%  # need both groups, a few obs each
  wilcox_test(log_rb_mean ~ presence) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

regen_metrics_subplot$all %>% select(plot,subplot) %>% unique() %>% 
  expand_grid(Species=unique(regen_metrics_subplot$species$Species)) %>% 
  filter(Species!="NA") %>% 
  left_join(regen_metrics_subplot$species) %>% 
  select(plot,subplot,species=Species,abundance) %>% 
  mutate(presence=if_else(is.na(abundance),"absent","present")) %>% 
  left_join(metric3d_all %>% 
              filter(time==12,subplot!="plot",slice_num==1) %>% 
              select(plot,subplot,rb_mean,rb_cv,occupied_gf)) %>% 
  ggplot(aes(species,log(rb_mean),color=presence))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter()+
  ylim(c(5.5,9))
  



df <- regen_metrics_plot$all %>% select(plot) %>% unique() %>% 
  expand_grid(Species=unique(regen_metrics_plot$species$Species)) %>% 
  filter(Species!="NA") %>% 
  left_join(regen_metrics_plot$species) %>% 
  select(plot,species=Species,abundance) %>% 
  mutate(presence=if_else(is.na(abundance),"absent","present")) %>% 
  left_join(metric3d_all %>% 
              filter(time==12,subplot=="plot",slice_num==1) %>% 
              select(plot,rb_mean,rb_cv,occupied_gf)) %>% 
  mutate(log_rb_mean = log(rb_mean)) %>%
  filter(is.finite(log_rb_mean))

regen_metrics_plot$all %>% select(plot) %>% unique() %>% 
  expand_grid(Species=unique(regen_metrics_plot$species$Species)) %>% 
  filter(Species!="NA") %>% 
  left_join(regen_metrics_plot$species) %>% 
  select(plot,species=Species,abundance) %>% 
  mutate(presence=if_else(is.na(abundance),"absent","present")) %>% 
  left_join(metric3d_all %>% 
              filter(time==12,subplot=="plot",slice_num==1) %>% 
              select(plot,rb_mean,rb_cv,occupied_gf)) %>% 
  ggplot(aes(species,log(rb_mean),color=presence))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter()+
  ylim(c(5.5,9))
  
  
library(lme4); library(lmerTest)

species_list <- df %>%
  group_by(species) %>%
  filter(n_distinct(presence) == 2, n_distinct(plot) >= 3) %>%
  pull(species) %>% unique()

results <- lapply(species_list, function(sp) {
  d <- df[df$species == sp, ]
  tryCatch({
    m  <- lm(log_rb_mean ~ presence , data = d)
    co <- summary(m)$coefficients["presencepresent", ]
    data.frame(species = sp,
               estimate  = co["Estimate"],
               std.error = co["Std. Error"],
               statistic = co["t value"],
               p.value   = co["Pr(>|t|)"])
  }, error = function(e)
    data.frame(species = sp, estimate = NA, std.error = NA,
               statistic = NA, p.value = NA))
})

res <- do.call(rbind, results)
res$p.adj <- p.adjust(res$p.value, method = "BH")
res




df$rb_c <- df$log_rb_mean - mean(df$log_rb_mean, na.rm = TRUE)   # centred predictor
df$species <- factor(df$species)

library(glmmTMB)

mnb <- glmmTMB(abundance ~ species * rb_c + (1 | plot),
               data = df, family = nbinom2)

# if there are more zeros than even NB expects (common), add a zero-inflation part:
mzi <- glmmTMB(abundance ~ species * rb_c + (1 | plot),
               ziformula = ~1, data = df, family = nbinom2)

AIC(mnb, mzi)  


library(emmeans)

et <- emtrends(mnb, ~ species, var = "rb_c")   # or mp / mzi
et                                             # per-species slope on log scale

# as rate ratios (abundance multiplies by this per 1-unit rb_mean):
summary(et, infer = TRUE) |>
  transform(RR = exp(rb_c.trend))
```

