---
title: "Figures exploration"
author: "Anne Baranger"
date: "2026-05-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE,message=FALSE)
library(targets)
lapply(c("stringr","tidyr","dplyr","tibble", # data analysis
         "ggplot2","viridis", "ggridges","wesanderson",# plot
         "factoextra", "nlme","modi","rstan","shinystan",# stat & models
         "rworldmap" # geographic data #terre, sf
         ),require,character.only=TRUE)
theme_set(theme_bw()) # ggplot theme
common_theme <- theme(
  # legend.position = "none",
  strip.background = element_rect(fill = "white",
                                        colour = "gray12"),
  panel.background = element_rect(fill = "white",
                                  colour = "gray12",
                                  size = 0.5, linetype = "solid"),
  plot.title = element_text(size = 12),
  axis.title = element_text(size = 10),
  axis.text = element_text(size = 10),
  strip.text = element_text(size =10),
  plot.margin = unit(c(0, 0.2, 0, 0.2), "cm")  # Example margin; adjust as needed
)

loadRData <- function(fileName){
  #loads an RData file, and returns it
  load(fileName)
  get(ls()[ls() != "fileName"])
}

```

# Data preparation 

## Loading classic data. 

```{r}
tar_load(metric3d_all)
tar_load(regen_metrics_plot)
tar_load(regen_metrics_subplot)
tar_load(plot_df)


metric3d_all %>% 
  select(plot,time) %>% 
  unique() %>% group_by(plot) %>% 
  summarise(`Number of simulated time slots`=n())

metric3d_all %>% 
  select(plot,subplot) %>% 
  unique() %>% group_by(plot) %>% 
  summarise(`Number of subplots`=n()-1)

```


# Exploration of complexity measures 

```{r,fig.height=10,fig.width=10}

metric3d_all_plot=metric3d_all %>% 
  filter(subplot=="plot") %>% 
  pivot_longer(cols =  c(matches("h_"),"occupied_gf",matches("pad")))

metric3d_all %>% 
  filter(subplot!="plot") %>% 
  pivot_longer(cols = c(matches("h_"),"occupied_gf",matches("pad"))) %>% 
  filter(!grepl("_na",name)) %>% 
  ggplot(aes(slice_num,value,color=plot))+
  geom_line(aes(group=interaction(plot,subplot)),alpha = 0.4)+
  geom_point(data=metric3d_all_plot,aes(slice_num,value,color=plot))+
  geom_line(data=metric3d_all_plot,aes(slice_num,value,color=plot,group=interaction(plot,subplot)))+
  facet_grid(name~plot,scale="free_y")+
  theme_bw()


# metric3d_all_plot
metric3d_all %>% 
  filter(subplot!="plot") %>% 
  filter(time==12) %>% 
  ggplot(aes(h_mean,h_sd,color=time))+
  geom_point(aes(group=interaction(time,subplot)),alpha = 0.4)+
  geom_line(aes(group=interaction(time,subplot)),alpha = 0.4)+
  geom_point(data=metric3d_all %>% filter(subplot=="plot") %>% filter(time==12),
             aes(h_mean,h_sd,color=time))+
  geom_line(data=metric3d_all %>% filter(subplot=="plot") %>% filter(time==12),
            aes(h_mean,h_sd,color=time,group=interaction(time,subplot)))+
  facet_wrap(~plot,scale="free_y")+
  theme_bw()
```

Main outcomes :
* measure height without accounting for empty areas doesn't make sense. Drop variables "_na"
* median doesn't make sense when accounting for empty areas, drop "h_med"
* "coef_lin" is not informative
* The shape of h_mean=f(slice_num) seems to provide insight on the understory structure. Linear form are more common for plots with less regeneration, while saturating form could be more common for dense understory

## Rank according to abondance

```{r}
metric3d_all_plot=metric3d_all %>% 
  filter(subplot=="plot")

regen_metrics_plot$all %>% 
  select(plot,n_sapling) %>% 
  mutate(n_sapling=replace_values(n_sapling, NA ~ 0)) %>% 
  unique() %>% 
  arrange(n_sapling) %>% 
  pull(plot)->plot_arrange
  
metric3d_all %>% 
  filter(subplot != "plot") %>% 
  mutate(plot = factor(plot, levels = plot_arrange)) %>% 
  ggplot(aes(slice_num, occupied_gf, color = plot)) +
  geom_line(aes(group = interaction(plot, subplot)), alpha = 0.4) +
  geom_point(
    data = metric3d_all_plot %>% 
      mutate(plot = factor(plot, levels = plot_arrange)),
    aes(slice_num, occupied_gf, color = plot)
  ) +
  geom_line(
    data = metric3d_all_plot %>% 
      mutate(plot = factor(plot, levels = plot_arrange)),
    aes(slice_num, occupied_gf, color = plot, group = interaction(plot, subplot))
  ) +
  facet_grid( ~ plot, scales = "free_y") +
  theme_bw()





regen_metrics_subplot$all %>% 
  mutate(plot_sub=paste0(plot,"_",substr(subplot,9,9))) %>% 
  select(plot_sub,n_sapling) %>%
  mutate(n_sapling=replace_values(n_sapling, NA ~ 0)) %>% 
  unique() %>% 
  arrange(n_sapling) ->subplot_arrange


metric3d_all %>% 
  filter(subplot != "plot") %>% 
  mutate(plot_sub = paste0(plot, "_", substr(subplot, 9, 9))) %>% 
  mutate(plot_sub = factor(plot_sub, levels = subplot_arrange$plot_sub)) %>% 
  left_join(subplot_arrange) %>% 
  ggplot(aes(slice_num, occupied_gf,
             group = plot_sub,
             color = n_sapling)) +
  geom_line(size = 0.9, alpha = 0.9) +
  scale_color_viridis_c(
    option = "plasma",
    trans = "log10"
  ) +
  theme_bw()


df_plot <- metric3d_all %>% 
  filter(subplot != "plot") %>% 
  mutate(
    plot_sub = paste0(plot, "_", substr(subplot, 9, 9)),
    plot_sub = factor(plot_sub, levels = subplot_arrange$plot_sub)
  ) %>% 
  left_join(subplot_arrange) %>% 
  mutate(
    abundance_cat = ntile(n_sapling, 5),
    abundance_cat = factor(
      abundance_cat,
      levels = 1:5
    )
  )

ggplot(df_plot, aes(slice_num, occupied_gf, group = plot_sub, color = n_sapling)) +
  geom_line(size = 1.1, alpha = 0.9) +
  scale_color_viridis_c(
    option = "plasma",
    trans = "log10"
  ) +
  facet_wrap(~ abundance_cat,ncol=5) +
  theme_bw()
```

# Exploration of field metrics

## relevance of seedling number
```{r}
regen_metrics_subplot$all %>% 
  ggplot(aes(abundance,n_seedling))+
  geom_point()+
  geom_abline(slope=1)+
  labs(x="Total abundance",
       y="Seedlings number (class 1 and 2)",
       title="Subplot level")

regen_metrics_plot$all %>% 
  ggplot(aes(abundance,n_seedling))+
  geom_point()+
  geom_abline(slope=1)+
  labs(x="Total abundance",
       y="Seedlings number (class 1 and 2)",
       title="Plot level")
```
Seedling number is highly correlated to total abundance at both subplot and plot level.

## relevance of sapling number
```{r}
regen_metrics_subplot$all %>% 
  ggplot(aes(abundance,n_sapling))+
  geom_point()+
  geom_smooth(method="lm")+
  labs(x="Total abundance",
       y="Saplings number (class 1 and 2)",
       title="Subplot level")

regen_metrics_plot$all %>% 
  ggplot(aes(abundance,n_sapling))+
  geom_point()+
  geom_smooth(method="lm")+
  labs(x="Total abundance",
       y="Saplings number (class 1 and 2)",
       title="Plot level")
```


## Browsing metrics
```{r}
regen_metrics_plot$all %>% 
  pivot_longer(cols=-c("plot","browsing")) %>% 
  mutate(name=factor(name,levels = c("abundance","abundance_cv","n_seedling","richness","richness_sd",
                                     "Height_increment_mn_hclass2","Height_increment_mn_hclass3",
                                     "Height_increment_cv_hclass2","Height_increment_cv_hclass3",
                                     "H",
                                     "H_hclass1","H_hclass2","H_hclass3","H_hclass4","H_hclass5"))) %>% 
  ggplot(aes(browsing,value))+
  geom_point()+
  geom_smooth(method="lm")+
  facet_wrap(~name,scale="free_y",ncol=5)+
  theme_bw()+
  labs(title = "At the plot scale")



regen_metrics_subplot$all %>% 
  pivot_longer(cols=-c("plot","subplot","browsing")) %>% 
  mutate(name=factor(name,levels = c("abundance","abundance_cv","n_seedling","richness","richness_sd",
                                     "Height_increment_mn_hclass2","Height_increment_mn_hclass3",
                                     "Height_increment_cv_hclass2","Height_increment_cv_hclass3",
                                     "H",
                                     "H_hclass1","H_hclass2","H_hclass3","H_hclass4","H_hclass5"))) %>% 
  ggplot(aes(browsing,value))+
  geom_point()+
  geom_smooth(method="lm")+
  facet_wrap(~name,scale="free_y",ncol=5)+
  theme_bw()+
  labs(title = "At the subplot scale")
```
Browsing reduces height increment, probably differs between species -> test it
It probably reduces abundance, but this is uncertain. 
Linked to higher richness and to higher sd in richness, at the plot scale.

### by species
```{r}
regen_metrics_subplot$species %>% 
  select(plot,browsing,Species,matches("Height")) %>% 
  pivot_longer(cols=-c("plot","browsing","Species")) %>% 
  mutate(name=factor(name,levels = c("Height_increment_mn","Height_increment_cv",
                                     "Height_increment_mn_hclass2","Height_increment_mn_hclass3",
                                     "Height_increment_cv_hclass2","Height_increment_cv_hclass3"))) %>% 
  ggplot(aes(browsing,value,color=Species))+
  geom_point()+
  geom_smooth(method="gam",se=FALSE)+
  facet_wrap(~name,scale="free_y",ncol=2)+
  theme_bw()
```
Not clear. Plot not relevant at the plot scale.


## Difference between plot scale and subplot scale

```{r}


regen_metrics_subplot$all %>%
  left_join(
    regen_metrics_plot$all %>% select(plot, H_plot = H, n_seedling_plot = n_seedling),
    by = "plot"
  ) %>%
  ggplot(aes(color = plot)) +
  # Lines connecting each subplot point to its plot-level point
  geom_segment(aes(
    x = n_seedling, y = H,
    xend = n_seedling_plot, yend = H_plot
  ), alpha = 0.4, linewidth = 0.4) +
  # Subplot points (filled circles)
  geom_point(aes(n_seedling, H)) +
  # Plot-level points (triangles)
  geom_point(aes(n_seedling, H_plot), size = 3)

```

## Link between richness and abundance

```{r}
regen_metrics_plot$all %>% 
  ggplot(aes(n_sapling,richness))+
  geom_point()+
  geom_smooth(method="gam")
```


# Exploration of radiative budget 

## Radiative budget throughout the day
```{r}
nb_time=metric3d_all %>% select(plot,time) %>% group_by(plot) %>% summarise(ntime=n_distinct(time)) %>% 
  filter(ntime>2) %>% pull(plot)

metric3d_all %>% 
  filter(plot %in% nb_time) %>% 
  ggplot(aes(time,rb_mean,color=as.factor(slice_num)))+
  geom_point()+
  geom_line()+
  facet_grid(plot~subplot)+
  scale_x_continuous(breaks=seq(8, 18, 1))
```
12 or 13 is probably the maximum for rb at the plot scale.
Huge variability though at the subplot scale. For GER11 two peaks corresponding to gaps of light coming into the subplot.


# Link between light distribution and structure

## Between heterogeneity of light and abundance/richness/H/brwosing

```{r}
reg_ext=regen_metrics_subplot$all %>% 
  select(plot,subplot,
         richness,richness_sd,richness_seedling,richness_sapling,
         abundance,abundance_sd,abundance_cv,
         H,H_height,browsing) %>% unique() %>% 
  pivot_longer(cols=-c(plot,subplot),names_to = "y",values_to = "y_val")


reg_ext %>% 
  left_join(metric3d_all %>% 
              select(plot,subplot,slice_num,time,rb_sd,occupied_gf) %>% 
              # filter(slice_num==1) %>% 
              filter(time==12) %>% 
              filter(subplot!="plot"),by=c("plot","subplot")) %>% 
  filter(plot%in%metric3d_all$plot) %>% 
  ggplot(aes(rb_sd,y_val,col=plot))+
  geom_point()+geom_smooth(aes(color=NULL),method="gam")+
  scale_x_log10()+
  facet_wrap(slice_num~y,scale="free_y",nrow=4)



reg_ext=regen_metrics_plot$all %>% 
  select(plot,plot,
         richness,richness_sd,richness_seedling,richness_sapling,
         abundance,abundance_sd,abundance_cv,
         Height_increment_mn,Height_increment_cv,
         H,H_height,browsing) %>% unique() %>% 
  pivot_longer(cols=-c(plot),names_to = "y",values_to = "y_val")


reg_ext %>% 
  left_join(metric3d_all %>% 
              select(plot,subplot,slice_num,time,rb_cv,rb_mean,occupied_gf) %>% 
              # filter(slice_num==1) %>% 
              filter(time==12) %>% 
              filter(subplot=="plot"),by=c("plot")) %>% 
  filter(plot%in%metric3d_all$plot) %>% 
  ggplot(aes(rb_cv,y_val,col=plot))+
  geom_point()+geom_smooth(aes(color=NULL),method="lm")+
  scale_x_log10()+
  facet_wrap(slice_num~y,scale="free_y",nrow=4)
```

```{r}
regen_metrics_subplot %>% 
  select(plot,subplot,richness,abundance,n_sapling,H,Height_increment_mn_hclass2) %>% 
  group_by(plot,subplot) %>% 
  summarise(across(everything(), \(x) mean(x, na.rm = TRUE)),
            .groups = "drop") %>% 
  pivot_longer(cols=-c("plot","subplot")) %>% 
  mutate(slice_num=NA) %>% 
  filter(plot%in%metric3d_all$plot)->regen_ext
metric3d_all %>% 
  select(plot,subplot,slice_num,rb_mean,rb_sd,rb_cv) %>% 
  filter(slice_num%in%c(1,4)) %>% 
  pivot_longer(cols=c("rb_mean","rb_sd","rb_cv")) %>% 
  bind_rows(regen_ext) %>% 
  mutate(name=factor(name,level=c("rb_mean","rb_sd","rb_cv",
                                  "abundance","n_sapling","richness","H","Height_increment_mn_hclass2"))) %>% 
  ggplot(aes(value,plot))+
  geom_boxplot(aes(color=as.factor(slice_num)))+
  facet_wrap(~name,scale="free_x",ncol=3)+
  theme_bw()+
  labs(color="Height slice",x="Radiative budget metric or Structural metric",y="")



metric3d_all %>% 
  select(plot,subplot,time,slice_num,rb_mean,rb_sd,rb_cv) %>% 
  filter(slice_num%in%c(1,4)) %>% 
  filter(!time%in%c(11,13,15)) %>% 
  left_join(regen_ext %>% 
              pivot_wider(names_from = name,values_from = value) %>% 
              select(-slice_num),by=c("plot","subplot")) %>% 
  ggplot(aes(rb_mean,richness))+
  geom_point(aes(color=plot))+
  geom_smooth(method="gam")+
  facet_grid(time~slice_num,scale="free")+theme_bw()->p
p
```

# ACP regen light

## subplot

```{r}
regen_metrics_subplot %>% 
  select(-matches("Height"),-H_hclass3,-H_hclass4,-H_hclass5,-Species) %>% 
  unique() %>% 
  mutate(n_sapling=case_when(is.na(n_sapling)~0,
                             TRUE~n_sapling),
         H_hclass1=case_when(is.na(H_hclass1)~0,
                             TRUE~H_hclass1)) %>% 
  left_join(metric3d_all %>% 
              filter(subplot!="plot")%>% 
              filter(time==12&slice_num==1) %>% 
              select(plot,subplot,h_mean,h_sd,h_cv,rb_mean,rb_sd,rb_cv),
            by=c("plot","subplot")) %>%
  filter(plot%in%unique(metric3d_all$plot)) %>% 
  select(-plot,-subplot)->pca_df

library(factoextra)

pca_df <- pca_df[sapply(pca_df, is.numeric)]
pca_df <- na.omit(pca_df)

# Run PCA (center and scale recommended)
pca_res <- prcomp(pca_df[,1:10], center = TRUE, scale. = TRUE)

# Summary of variance explained
summary(pca_res)

# Correlation circle (variables factor map)
fviz_pca_var(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE                   # avoid label overlap
)

fviz_pca_biplot(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE                   # avoid label overlap
)


fviz_pca_var(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE,                   # avoid label overlap
  axes = c(1,3)
)

fviz_eig(pca_res, addlabels = TRUE, ylim = c(0, 50))

```
## Plot
```{r}
regen_metrics_plot %>% 
  select(-matches("Height"),-H_hclass3,-H_hclass4,-H_hclass5,-Species) %>% 
  unique() %>% 
  mutate(n_sapling=case_when(is.na(n_sapling)~0,
                             TRUE~n_sapling),
         H_hclass1=case_when(is.na(H_hclass1)~0,
                             TRUE~H_hclass1)) %>% 
  left_join(metric3d_all %>% 
              filter(subplot=="plot")%>% 
              filter(time==12&slice_num==1) %>% 
              select(plot,subplot,h_mean,h_sd,h_cv,rb_mean,rb_sd,rb_cv),
            by=c("plot")) %>%
  filter(plot%in%unique(metric3d_all$plot)) %>% 
  select(-plot)->pca_df

library(factoextra)

pca_df <- pca_df[sapply(pca_df, is.numeric)]
pca_df <- na.omit(pca_df)

# Run PCA (center and scale recommended)
pca_res <- prcomp(pca_df, center = TRUE, scale. = TRUE)

# Summary of variance explained
summary(pca_res)

# Correlation circle (variables factor map)
fviz_pca_var(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE                   # avoid label overlap
)

fviz_pca_biplot(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE                   # avoid label overlap
)


fviz_pca_var(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE,                   # avoid label overlap
  axes = c(1,3)
)

fviz_eig(pca_res, addlabels = TRUE, ylim = c(0, 50))

```


# Canopy species

```{r}

plot_df %>% 
  left_join(regen_metrics_plot %>% 
              select(plot,richness,abundance,H) %>% 
              unique())%>% 
  dplyr::select(-cutting_rate,-recruitment_rate,-n_cut,-cutting_rate,-n_recruits)->a


pca_df <-sapply(a[,2:22], as.numeric)
pca_df <- na.omit(pca_df) 

# Run PCA (center and scale recommended)
pca_res <- prcomp(pca_df, center = TRUE, scale. = TRUE)

# Summary of variance explained
summary(pca_res)

# Correlation circle (variables factor map)
fviz_pca_var(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE                   # avoid label overlap
)


fviz_pca_var(
  pca_res,
  col.var = "contrib",            # color by contribution
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE,                   # avoid label overlap
  axes = c(1,3)
)

fviz_eig(pca_res, addlabels = TRUE, ylim = c(0, 50))
```
## Effect of adults diversity on light heterogeneity / quantity

```{r}
plot_df %>% 
  select(plot,n_trees_alive,species_richness_alive) %>% 
  left_join(metric3d_all_plot %>% 
              filter(time==12,slice_num==1)%>% 
              select(plot,rb_mean,rb_sd,rb_cv) ) %>% 
  ggplot(aes(species_richness_alive,rb_sd))+
  geom_point()+
  geom_smooth(method="lm")


plot_df %>% 
  select(plot,n_trees_alive,species_richness_alive) %>% 
  left_join(regen_metrics_plot$all %>% 
              select(plot,richness,richness_seedling,richness_sapling) ) %>% 
  ggplot(aes(species_richness_alive,richness_sapling))+
  geom_point()+
  geom_smooth(method="lm")


plot_df %>% 
  select(plot,species_richness_alive) %>% 
  left_join(regen_metrics_plot$all %>% 
              select(plot,richness)) %>% 
  left_join(metric3d_all_plot %>%
              filter(time==12,slice_num==1)%>% 
              select(plot,rb_sd,rb_mean,rb_cv))%>%
  mutate(rb_sd=log(rb_sd),
         rb_mean=log(rb_mean)) %>%
  mutate(across(c(species_richness_alive, richness, rb_sd, rb_mean), scale))->sem_df

library(piecewiseSEM)
library(semPlot)
library(lavaan)

lavaan_model <- '
  richness ~ species_richness_alive + rb_sd + rb_mean
  rb_sd    ~ species_richness_alive
  rb_mean  ~ species_richness_alive
'

fit_lav <- sem(lavaan_model, data = sem_df)
summary(fit_lav)
# Plot
semPaths(
  object         = fit_lav,
  what           = "std",
  whatLabels     = "std",
  layout         = "tree2",
  edge.label.cex = 1.2,
  node.label.cex = 1.2,
  residuals      = FALSE,
  intercepts     = FALSE
)




# 1. Extract coefficients from piecewiseSEM
coef_table <- coefs(sem_model, standardize = "scale")

# 2. Add significance stars
coef_table$label <- with(coef_table, ifelse(P.Value < 0.001, paste0(round(Std.Estimate, 2), "***"),
                                    ifelse(P.Value < 0.01,  paste0(round(Std.Estimate, 2), "**"),
                                    ifelse(P.Value < 0.05,  paste0(round(Std.Estimate, 2), "*"),
                                    ifelse(P.Value < 0.1,   paste0(round(Std.Estimate, 2), "."),
                                                             round(Std.Estimate, 2))))))

# 3. Plot with custom labels
semPaths(
  object         = fit_lav,
  what           = "std",
  whatLabels     = "std",
  layout         = "tree2",
  edge.label.cex = 1.2,
  node.label.cex = 1.2,
  residuals      = FALSE,
  intercepts     = FALSE,
  edgeLabels     = coef_table$label  # inject piecewiseSEM labels with stars
)

regen_metrics_plot$all %>%
  select(plot,plot,richness) %>% 
  left_join(plot_df %>% 
              select(plot,species_richness_alive)) %>% 
  left_join(metric3d_all %>%
              filter(subplot=="plot") %>% 
              filter(time==12,slice_num==1)%>% 
              select(plot,subplot,rb_sd,rb_mean,rb_cv))%>%
  mutate(rb_sd=log(rb_sd),
         rb_mean=log(rb_mean)) %>%
  mutate(across(c(species_richness_alive, richness, rb_sd, rb_mean), scale))->sem_df
summary(lm(richness~rb_sd,data=sem_df))

regen_metrics_plot$all %>% 
  select(plot,plot,
         richness,richness_sd,richness_seedling,richness_sapling,
         abundance,abundance_sd,abundance_cv,
         Height_increment_mn,Height_increment_cv,
         H,H_height,browsing) %>% unique() %>% 
  pivot_longer(cols=-c(plot),names_to = "y",values_to = "y_val") %>% 
  left_join(metric3d_all %>% 
              select(plot,subplot,slice_num,time,rb_sd,rb_mean,occupied_gf) %>% 
              # filter(slice_num==1) %>% 
              filter(time==12) %>% 
              filter(subplot=="plot"),by=c("plot")) %>% 
  filter(plot%in%metric3d_all$plot) %>% 
  ggplot(aes(rb_mean,y_val,col=plot))+
  geom_point()+geom_smooth(aes(color=NULL),method="lm")+
  scale_x_log10()+
  facet_wrap(slice_num~y,scale="free_y",nrow=4)

regen_metrics_subplot$all %>% 
  select(plot,subplot,
         richness,richness_sd,richness_seedling,richness_sapling,
         abundance,abundance_sd,abundance_cv,
         Height_increment_mn,Height_increment_cv,
         H,H_height,browsing) %>% unique() %>% 
  pivot_longer(cols=-c(plot,subplot),names_to = "y",values_to = "y_val") %>% 
  left_join(metric3d_all %>% 
              select(plot,subplot,slice_num,time,rb_sd,rb_mean,occupied_gf) %>% 
              # filter(slice_num==1) %>% 
              filter(time==12) %>% 
              filter(subplot!="plot"),by=c("plot","subplot")) %>% 
  filter(plot%in%metric3d_all$plot) %>% 
  ggplot(aes(rb_mean,y_val,col=plot))+
  geom_point()+geom_smooth(aes(color=NULL),method="lm")+
  scale_x_log10()+
  facet_wrap(slice_num~y,scale="free_y",nrow=4)




library(tidyverse)
library(boot)

# model function
boot_fn <- function(data, indices){

  d <- data[indices, ]

  fit <- lm(richness ~ rb_sd, data = d)

  coef(fit)
}

set.seed(123)

boot_res <- boot(
  data = sem_df,
  statistic = boot_fn,
  R = 10000
)

boot_res

boot_slopes <- tibble(
  intercept = boot_res$t[,1],
  slope = boot_res$t[,2]
)

ggplot(boot_slopes, aes(slope)) +
  geom_histogram(bins = 40) +
  geom_vline(xintercept = coef(lm(richness ~ rb_sd, data = sem_df))[2],
             color = "red")

boot.ci(boot_res, type = "perc", index = 2)
```


