---
title: "Fig_paper"
author: "Anne Baranger"
date: "2026-06-24"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE,message=FALSE)
library(targets)
lapply(c("stringr","tidyr","dplyr","tibble","tidyverse","forcats", "purrr",# data analysis
         "ggplot2","viridis", "ggridges","wesanderson","ggridges","ggrepel","ggnewscale","ggside", # plot
         "factoextra", "nlme","modi","rstan","shinystan","boot", "rstatix","lme4","multcompView","vegan",# stat & models
         "rworldmap" # geographic data #terre, sf
         ),require,character.only=TRUE)
theme_set(theme_bw()) # ggplot theme
common_theme <- theme(
  # legend.position = "none",
  strip.background = element_rect(fill = "white",
                                        colour = "gray12"),
  panel.background = element_rect(fill = "white",
                                  colour = "gray12",
                                  size = 0.5, linetype = "solid"),
  plot.title = element_text(size = 12),
  axis.title = element_text(size = 10),
  axis.text = element_text(size = 10),
  strip.text = element_text(size =10),
  plot.margin = unit(c(0, 0.2, 0, 0.2), "cm")  # Example margin; adjust as needed
)

loadRData <- function(fileName){
  #loads an RData file, and returns it
  load(fileName)
  get(ls()[ls() != "fileName"])
}

```

# Data preparation 

```{r}
# load targets objects
tar_load(c(metric3d_all, metric3d_all_unsliced, metric3d_all_seg,
           regen_metrics_plot, regen_metrics_subplot, plot_df, H_ext_all,
           regen_cat, regen_cat_subplot, var_meta, gini_all))

# def of var_meta
var_meta=tribble(
  ~variable,                ~type,                            ~label,                                 ~abrv,
  "richness",               "Composition",                    "Richness \ntotal",                     "Richn. tot.",
  "richness_sapling",       "Composition",                    "Richness \nsaplings",                  "Richn. sap.", 
  "H",                      "Composition",                    "Shannon \nindex",                      "H",
  "cwm_shade",              "Composition",                    "Shade tolerance \n (CWM)",             "Shade tol.",
  "abundance",              "Quantity",                       "Abundance",                            "Ab.",
  "abundance_cv",           "Quantity",                       "Abundance CV",                         "Ab. (cv)",
  "n_sapling",              "Quantity",                       "Abundance of \nsaplings",              "Ab. sap.",
  "h_mean",                 "Quantity",                       "Mean \nheight",                        "h mean",
  "pad_tot",                "Quantity",                       "PAD \ntotal",                          "PAD tot.",
  "empty_gf",               "Quantity",                       "Gap fraction",                         "Gap frac.",
  "Height_increment_mn",    "Quantity",                       "Height \nincrement",                   "h incr.",
  "H_height",               "Complexity",                     "Vertical \ndistribution",              "Vert. dist.",
  "h_cv",                   "Complexity",                     "Height \nvariability",                 "h var.",
  "pad_cv",                 "Complexity",                     "PAD \nvariability",                    "PAD (cv)",
  "ri",                     "Complexity",                     "Rumple index",                         "RI",
  "fhd",                    "Complexity",                     "Foliage height \ndiversity",           "FHD",
  "browsing",               "Other",                          "Browsing",                             "Brows.",
  "lrb_mean",               "Light quantity",                 "Radiative budget \nmean (logged)",     "RB (log)",
  "rb_mean",                "Light quantity",                 "Radiative\nbudget mean",               "RB",
  "rb_cv",                  "Light horizontal heterogeneity", "Radiative\nbudget CV",                 "RB (cv)",
  "rb_sd",                  "Light horizontal heterogeneity", "Radiative\nbudget SD",                 "",
  "ext_ref_agg",            "Light quantity",                 "Mean extinction \n at 4.5m",           "Ext. mean",
  "ext_ref_cv",             "Light horizontal heterogeneity", "Mean extinction \n at 4.5m CV",        "",
  "ext_ref_sd",             "Light horizontal heterogeneity", "Mean extinction \n at 4.5m SD",        "",
  "padAbove_agg",           "PAD",                            "PAD above \nregeneration",             "",
  "padAbove_cv",            "PAD",                            "PAD above \nCV",                       "",
  "padAbove_sd",            "PAD",                            "PAD above \nSD",                       "",
  "padTot_agg",             "PAD",                            "PAD total",                            "",
  "padTot_cv",              "PAD",                            "PAD total CV",                         "",
  "padTot_sd",              "PAD",                            "PAD total sd",                         "",
  "rb_ref_agg",             "Light quantity",                 "RB. at 4.5m",                          "",
  "rb_ref_cv",              "Light horizontal heterogeneity", "RB. at 4.5m \nCV",                     "",
  "rb_ref_sd",              "Light horizontal heterogeneity", "RB. at 4.5m \nSD",                     "",
  "slope_cross_agg",        "Light vertical heterogeneity",   "Light profile slope \nat ext = ",      "",
  "slope_cross_cv",         "Light vertical heterogeneity",   "Light profile slope, CV \nat ext = ",  "",
  "slope_cross_sd",         "Light vertical heterogeneity",   "Light profile slope, SD \nat ext = ",  "",
  "slope_ref_agg",          "Light vertical heterogeneity",   "Light profile slope \nat 4.5m",        "Light profile \nslope",
  "slope_ref_cv",           "Light vertical heterogeneity",   "Light profile slope, CV\nat 4.5m",     "",
  "slope_ref_sd",           "Light vertical heterogeneity",   "Light profile slope, SD\nat 4.5m",     "",
  "z_cross_agg",            "Light",                          "Height\nat ext = ",                    "",
  "z_cross_cv",             "Light",                          "Height, CV\nat ext = ",                "",
  "z_cross_sd",             "Light",                          "Height, SD\nat ext = ",                "",
  "zPAD50_agg",             "PAD",                            "Height at\n 50% of max(PAD)",          "",
  "zPAD50_cv",              "PAD",                            "Height (CV) at\n 50% of max(PAD)",     "",
  "zPAD50_sd",              "PAD",                            "Height (SD) at\n 50% of max(PAD)",     "",
  "gini_h_rb",              "Light horizontal heterogeneity", "Gini index",                           "Gini",
  "gini_v_rb",              "Light vertical heterogeneity",   "Gini vert.",                           "",
  "bl_diff_abs",            "Light vertical heterogeneity",   "Big leaf dif.",                        "Big leaf dif.",
  "bl_slope_abs",           "Light vertical heterogeneity",   "Big leaf \nslope dif.",                "",
  "PC1",                    "Light",                          "Light PCA axis 1",                     "",
  "n_trees_alive",          "Adult",                          "Number of\n living trees",             "",
  "species_richness_alive", "Adult",                          "Adult richness",                       "",
  "mean_dbh_alive_cm",      "Adult",                          "Mean diameter",                        "",
  "mean_height_alive_m",    "Adult",                          "Mean height",                          "",
  "mortality_rate",         "Adult",                          "Mortality",                            "",
  "basal_area_alive_m2",     "Adult",                         "Total basal area",                     ""
) 

# all tls metrics
tls_metrics = rbind(metric3d_all,metric3d_all_unsliced) %>% 
  left_join(H_ext_all,by=c("plot","subplot","time"))%>% 
  left_join(gini_all,by=c("plot","subplot","time")) %>% 
  mutate(lrb_mean = log(rb_mean))

# inventory 
forest_data_plot=regen_metrics_plot$all %>% 
  left_join(regen_cat) %>% 
  left_join(plot_df)%>%
  mutate(across(c(richness_sapling, n_sapling), ~ replace_na(.x, 0))) %>% 
  mutate(subplot="plot")
forest_data_subplot=regen_metrics_subplot$all %>% 
  left_join(regen_cat_subplot)%>% 
  left_join(plot_df) %>% 
  unique()%>%
  mutate(across(c(richness_sapling, n_sapling), ~ replace_na(.x, 0)))

# function to create dataset
prep_data<-function(df,df_2=NULL,id_cols,plot_level,t=12,s=0,tg=0.9){
  df_f = df %>% 
    filter(if (plot_level=="plot") subplot == "plot" else subplot != "plot",
           time == t,slice_num==s,
           if ("target" %in% colnames(df)) target == .env$tg else TRUE) %>% 
    dplyr::select(any_of(c( c("plot","subplot"),id_cols))) %>% 
    unique()
  if(!is.null(df_2)){
    df_filtered_2 = df_2 %>% 
      filter(
        if (plot_level == "plot") subplot == "plot" else subplot != "plot",
        if ("time" %in% colnames(df_2)) time == .env$t else TRUE,
        if ("slice_num" %in% colnames(df_2)) slice_num == .env$s else TRUE
      ) %>% 
      dplyr::select(any_of(c( c("plot","subplot"),id_cols))) %>% 
      unique()
    df_f=df_f%>% 
      left_join(df_filtered_2,by=c("plot","subplot"))
  }
  return(df_f)
}

save_fig <- function(plot, name, width, height) {
  ggsave(file.path("figures", paste0(name, ".png")), plot,
         width = width, height = height,units="cm", dpi = 600, bg = "white")
  ggsave(file.path("figures", paste0(name, ".pdf")), plot,
         width = width, height = height,units="cm", device = cairo_pdf)
}
```

# Regeneration structure pattern


```{r, fig.width=10, fig.height=6}
# varaibles for the pca
var_pca=c('richness','H','richness_sapling','cwm_shade','abundance','n_sapling','Height_increment_mn','H_height','empty_gf','pad_tot','pad_cv','ri','fhd')
var_tls=c('empty_gf','pad_tot','pad_cv','ri','fhd')

# create dataset 
pca_data_plot= prep_data(tls_metrics,forest_data_plot,var_pca,plot_level = "plot")
pca_data_subplot= prep_data(tls_metrics,forest_data_subplot,var_pca,plot_level = "subplot")


# compute pca on plots
pca_df   <- pca_data_plot %>% 
  dplyr::select(-plot,-subplot) %>% 
  mutate(across(everything(), as.numeric)) %>%
  drop_na()
pca_res  <- prcomp(pca_df, center = TRUE, scale. = TRUE)

# predictions
pca_tot <- bind_rows(pca_data_plot, pca_data_subplot) %>%
  left_join(regen_cat) %>%
  bind_cols(as_tibble(predict(pca_res, .[, colnames(pca_df)])[, 1:2]))
pts <- pca_tot %>%
  filter(subplot=="plot") %>%
  mutate(plot_level = "Plot")

# arrows coordinated
v <- get_pca_var(pca_res)
var_coord <- tibble(
  variable = rownames(v$coord),
  PC1      = v$coord[, 1],
  PC2      = v$coord[, 2],
  contrib  = rowSums(v$contrib[, 1:2]),
  cos2     = rowSums(v$cos2[, 1:2])
  ) %>%
  left_join(var_meta, by = "variable") %>%
  mutate(type = replace_na(type, "Other"),
         Source = case_when(variable%in%var_tls~"Lidar-derived",
                            TRUE~"Inventory-derived"))


# plot

# axis
all_vals <- c(pts$PC1, pts$PC2, var_coord$PC1 * 5, var_coord$PC2 * 5)
brks <- pretty(range(all_vals, na.rm = TRUE), n = 5)
lims <- c(floor(range(all_vals, na.rm = TRUE)[1]),ceiling(range(all_vals, na.rm = TRUE)[2]))   # limits snap to the outer breaks

# labels
scale_f <- 5
lab_df <- var_coord %>%
  mutate(x = PC1 * scale_f *1.01 ,
         y = PC2 * scale_f *1.01 ,
         abrv, type,.keep="none")
arrow_pts <- var_coord %>%
  crossing(t = seq(0.05, 1, length.out = 7)) %>%
  mutate(x = PC1 * scale_f * t,
         y = PC2 * scale_f * t,
         abrv = "", type,.keep="none")

repel_df <- bind_rows(lab_df, arrow_pts)



# kw tests
test_df <- pts %>%
  filter(!is.na(reg_category)) %>%
  mutate(reg_category = droplevels(factor(reg_category))) %>%
  pivot_longer(c(PC1, PC2), names_to = "axis", values_to = "score") %>%
  group_by(axis) %>%
  kruskal_test(score ~ reg_category)%>%
  add_significance("p") %>%   # adds stars column "p.signif" (ns, *, **, ***)
  mutate(label = paste0(
    "KW :\n ",
    ifelse(p < 0.001, "p < 0.001", paste0("p = ", formatC(p, format = "f", digits = 3))),
    " ", ifelse(p.signif == "ns", "", p.signif)
  ))


# plots
ggplot() +
  geom_vline(xintercept = 0,colour="darkgrey",linewidth=0.2)+geom_hline(yintercept = 0,colour="darkgrey",linewidth=0.2)+
  geom_point(
    data = pts,
    aes(PC1, PC2, shape = plot_level, size = plot_level, color = reg_category),
    size=0.5
  ) +
  geom_xsideboxplot(
    data = pts %>% filter(!is.na(reg_category)),
    aes(x = PC1, y = reg_category, fill = reg_category),
    whisker.linewidth =0.2,median.linewidth =0.2,box.linewidth=0.2,
    orientation = "y", outliers = FALSE
  ) +
  geom_ysideboxplot(
    data = pts %>% filter(!is.na(reg_category)),
    aes(y = PC2, x = reg_category, fill = reg_category),
    whisker.linewidth =0.2,median.linewidth =0.2,box.linewidth=0.2,
    orientation = "x", outliers = FALSE
  )+
  geom_xsidetext(
    data = filter(test_df, axis == "PC1"),
    aes(x = 2.7, y = "Medium", label = label),
    size = 7.5,size.unit="pt",
    vjust=-0.02,
    inherit.aes = FALSE
  ) +
  geom_ysidetext(
    data = filter(test_df, axis == "PC2"),
    aes(x = "Medium", y = 2.5, label = label),
    size = 7.5, size.unit="pt", inherit.aes = FALSE
  ) +
  scale_fill_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  scale_shape_manual(name = "Spatial scale", values = c("Plot" = 19, "Subplot" = 1)) +
  scale_size_manual(name = "Spatial scale", values = c("Plot" = 2.5, "Subplot" = 1.5)) +
  scale_color_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  guides(color="none",size="none",shape="none")+
  new_scale_color() +
  geom_segment(
    data = var_coord,
    aes(x = 0, y = 0, xend = PC1*5, yend = PC2*5, color = type,linewidth = Source),
    arrow = arrow(length = unit(0.10, "cm")), lineend = "round"
  ) +
  geom_text_repel(
    data = repel_df,
    aes(x = x, y = y, label = abrv, color = type),
    size = 8*0.353,
    # box.padding = 0.4,        # space around each label
    # point.padding = 0.15,     # space kept from the obstacle points
    # # min.segment.length = 0.3,   # always draw a leader line if a label moves
    segment.color = "grey50",
    segment.size = 0.1,
    # max.overlaps = Inf,       # never drop labels
    # force = 2,                # push harder
    seed = 58,                # reproducible layout
    show.legend = FALSE
  ) +
  scale_color_brewer(name = "Variable type", palette = "Dark2") +
  scale_linewidth_manual(name = "Data source", 
                         values = c("Lidar-derived" = 0.45, "Inventory-derived" = 0.2)) +
  theme_minimal() +
  theme(
    # --- main panel: black frame ---
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
    panel.grid = element_line(linewidth = 0.2),
    # --- side panels
    ggside.panel.scale = 0.15,          # smaller = thinner strips (try 0.10–0.20)
    ggside.panel.grid = element_blank(),
    ggside.panel.border = element_blank(),
    ggside.panel.background = element_blank(),
    ggside.axis.text  = element_blank(),
    ggside.axis.ticks = element_blank(),

    legend.position = "bottom",
    legend.location = "pannel",
    legend.direction = "vertical",
    legend.justification = "left",
    legend.key.spacing = unit(0,"pt"),
    legend.text=element_text(margin=margin(l=0),size=8),

    aspect.ratio = 1,
    text=element_text(size=8)
  )+
  labs(
    x = paste0("PCA axis 1 (", round(summary(pca_res)$importance["Proportion of Variance","PC1"]*100, 1), "% of variation)"),
    y = paste0("PCA axis 2 (", round(summary(pca_res)$importance["Proportion of Variance","PC2"]*100, 1), "% of variation)")
  ) + 
  scale_x_continuous(limits = lims, breaks = brks) +
  scale_y_continuous(limits = lims, breaks = brks) ->c
c



ggsave(file.path("figures", paste0("mainfig1_PCA_regen", ".png")), c,
         width=8.5,height=11,units="cm", dpi = 600, bg = "white",scale = 1.2)

# save_fig(c,"mainfig1_PCA_regen",width=16,height=18)
```


## correlation table 

```{r}
# all pairwise correlation tests
pairs <- combn(names(pca_df), 2, simplify = FALSE)

cor_tab <- map_dfr(pairs, function(v) {
  ct <- cor.test(pca_df[[v[1]]], pca_df[[v[2]]], method = "pearson", exact = FALSE)
  tibble(var1 = v[1], var2 = v[2],
         rho = unname(ct$estimate),
         p = ct$p.value)
}) %>%
  mutate(
    p_adj  = p.adjust(p, method = "bonferroni"),
    signif = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE          ~ "ns"
    )
  ) %>%
  arrange(desc(abs(rho)))

cor_tab
```

- hard to distinguish different regeneration clusters
- variable type are not necessarly colinear, can even be non correlated
- field-derived and lidar-derived variables roughly non correlated, catching different dimensions
- prediction at the subplot scale = add variability


# Light distribution descriptors - plot level

## PCA of light descriptors

```{r}
# variables of the pca
var_pca_light=c("lrb_mean","ext_ref_agg","rb_cv","gini_h_rb","bl_diff_abs","slope_ref_agg","reg_category")

# create dataset
light_pca <- prep_data(tls_metrics,forest_data_plot,var_pca_light,plot_level = "plot",t=12,s=0)  

# compute pca
df_pca=light_pca[,var_pca_light] %>% select(-reg_category)#"rb_ref_agg","padAbove_cv","rb_cv",
pca_light <- prcomp(df_pca,center = TRUE, scale. = TRUE)

light_pca_pred = cbind(light_pca,predict(pca_light,df_pca)[,c(1,2)])


# get arrows
var <- get_pca_var(pca_light)
var_coord <- as.data.frame(var$coord[, 1:2])
colnames(var_coord) <- c("PC1", "PC2")
var_coord$variable <- rownames(var_coord)
var_coord$contrib <- var$contrib[, 1] + var$contrib[, 2]
var_coord$cos2 <- var$cos2[, 1] + var$cos2[, 2]

var_coord=var_coord %>%
  left_join(var_meta,by="variable") %>% 
  mutate(label=case_when(variable=="slope_cross_agg"~paste0(label," 0.9"),
                         variable=="z_cross_agg"~paste0(label," 0.9"),
                         TRUE~label),
         type=if_else(is.na(type),"Other",type))

# plot
all_vals <- c(pts$PC1, pts$PC2, var_coord$PC1 * 5, var_coord$PC2 * 5)
brks <- pretty(range(all_vals, na.rm = TRUE), n = 5)
lims <- c(floor(range(all_vals, na.rm = TRUE)[1]),ceiling(range(all_vals, na.rm = TRUE)[2]))   # limits snap to the outer breaks


scale_f <- 5
lab_df <- var_coord %>%
  mutate(x = PC1 * scale_f *1.01 ,
         y = PC2 * scale_f *1.01 ,
         abrv, type,.keep="none")
arrow_pts <- var_coord %>%
  crossing(t = seq(0.05, 1, length.out = 7)) %>%
  mutate(x = PC1 * scale_f * t,
         y = PC2 * scale_f * t,
         abrv = "", type,.keep="none")

repel_df <- bind_rows(lab_df, arrow_pts)


ggplot() +
  geom_point(
    data = light_pca_pred,
    aes(PC1, PC2, color = reg_category),
    size=0.8 #,show.legend = FALSE
  ) +
  scale_color_brewer(name = "Regeneration cover", palette = "YlGn",na.value = "grey") +
  
  new_scale_color() +
  
  geom_segment(
    data = var_coord,
    aes(x = 0, y = 0, xend = PC1*5, yend = PC2*5,  color = type), #alpha = contrib,
    arrow = arrow(length = unit(0.10, "cm")),
    lineend = "round",linewidth=0.3
  ) +
  geom_text_repel(
    data = repel_df,
    aes(x = x, y = y, label = abrv,  color = type), #alpha = contrib,
    size = 8*0.353, show.legend = FALSE
  ) +
  scale_fill_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  scale_color_brewer(name = "Variable type", palette = "Dark2") +
  guides(color="none")+
  
  coord_equal() +
  theme_minimal() +
  guides(
    alpha = "none",
    color = guide_legend(override.aes = list(alpha = 1))
  ) +
  labs(
    x = paste0("PCA axis 1 (", round(summary(pca_light)$importance["Proportion of Variance","PC1"]*100, 1), "% of variation)"),
    y = paste0("PCA axis 2 (", round(summary(pca_light)$importance["Proportion of Variance","PC2"]*100, 1), "% of variation)")
  )+
  scale_y_continuous(breaks = seq(-10, 10, by = 2))+
  scale_x_continuous(breaks = seq(-10, 10, by = 2))+
  theme(
    # --- main panel: black frame ---
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
    panel.grid = element_line(linewidth = 0.2),
    # --- side panels
    # ggside.panel.scale = 0.15,          # smaller = thinner strips (try 0.10–0.20)
    # ggside.panel.grid = element_blank(),
    # ggside.panel.border = element_blank(),
    # ggside.panel.background = element_blank(),
    # ggside.axis.text  = element_blank(),
    # ggside.axis.ticks = element_blank(),

    legend.position = "bottom",
    legend.location = "pannel",
    legend.direction = "vertical",
    legend.justification = "left",
    legend.key.spacing = unit(0,"pt"),
    legend.text=element_text(margin=margin(l=0),size=8),

    aspect.ratio = 1,
    text=element_text(size=8)
  ) + 
  scale_x_continuous(limits = lims, breaks = brks) +
  scale_y_continuous(limits = lims, breaks = brks)-> pca_light

pca_light


# ggsave(file.path("figures", paste0("mainfig2_PCA_light", ".png")), pca_light,
#          width=8.5,height=11,units="cm", dpi = 600, bg = "white",scale = 1.1)
```

## comparative effects of plant material and light distribution on the structure of regeneration

```{r}
light_pca_comp <- prep_data(tls_metrics,forest_data_plot,
                            c(var_pca_light,"padAbove_agg","padAbove_cv"),
                            plot_level = "plot",t=12,s=0)  %>% 
  pivot_longer(cols = -c("plot","subplot","reg_category")) %>%
  filter(!is.na(reg_category)) %>%
  mutate(name=factor(name,levels = c(var_pca_light,"padAbove_agg","padAbove_cv")),
         reg_category = factor(reg_category)) %>%
  group_by(name) %>% mutate(value = as.numeric(scale(value))) %>% ungroup() %>%
  left_join(var_meta, by = c("name" = "variable")) %>% 
  arrange(name) %>%        # type's level order drives it
  mutate(label=case_when(name=="slope_cross_agg"~paste0(label," 0.9"),
                         name=="z_cross_agg"~paste0(label," 0.9"),
                         TRUE~label),
         label = fct_inorder(label))

# 
# cld_one <- function(df) {
#   lev <- levels(droplevels(df$reg_category))
#   if (length(lev) < 2) return(tibble(reg_category = lev, letter = ""))
#   
#   kw <- kruskal.test(value ~ droplevels(reg_category), data = df)
#   if (is.na(kw$p.value) || kw$p.value > 0.05)
#     return(tibble(reg_category = lev, letter = "a"))   # no differences
#   
#   dt <- dunn_test(df, value ~ reg_category, p.adjust.method = "BH")
#   pv <- setNames(dt$p.adj, paste(dt$group1, dt$group2, sep = "-"))
#   l  <- multcompLetters(pv)$Letters
#   tibble(reg_category = names(l), letter = unname(l))
# }
# 
# cld_df <- light_pca_comp %>%
#   group_by(type, label) %>%
#   group_modify(~ cld_one(.x)) %>%
#   ungroup() %>%
#   mutate(reg_category = factor(reg_category, levels = levels(light_pca_comp$reg_category))) %>%
#   # assign y coordinates
#   left_join(
#     light_pca_comp %>% group_by(type, label, reg_category) %>%
#       summarise(y = max(value, na.rm = TRUE), .groups = "drop"),
#     by = c("type", "label", "reg_category")
#   )

kw_df <- light_pca_comp %>%
  mutate(type=case_when(type=="PAD"~"Plant area density\n distribution",
                        TRUE~type)) %>% 
  group_by(type, label) %>%
  summarise(
    p = if (n_distinct(reg_category) < 2) NA_real_
        else kruskal.test(value ~ droplevels(reg_category))$p.value,
    y = max(value, na.rm = TRUE),          # top of the highest point for this variable
    .groups = "drop"
  ) %>%
  # mutate(p = p.adjust(p, method = "BH")) %>%
  mutate(
    stars = case_when(
      is.na(p)  ~ "",
      p < 0.001 ~ "***",
      p < 0.01  ~ "**",
      p < 0.05  ~ "*",
      p < 0.1   ~ "(*)",
      TRUE      ~ "ns"                     # use "" to show nothing when not significant
    )
  )

# kw_df

light_pca_comp %>% 
  mutate(type=case_when(type=="PAD"~"Plant area density\n distribution",
                        TRUE~type)) %>% 
  ggplot(aes(label, value, fill = reg_category)) +
  geom_boxplot(position = position_dodge(width = 0.8, preserve = "single"),
               width=0.5,outlier.shape = NA) +
  geom_point(aes(color=reg_category),
             position = position_jitterdodge(jitter.width = 0.15,
                                             dodge.width = 0.8,
                                             seed = 1),
             size = 1, alpha = 0.8, shape = 16,
             show.legend = FALSE) +
    geom_text(data = kw_df,
            aes(x = label, y = 2.5, label = stars),
            inherit.aes = FALSE, size = 3, vjust = 0) +
  # geom_text(data = cld_df,
  #           aes(x = label, y = y + 0.25, label = letter, group = reg_category),
  #           position = position_dodge(width = 0.8),
  #           inherit.aes = FALSE, size = 3, vjust = 0) +
  scale_fill_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  scale_color_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  facet_grid(~type, scales = "free_x") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.12)))+
  labs(y="Scaled values",x="")+
  theme_minimal(base_size = 9)+
  theme(legend.position = "none")->a
a

cowplot::plot_grid(pca_light+theme(legend.position="left"),a,ncol=2,align = "h", axis = "tb",
                   rel_widths = c(1, 1.5))


```


## correlation table 

```{r}
# all pairwise correlation tests
pairs <- combn(names(df_pca), 2, simplify = FALSE)

cor_tab <- map_dfr(pairs, function(v) {
  ct <- cor.test(df_pca[[v[1]]], df_pca[[v[2]]], method = "pearson", exact = FALSE)
  tibble(var1 = v[1], var2 = v[2],
         rho = unname(ct$estimate),
         p = ct$p.value)
}) %>%
  mutate(
    p_adj  = p.adjust(p, method = "bonferroni"),
    signif = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE          ~ "ns"
    )
  ) %>%
  arrange(desc(abs(rho)))

cor_tab
```

# Light distribution effect on regeneration structure - plot level

```{r}
explained <- c("richness","richness_sapling","H","cwm_shade","abundance","n_sapling","Height_increment_mn","empty_gf","H_height","fhd","ri","pad_cv")
explaining <- c("lrb_mean","gini_h_rb","slope_ref_agg","bl_diff_abs")

light_effect <- prep_data(tls_metrics,forest_data_plot,c(explained,explaining),plot_level="plot",
                          t=12,s=0,tg=0.9) %>% 
  mutate(slope_ref_agg=-slope_ref_agg)

order=c(var_meta$label[match(explaining,var_meta$variable)],
        var_meta$label[match(explained,var_meta$variable)])


std_coef <- function(d, y, x) { #, covar = "browsing"
  d <- d %>%
    dplyr::select(all_of(c(y, x))) %>% #, covar
    drop_na() %>%
    mutate(across(where(is.numeric), ~ as.numeric(scale(.))))  # factors (e.g. browsing) left alone
  f <- reformulate(c( x), response = y) #covar,
  unname(coef(lm(f, data = d))[x])
}

# --- bootstrap distribution of that coef ---
boot_coef <- function(d, y, x, B = 1000) { #, covar = "browsing"
  d <- d %>% dplyr::select(all_of(c(y, x))) %>% drop_na() %>% # , covar
      mutate(across(where(is.numeric), ~ as.numeric(scale(.)))) 
  n <- nrow(d)
  f <- reformulate(c( x), response = y) # covar,
  map_dbl(seq_len(B), function(b) {
    db <- d[sample(n, n, replace = TRUE), ]  # <- move this line out of the loop to scale once
    unname(coef(lm(f, data = db))[x])
  })
}

results <- expand_grid(explained = explained, explaining = explaining) %>%
  mutate(
    estimate = map2_dbl(explained, explaining, ~ std_coef(light_effect, .x, .y)),
    draws    = map2(explained, explaining, ~ boot_coef(light_effect, .x, .y, B = 1000))
  )


boot_long <- results %>%
  dplyr::select(explained, explaining, draws) %>%
  unnest_longer(draws) %>%
  rename(coef = draws)

boot_stats <- boot_long %>%
  group_by(explained, explaining) %>%
  summarise(
    lo   = unname(quantile(coef, 0.10)),
    lo5  = unname(quantile(coef, 0.25)),
    med  = unname(quantile(coef, 0.50)),
    hi5  = unname(quantile(coef, 0.75)),
    hi   = unname(quantile(coef, 0.90)),
    nrep = dplyr::n(),
    # test bilatéral : 2 x la proportion de tirages du "mauvais" côté de 0,
    # bornée par la résolution du bootstrap 1/(B+1)
    p_boot = max(2 * min(mean(coef <= 0), mean(coef >= 0)), 1 / (nrep + 1)),
    .groups = "drop"
  )



boot_stats %>% filter(explaining !="bl_diff_abs") %>% 
  mutate(
    p_class = cut(p_boot,
                  breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                  labels = c("< 0.001", "< 0.01", "< 0.05", "< 0.1", "ns"),
                  right = FALSE)
  ) %>%
  pivot_longer(cols = c("explained", "explaining")) %>%
  left_join(var_meta, by = c("value" = "variable")) %>% 
  dplyr::select(-value,-abrv) %>%
  pivot_wider(names_from = "name", values_from = c("label", "type")) %>%
  mutate(
    type_explained   = factor(type_explained,
                              levels = c("Composition", "Quantity", "Complexity")),
    label_explained  = factor(label_explained,  levels = rev(order)),
    label_explaining = factor(label_explaining, levels = order)
  ) %>%
  ggplot(aes(x = med, y = label_explained)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = .2, colour = "grey70") +
  geom_linerange(aes(xmin = lo,  xmax = hi,  colour = p_class), linewidth = .2) +
  geom_linerange(aes(xmin = lo5, xmax = hi5, colour = p_class), linewidth = .5) +
  geom_point(aes(colour = p_class), size = 1.3) +
  scale_colour_manual(
    values = c("< 0.001" = "#08306b", "< 0.01" = "#2171b5",
               "< 0.05"  = "#6baed6", "< 0.1"  = "#c6dbef",
               "ns"      = "grey78"),
    drop = FALSE, name = "p (bootstrap)"
  ) +
  facet_grid(type_explained ~ label_explaining, scales = "free_y", space = "free_y") +
  labs(x = "Standardized coefficient", y = NULL) +
  theme_bw(base_size = 9) +
  theme(
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
    panel.grid = element_line(linewidth = 0.2),
    strip.text.y     = element_text(angle = 0),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom",
    text=element_text(size=8)
  )->light_eff


# ggsave(file.path("figures", paste0("mainfig3_light_eff", ".png")), light_eff,
#          width=16,height=9,units="cm", dpi = 600, bg = "white",scale = 1.2)
```

# From plot to subplot scale

## PCA of light distribution at subplot scale

```{r}
var_pca_light=c("lrb_mean","ext_ref_agg","rb_cv","gini_h_rb","bl_diff_abs","slope_ref_agg","reg_category")

# create dataset
light_pca_s <- prep_data(tls_metrics,forest_data_subplot,var_pca_light,plot_level = "subplot",t=12,s=0)  

# compute pca
df_pca_s=light_pca_s[,var_pca_light] %>% select(-reg_category)#"rb_ref_agg","padAbove_cv","rb_cv",
pca_light_s <- prcomp(df_pca_s,center = TRUE, scale. = TRUE)

light_pca_pred_s = cbind(light_pca_s,predict(pca_light_s,df_pca_s)[,c(1,2)])


# get arrows
var <- get_pca_var(pca_light_s)
var_coord <- as.data.frame(var$coord[, 1:2])
colnames(var_coord) <- c("PC1", "PC2")
var_coord$variable <- rownames(var_coord)
var_coord$contrib <- var$contrib[, 1] + var$contrib[, 2]
var_coord$cos2 <- var$cos2[, 1] + var$cos2[, 2]

var_coord=var_coord %>%
  left_join(var_meta,by="variable") %>% 
  mutate(label=case_when(variable=="slope_cross_agg"~paste0(label," 0.9"),
                         variable=="z_cross_agg"~paste0(label," 0.9"),
                         TRUE~label),
         type=if_else(is.na(type),"Other",type))

# plot
ggplot() +
  geom_point(
    data = light_pca_pred_s,
    aes(PC1, PC2, color = reg_category),
    size=2.5 #,show.legend = FALSE
  ) +
  scale_color_brewer(name = "Regeneration cover", palette = "YlGn",na.value = "grey") +
  
  new_scale_color() +
  
  geom_segment(
    data = var_coord,
    aes(x = 0, y = 0, xend = PC1*5, yend = PC2*5,  color = type), #alpha = contrib,
    arrow = arrow(length = unit(0.20, "cm")),
    lineend = "round"
  ) +
  geom_text_repel(
    data = var_coord,
    aes(x = PC1*5, y = PC2*5, label = label,  color = type), #alpha = contrib,
    size = 3, show.legend = FALSE
  ) +
  # geom_xsideboxplot(
  #   data = light_pca_pred_s %>% filter(!is.na(reg_category)),
  #   aes(x = PC1, y = reg_category, fill = reg_category),
  #   orientation = "y", outliers = FALSE
  # ) +
  # geom_ysideboxplot(
  #   data = light_pca_pred_s %>% filter(!is.na(reg_category)),
  #   aes(y = PC2, x = reg_category, fill = reg_category),
  #   orientation = "x", outliers = FALSE
  # )+
  scale_fill_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  scale_color_brewer(name = "Variable type", palette = "Dark2") +
  guides(color="none")+
  
  coord_equal() +
  theme_minimal() +
  guides(
    alpha = "none",
    color = guide_legend(override.aes = list(alpha = 1))
  ) +
  labs(
    x = paste0("PCA axis 1 (", round(summary(pca_light_s)$importance["Proportion of Variance","PC1"]*100, 1), "% of variation)"),
    y = paste0("PCA axis 2 (", round(summary(pca_light_s)$importance["Proportion of Variance","PC2"]*100, 1), "% of variation)")
  )+
  scale_y_continuous(breaks = seq(-10, 10, by = 2))+
  scale_x_continuous(breaks = seq(-10, 10, by = 2))+
  theme(
    legend.position = "bottom",
    legend.direction = "vertical",
    # --- main panel: black frame ---
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    
    # --- side panels: small ---
    ggside.panel.scale = 0.15,          # smaller = thinner strips (try 0.10–0.20)
    
    # --- side panels: no grid ---
    ggside.panel.grid = element_blank(),
    
    # --- side panels: no frame/border of their own, hide their axes ---
    ggside.panel.border = element_blank(),
    ggside.panel.background = element_blank(),
    ggside.axis.text  = element_blank(),
    ggside.axis.ticks = element_blank()
  )-> pca_light_s

# pca_light

light_pca_comp_s <- prep_data(tls_metrics,forest_data_subplot,
                            c(var_pca_light,"padAbove_agg","padAbove_cv"),
                            plot_level = "subplot",t=12,s=0)  %>% 
  pivot_longer(cols = -c("plot","subplot","reg_category")) %>%
  filter(!is.na(reg_category)) %>%
  mutate(name=factor(name,levels = c(var_pca_light,"padAbove_agg","padAbove_cv")),
         reg_category = factor(reg_category)) %>%
  group_by(name) %>% mutate(value = as.numeric(scale(value))) %>% ungroup() %>%
  left_join(var_meta, by = c("name" = "variable")) %>% 
  arrange(name) %>%        # type's level order drives it
  mutate(label=case_when(name=="slope_cross_agg"~paste0(label," 0.9"),
                         name=="z_cross_agg"~paste0(label," 0.9"),
                         TRUE~label),
         label = fct_inorder(label))


cld_one <- function(df) {
  lev <- levels(droplevels(df$reg_category))
  if (length(lev) < 2) return(tibble(reg_category = lev, letter = ""))
  
  kw <- kruskal.test(value ~ droplevels(reg_category), data = df)
  if (is.na(kw$p.value) || kw$p.value > 0.05)
    return(tibble(reg_category = lev, letter = "a"))   # no differences
  
  dt <- dunn_test(df, value ~ reg_category, p.adjust.method = "BH")
  pv <- setNames(dt$p.adj, paste(dt$group1, dt$group2, sep = "-"))
  l  <- multcompLetters(pv)$Letters
  tibble(reg_category = names(l), letter = unname(l))
}

cld_df <- light_pca_comp_s %>%
  group_by(type, label) %>%
  group_modify(~ cld_one(.x)) %>%
  ungroup() %>%
  mutate(reg_category = factor(reg_category, levels = levels(light_pca_comp_s$reg_category))) %>%
  # assign y coordinates
  left_join(
    light_pca_comp_s %>% group_by(type, label, reg_category) %>%
      summarise(y = max(value, na.rm = TRUE), .groups = "drop"),
    by = c("type", "label", "reg_category")
  )


ggplot(light_pca_comp_s, aes(label, value, fill = reg_category)) +
  geom_boxplot(position = position_dodge(width = 0.8, preserve = "single"),
               width=0.5,outlier.shape = NA) +
  geom_point(aes(color=reg_category),
             position = position_jitterdodge(jitter.width = 0.15,
                                             dodge.width = 0.8,
                                             seed = 1),
             size = 1, alpha = 0.8, shape = 16,
             show.legend = FALSE) +
  geom_text(data = cld_df,
            aes(x = label, y = y + 0.25, label = letter, group = reg_category),
            position = position_dodge(width = 0.8),
            inherit.aes = FALSE, size = 3, vjust = 0) +
  scale_fill_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  scale_color_brewer(name = "Regeneration cover", palette = "YlGn", na.value = "grey") +
  facet_grid(~type, scales = "free_x") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.12)))+
  labs(y="Scaled values",x="")+
  theme_minimal(base_size = 9)+
  theme(legend.position = "none")->a_s
# a

cowplot::plot_grid(pca_light_s,a_s,ncol=2,align = "h", axis = "tb",
                   rel_widths = c(1, 1.5))


```


## Light distribution effect on regeneration structure - subplot level

```{r}
explained <- c("richness","richness_sapling","H","cwm_shade","abundance","n_sapling","Height_increment_mn","empty_gf","H_height","fhd","ri","pad_cv")
explaining <- c("lrb_mean","gini_h_rb","slope_ref_agg","bl_diff_abs")

light_effect_s <- prep_data(tls_metrics,forest_data_subplot,c(explained,explaining),
                            plot_level="subplot",
                            t=12,s=0,tg=0.9) %>% 
  mutate(slope_ref_agg=-slope_ref_agg)

order=c(var_meta$label[match(explaining,var_meta$variable)],
        var_meta$label[match(explained,var_meta$variable)])


std_coef <- function(d, y, x) {#, covar = "browsing"
  d <- d %>%
    dplyr::select(all_of(c(y, x, "plot"))) %>%#, covar
    drop_na() %>%
    mutate(plot = factor(plot)) %>%                              # grouping factor, kept out of scaling
    mutate(across(where(is.numeric), ~ as.numeric(scale(.))))    # factors (browsing, plot) left alone
  f <- as.formula(paste0(y, " ~ ",  " + ", x, " + (1 | plot)"))#covar,
  m <- lmer(f, data = d, control = lmerControl(calc.derivs = FALSE))
  unname(fixef(m)[x])
}

# --- bootstrap distribution of that coef ---
boot_coef <- function(d, y, x, B = 1000) {# covar = "browsing",
  d <- d %>% dplyr::select(all_of(c(y, x, "plot"))) %>% drop_na() %>%#covar, 
    mutate(plot = factor(plot)) %>%
    mutate(across(where(is.numeric), ~ as.numeric(scale(.))))
  n <- nrow(d)
  f <- as.formula(paste0(y, " ~ ", " + ", x, " + (1 | plot)"))# covar,
  map_dbl(seq_len(B), function(b) {
    db <- d[sample(n, n, replace = TRUE), ]
    m  <- suppressMessages(suppressWarnings(
      lmer(f, data = db, control = lmerControl(calc.derivs = FALSE))))
    unname(fixef(m)[x])
  })
}

results <- expand_grid(explained = explained, explaining = explaining) %>%
  mutate(
    estimate = map2_dbl(explained, explaining, ~ std_coef(light_effect_s, .x, .y)),
    draws    = map2(explained, explaining, ~ boot_coef(light_effect_s, .x, .y, B = 1000))
  )


boot_long <- results %>%
  dplyr::select(explained, explaining, draws) %>%
  unnest_longer(draws) %>%
  rename(coef = draws)

boot_stats <- boot_long %>%
  group_by(explained, explaining) %>%
  summarise(
    lo   = unname(quantile(coef, 0.10)),
    lo5  = unname(quantile(coef, 0.25)),
    med  = unname(quantile(coef, 0.50)),
    hi5  = unname(quantile(coef, 0.75)),
    hi   = unname(quantile(coef, 0.90)),
    nrep = dplyr::n(),
    # test bilatéral : 2 x la proportion de tirages du "mauvais" côté de 0,
    # bornée par la résolution du bootstrap 1/(B+1)
    p_boot = max(2 * min(mean(coef <= 0), mean(coef >= 0)), 1 / (nrep + 1)),
    .groups = "drop"
  )


boot_stats %>% filter(explaining !="bl_diff_abs") %>% 
  mutate(
    p_class = cut(p_boot,
                  breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                  labels = c("< 0.001", "< 0.01", "< 0.05", "< 0.1", "ns"),
                  right = FALSE)
  ) %>%
  pivot_longer(cols = c("explained", "explaining")) %>%
  left_join(var_meta, by = c("value" = "variable")) %>% 
  dplyr::select(-value) %>%
  pivot_wider(names_from = "name", values_from = c("label", "type")) %>%
  mutate(
    type_explained   = factor(type_explained,
                              levels = c("Composition", "Quantity", "Complexity")),
    label_explained  = factor(label_explained,  levels = rev(order)),
    label_explaining = factor(label_explaining, levels = order)
  ) %>%
  ggplot(aes(x = med, y = label_explained)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = .3, colour = "grey70") +
  geom_linerange(aes(xmin = lo,  xmax = hi,  colour = p_class), linewidth = .9) +
  geom_linerange(aes(xmin = lo5, xmax = hi5, colour = p_class), linewidth = 1.5) +
  geom_point(aes(colour = p_class), size = 1.3) +
  scale_colour_manual(
    values = c("< 0.001" = "#08306b", "< 0.01" = "#2171b5",
               "< 0.05"  = "#6baed6", "< 0.1"  = "#c6dbef",
               "ns"      = "grey78"),
    drop = FALSE, name = "p (bootstrap)"
  ) +
  facet_grid(type_explained ~ label_explaining, scales = "free_y", space = "free_y") +
  labs(x = "Standardized coefficient", y = NULL) +
  theme_bw(base_size = 9) +
  theme(
    panel.spacing.x  = unit(.6, "lines"),
    strip.text.y     = element_text(angle = 0),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )

```

 # Temporal variability and spatial variability

```{r}
# create data
var_metrics <- c('ext_ref_agg', "slope_ref_agg", "gini_h_rb")
light_variability =tls_metrics %>% 
  filter(slice_num==0,subplot!="plot") %>% 
  dplyr::select(any_of(c( c("plot","subplot","time"),var_metrics))) %>% 
  unique() %>%
  mutate(across(all_of(var_metrics), as.numeric)) %>%
  pivot_longer(all_of(var_metrics), names_to = "metric", values_to = "value")


# compute temporal and spatial variability
sd_temporal <- light_variability %>%
  group_by(plot, metric, subplot) %>%
  summarise(sd_t = sd(value, na.rm = TRUE),
            m_sub = mean(value, na.rm = TRUE), .groups = "drop") %>%
  group_by(plot, metric) %>%
  summarise(temporal_within = mean(sd_t, na.rm = TRUE),
            spatial_marg    = sd(m_sub, na.rm = TRUE), .groups = "drop")

sd_spatial <- light_variability %>%
  group_by(plot, metric, time) %>%
  summarise(sd_s = sd(value, na.rm = TRUE),
            m_time = mean(value, na.rm = TRUE), .groups = "drop") %>%
  group_by(plot, metric) %>%
  summarise(spatial_within  = mean(sd_s, na.rm = TRUE),
            temporal_marg   = sd(m_time, na.rm = TRUE), .groups = "drop")

res <- light_variability %>%
  group_by(plot, metric) %>%
  summarise(mean = mean(value, na.rm = TRUE),
            n_sub = n_distinct(subplot), n_date = n_distinct(time),
            .groups = "drop") %>%
  left_join(sd_temporal, by = c("plot", "metric")) %>%
  left_join(sd_spatial,  by = c("plot", "metric")) %>%
  mutate(
    # part de la variabilité totale expliquée par le temps (0 = tout spatial,
    # 1 = tout temporel) -> utile pour colorer / trier les plots
    temporal_part = temporal_within / (temporal_within + spatial_within),
    # versions normalisées (CV) pour comparer des métriques d'échelles
    # différentes ; NA si la moyenne est proche de 0 (cas de slope_ref_agg)
    cv_temporal = ifelse(abs(mean) > 1e-8, temporal_within / abs(mean), NA),
    cv_spatial  = ifelse(abs(mean) > 1e-8, spatial_within  / abs(mean), NA)
  )
res %>% 
  left_join(var_meta,by=c("metric"="variable")) %>% 
  mutate(label=factor(label,levels=c("Mean extinction \n at 4.5m","Gini index","Light profile slope \nat 4.5m"))) %>% 
  ggplot(aes(x = spatial_within, y = temporal_within)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point( size = 2.5) +
  # geom_text(aes(label = plot), vjust = -0.9, size = 3, show.legend = FALSE) +
  scale_colour_gradient2(low = "#2c7bb6", mid = "grey80", high = "#d7191c",
                         midpoint = 0.5, limits = c(0, 1),
                         name = "Part\ntemporelle") +
  facet_wrap(~ label) +
  expand_limits(x = 0, y = 0) +
  # coord_equal() +          # retirer si les 2 axes ont des ordres de grandeur très différents
  labs(
    # title = "Variabilité temporelle vs spatiale, par plot",
    # subtitle = "Au-dessus de la diagonale : la variabilité temporelle domine",
    x = "Spatial variability (between subplots)",
    y = "Temporal variability (between simulation times)"
  ) +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank(), strip.background = element_blank())
```

# Confounding effect of adult structure

## Effect size

```{r}
explained <- c("richness","richness_sapling","H","cwm_shade","abundance","n_sapling","Height_increment_mn","empty_gf","H_height","fhd","ri","pad_cv") 
explaining <- c("n_trees_alive","species_richness_alive","mean_dbh_alive_cm","basal_area_alive_m2","mean_height_alive_m","browsing")

adult_effect<- prep_data(tls_metrics,forest_data_plot,id_cols = c(explained,explaining),
                         plot_level = "plot")


order=c(var_meta$label[match(explaining,var_meta$variable)],
        var_meta$label[match(explained,var_meta$variable)])



std_coef <- function(d, y, x) {
  d <- d %>%
    dplyr::select(all_of(c(y, x))) %>%
    drop_na() %>%
    mutate(across(where(is.numeric), ~ as.numeric(scale(.))))  # factors (e.g. browsing) left alone
  f <- reformulate(c( x), response = y)
  unname(coef(lm(f, data = d))[x])
}

# --- bootstrap distribution of that coef ---
boot_coef <- function(d, y, x,  B = 1000) {
  d <- d %>% dplyr::select(all_of(c(y, x))) %>% drop_na() %>%
      mutate(across(where(is.numeric), ~ as.numeric(scale(.)))) 
  n <- nrow(d)
  f <- reformulate(c( x), response = y)
  map_dbl(seq_len(B), function(b) {
    db <- d[sample(n, n, replace = TRUE), ]  # <- move this line out of the loop to scale once
    unname(coef(lm(f, data = db))[x])
  })
}

results <- expand_grid(explained = explained, explaining = explaining) %>%
  mutate(
    estimate = map2_dbl(explained, explaining, ~ std_coef(adult_effect, .x, .y)),
    draws    = map2(explained, explaining, ~ boot_coef(adult_effect, .x, .y, B = 1000))
  )


boot_long <- results %>%
  dplyr::select(explained, explaining, draws) %>%
  unnest_longer(draws) %>%
  rename(coef = draws)

boot_stats <- boot_long %>%
  group_by(explained, explaining) %>%
  summarise(
    lo   = unname(quantile(coef, 0.10)),
    lo5  = unname(quantile(coef, 0.25)),
    med  = unname(quantile(coef, 0.50)),
    hi5  = unname(quantile(coef, 0.75)),
    hi   = unname(quantile(coef, 0.90)),
    nrep = dplyr::n(),
    # test bilatéral : 2 x la proportion de tirages du "mauvais" côté de 0,
    # bornée par la résolution du bootstrap 1/(B+1)
    p_boot = max(2 * min(mean(coef <= 0), mean(coef >= 0)), 1 / (nrep + 1)),
    .groups = "drop"
  )



boot_stats %>% 
  mutate(
    p_class = cut(p_boot,
                  breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                  labels = c("< 0.001", "< 0.01", "< 0.05", "< 0.1", "ns"),
                  right = FALSE)
  ) %>%
  pivot_longer(cols = c("explained", "explaining")) %>%
  left_join(var_meta, by = c("value" = "variable")) %>% 
  dplyr::select(-value) %>%
  pivot_wider(names_from = "name", values_from = c("label", "type")) %>%
  mutate(
    type_explained   = factor(type_explained,
                              levels = c("Composition", "Quantity", "Complexity")),
    label_explained  = factor(label_explained,  levels = rev(order)),
    label_explaining = factor(label_explaining, levels = order)
  ) %>%
  ggplot(aes(x = med, y = label_explained)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = .3, colour = "grey70") +
  geom_linerange(aes(xmin = lo,  xmax = hi,  colour = p_class), linewidth = .9) +
  geom_linerange(aes(xmin = lo5, xmax = hi5, colour = p_class), linewidth = 1.5) +
  geom_point(aes(colour = p_class), size = 1.3) +
  scale_colour_manual(
    values = c("< 0.001" = "#08306b", "< 0.01" = "#2171b5",
               "< 0.05"  = "#6baed6", "< 0.1"  = "#c6dbef",
               "ns"      = "grey78"),
    drop = FALSE, name = "p (bootstrap)"
  ) +
  facet_grid(type_explained ~ label_explaining, scales = "free_y", space = "free_y") +
  labs(x = "Standardized coefficient", y = NULL) +
  theme_bw(base_size = 9) +
  theme(
    panel.spacing.x  = unit(.6, "lines"),
    strip.text.y     = element_text(angle = 0),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )

```
## Variance partition analysis


```{r}
library(vegan)
library(rdacca.hp)
library(dplyr)
library(ggplot2)
library(patchwork)   # multi-panel figures
library(ggrepel)     # non-overlapping labels

set.seed(42)         # reproducible permutations
n_perm    <- 999    # permutations for RDA tests
n_perm_hp <- 999     # permutations for permu.hp (slower)
alpha     <- 0.1

# --- Parameters: variable names ----------------------------------------------
adult_vars <- c("species_richness_alive", "mean_dbh_alive_cm", "mean_height_alive_m")
light_vars <- c("lrb_mean", "gini_h_rb", "slope_ref_agg")
regen_vars <- c("richness", "abundance", "ri")
log_vars   <- c("abundance")   # responses to log1p-transform (character(0) if none)

# Labels displayed in figures (adapt, especially gini / slope)
display_names <- c(
  species_richness_alive = "Adult species richness",
  mean_dbh_alive_cm      = "Mean DBH",
  mean_height_alive_m    = "Mean height",
  lrb_mean                = "Mean light",
  gini_h_rb              = "Horizontal heterogeneity",
  slope_ref_agg          = "Vertical heterogeneity",
  richness               = "Richness",
  abundance              = "Abundance",
  ri                     = "Complexity"
)

# =============================================================================
# 0. Data preparation
# =============================================================================
dat <- prep_data(tls_metrics,forest_data_plot,id_cols = c(regen_vars, adult_vars, light_vars),
                 plot_level = "plot") 

# Plots excluded because of missing values (to report in the paper)
na_plots <- which(!complete.cases(dat))
if (length(na_plots) > 0)
  message("Plots excluded (NA): ", paste(na_plots, collapse = ", "))
dat <- dat[complete.cases(dat), ]
n <- nrow(dat)
message("Number of plots analysed: n = ", n)

# Transform skewed responses, then standardise (mean 0, variance 1)
standardise <- function(d) d %>% mutate(across(everything(), ~ as.numeric(scale(.))))
regen  <- dat %>% select(all_of(regen_vars)) %>%
  mutate(across(all_of(log_vars), log1p)) %>% standardise()
adults <- dat %>% select(all_of(adult_vars)) %>% standardise()
light  <- dat %>% select(all_of(light_vars)) %>% standardise()
predictors <- cbind(adults, light)

# Collinearity among predictors (rule of thumb: VIF < 5 comfortable, > 10 problematic)
print(round(diag(solve(cor(predictors))), 2))

# =============================================================================
# Shared helpers
# =============================================================================
# Global test + variance partitioning for any response (matrix or single vector)
partition_analysis <- function(Y) {
  full <- rda(Y ~ ., data = predictors)
  vp   <- varpart(Y, light, adults)
  f    <- vp$part$fract$Adj.R.square   # [1] light alone, [2] adults alone, [3] both
  test <- function(m) anova(m, permutations = n_perm)$`Pr(>F)`[1]
  data.frame(
    full_adj_r2    = f[3],
    p_full         = test(full),
    light_total    = f[1],
    p_light_total  = test(rda(Y ~ ., data = light)),
    adult_total    = f[2],
    p_adult_total  = test(rda(Y ~ ., data = adults)),
    light_unique   = f[3] - f[2],
    p_light_unique = test(rda(Y ~ . + Condition(as.matrix(adults)), data = light)),
    shared         = f[1] + f[2] - f[3],             # not testable
    adult_unique   = f[3] - f[1],
    p_adult_unique = test(rda(Y ~ . + Condition(as.matrix(light)), data = adults)),
    residuals      = 1 - f[3]
  )
}

stars <- function(p, ns = "n.s.") ifelse(is.na(p), "",
                                         ifelse(p < 0.001, "***", 
                                                ifelse(p < 0.01, "**", 
                                                       ifelse(p < 0.05, "*",
                                                              ifelse(p < 0.1, "(*)", ns)))))
pretty_name <- function(v) unname(ifelse(v %in% names(display_names), display_names[v], v))

# =============================================================================
# PART A - Multivariate analysis
# =============================================================================
stopifnot(ncol(regen) == length(regen_vars))
model <- rda(regen ~ ., data = predictors)
stopifnot(abs(model$tot.chi - ncol(regen)) < 1e-6)   # total variance = number of responses

# A1. Global test
global_test <- anova(model, permutations = n_perm)
print(global_test)
model_r2 <- RsquareAdj(model)
message(sprintf("Multivariate model: raw R² = %.3f | adjusted R² = %.3f | p = %.4f",
                model_r2$r.squared, model_r2$adj.r.squared, global_test$`Pr(>F)`[1]))
if (global_test$`Pr(>F)`[1] > alpha)
  warning("Multivariate model NOT significant: partitioning results are exploratory only.")

# A2. Variance partitioning with tests
multi <- partition_analysis(regen)
multi_fractions <- data.frame(
  fraction = c("Full model [a+b+c]", "Light (total) [a+b]", "Adults (total) [b+c]",
               "Light (unique) [a]", "Shared [b]", "Adults (unique) [c]", "Residuals [d]"),
  adj_r2   = round(c(multi$full_adj_r2, multi$light_total, multi$adult_total,
                     multi$light_unique, multi$shared, multi$adult_unique, multi$residuals), 3),
  p_value  = c(multi$p_full, multi$p_light_total, multi$p_adult_total,
               multi$p_light_unique, NA, multi$p_adult_unique, NA)
)
print(multi_fractions)
# Reminder: [b] cannot be tested; a negative adjusted fraction is interpreted as 0.

# =============================================================================
# PART B - Univariate analyses (one per regeneration variable)
# =============================================================================
uni_fractions  <- list()
uni_importance <- list()

for (v in regen_vars) {
  y <- regen[[v]]
  
  # B1. Global test + partitioning
  uni_fractions[[v]] <- cbind(response = v, partition_analysis(y))
  
  # B2. Hierarchical partitioning + permutation test of each predictor
  hp   <- rdacca.hp(y, predictors, method = "RDA", type = "adjR2")$Hier.part
  perm <- permu.hp(y, predictors, method = "RDA", type = "adjR2", permutations = n_perm_hp)
  p_col <- grep("^Pr", colnames(perm), value = TRUE)[1]
  uni_importance[[v]] <- data.frame(
    response   = v,
    variable   = rownames(hp),
    group      = ifelse(rownames(hp) %in% adult_vars, "Adults", "Light"),
    unique     = hp[, "Unique"],
    shared     = hp[, "Average.share"],
    individual = hp[, "Individual"],
    percent    = hp[, "I.perc(%)"],
    p_perm     = as.numeric(gsub("[^0-9.eE-]", "", perm[rownames(hp), p_col]))
  )
}

# Holm correction across the 3 responses (one family per type of test)
uni_fractions <- bind_rows(uni_fractions) %>%
  mutate(across(starts_with("p_"), ~ p.adjust(.x, method = "holm"), .names = "{.col}_holm"))
print(uni_fractions, digits = 3)

# Importance: Holm correction across the 6 predictors within each response, then ranking
uni_importance <- bind_rows(uni_importance) %>%
  group_by(response) %>%
  mutate(p_holm = p.adjust(p_perm, method = "holm"),
         rank   = rank(-individual, ties.method = "min")) %>%
  ungroup() %>%
  arrange(response, rank)
print(as.data.frame(uni_importance), digits = 3)
# Only interpret the ranking for responses whose global model is significant.

# =============================================================================
# PART C - Jackknife (leave-one-out)
# =============================================================================
# Each plot is dropped in turn and the analysis refitted on the n - 1 others.
# - jackknife standard error -> approximate 95 % CI (estimate +/- 1.96 SE)
# - min / max over the n refits and the most influential plot -> sensitivity
# (The classic bootstrap is unsuitable with n = 19: resamples contain ~12 distinct
#  plots, which biases the adjusted R² and sometimes saturates the model.)

drop_k <- function(Y, k) if (is.null(dim(Y))) Y[-k] else Y[-k, , drop = FALSE]

# Fractions refitted without plot k (matrix: fractions x plots)
jk_fractions <- function(Y) {
  sapply(seq_len(n), function(k) {
    Yk <- drop_k(Y, k)
    r2 <- function(X) RsquareAdj(rda(Yk ~ ., data = X[-k, , drop = FALSE]))$adj.r.squared
    f_light <- r2(light); f_adult <- r2(adults); f_full <- r2(predictors)
    c(light_unique = f_full - f_adult, shared = f_light + f_adult - f_full,
      adult_unique = f_full - f_light, full_adj_r2 = f_full)
  })
}

# Summary of a jackknife matrix given the full-data estimates (named vector)
jk_summary <- function(jk, estimate) {
  est <- estimate[rownames(jk)]
  se  <- sqrt((n - 1) / n * rowSums((jk - rowMeans(jk))^2))
  data.frame(term = rownames(jk), estimate = est, se_jack = se,
             ci_low = est - 1.96 * se, ci_high = est + 1.96 * se,
             min_loo = apply(jk, 1, min), max_loo = apply(jk, 1, max),
             most_influential_plot = rownames(dat)[apply(abs(jk - est), 1, which.max)],
             row.names = NULL)
}

# C1. Multivariate fractions
multi_jack <- jk_summary(jk_fractions(regen),
                         c(light_unique = multi$light_unique, shared = multi$shared,
                           adult_unique = multi$adult_unique, full_adj_r2 = multi$full_adj_r2))
print(multi_jack, digits = 3)

# C2. Univariate fractions
uni_frac_jack <- bind_rows(lapply(regen_vars, function(v) {
  est <- unlist(uni_fractions[uni_fractions$response == v,
                              c("light_unique", "shared", "adult_unique", "full_adj_r2")])
  cbind(response = v, jk_summary(jk_fractions(regen[[v]]), est))
}))
print(uni_frac_jack, digits = 3)

# C3. Univariate importances
uni_imp_jack <- bind_rows(lapply(regen_vars, function(v) {
  y  <- regen[[v]]
  jk <- sapply(seq_len(n), function(k)
    rdacca.hp(y[-k], predictors[-k, ], method = "RDA", type = "adjR2")$Hier.part[, "Individual"])
  imp <- uni_importance[uni_importance$response == v, ]
  cbind(response = v, jk_summary(jk, setNames(imp$individual, imp$variable)))
}))
uni_importance <- uni_importance %>%
  left_join(uni_imp_jack %>% select(response, variable = term, se_jack, ci_low, ci_high,
                                    min_loo, max_loo, most_influential_plot),
            by = c("response", "variable"))
print(as.data.frame(uni_importance), digits = 3)

# =============================================================================
# FIGURES
# =============================================================================
dir.create("figures", showWarnings = FALSE)

group_colours    <- c("Light" = "#E0A526", "Adults" = "#2F6B4F")
fraction_colours <- c("Light (unique)" = "#E0A526", "Shared" = "#A7A77A",
                      "Adults (unique)" = "#2F6B4F")

theme_set(theme_minimal(base_size = 12) + theme(
  panel.grid.minor = element_blank(),
  plot.title       = element_text(face = "bold", size = 13),
  plot.subtitle    = element_text(colour = "grey35", size = 10),
  plot.caption     = element_text(colour = "grey45", size = 9, hjust = 0),
  axis.title       = element_text(colour = "grey25"),
  legend.position  = "bottom"))

save_fig <- function(plot, name, width, height) {
  ggsave(file.path("figures", paste0(name, ".png")), plot,
         width = width, height = height, dpi = 300, bg = "white")
  ggsave(file.path("figures", paste0(name, ".pdf")), plot,
         width = width, height = height, device = cairo_pdf)
}

# --- Figure 1: multivariate partitioning diagram ------------------------------
# (schematic circles: their size is not proportional to the fractions)
fmt_frac <- function(v) ifelse(v < 0, sprintf("%.2f (≈ 0)", v), sprintf("%.2f", v))
circle <- function(x0, g) {
  t <- seq(0, 2 * pi, length.out = 200)
  data.frame(x = x0 + cos(t), y = sin(t), group = g)
}
circles <- rbind(circle(-0.55, "Light"), circle(0.55, "Adults"))
venn_labels <- data.frame(
  x = c(-1.05, 0, 1.05), y = 0,
  text = c(sprintf("[a]\n%s %s", fmt_frac(multi$light_unique), stars(multi$p_light_unique)),
           sprintf("[b]\n%s",    fmt_frac(multi$shared)),
           sprintf("[c]\n%s %s", fmt_frac(multi$adult_unique), stars(multi$p_adult_unique))))
ci_txt <- function(t) with(multi_jack[multi_jack$term == t, ], sprintf("[%.2f; %.2f]", ci_low, ci_high))
venn_labels$text <- paste0(venn_labels$text, "\n", c(ci_txt("light_unique"), ci_txt("shared"), ci_txt("adult_unique")))
titles <- data.frame(x = c(-0.95, 0.95), y = 1.2, group = c("Light", "Adults"))

fig_partition <- ggplot() +
  geom_polygon(data = circles, aes(x, y, fill = group, group = group),
               alpha = 0.35, colour = "white", linewidth = 1.2) +
  geom_text(data = venn_labels, aes(x, y, label = text),
            size = 4.2, fontface = "bold", colour = "grey15", lineheight = 0.9) +
  geom_text(data = titles, aes(x, y, label = group, colour = group),
            size = 5, fontface = "bold") +
  annotate("text", x = 0, y = -1.3, colour = "grey40", size = 3.8,
           label = sprintf("Residuals = %.2f", multi$residuals)) +
  scale_fill_manual(values = group_colours) +
  scale_colour_manual(values = group_colours) +
  coord_equal(xlim = c(-1.8, 1.8), ylim = c(-1.45, 1.4)) +
  labs(title = "Variance partitioning of regeneration structure (multivariate RDA)",
       subtitle = sprintf("Adjusted R² [jackknife 95 %% CI]  |  full model: adj. R² = %.2f, p = %.3f",
                          multi$full_adj_r2, multi$p_full)) +
  theme_void() +
  theme(legend.position = "none",
        plot.title    = element_text(face = "bold", size = 13, hjust = 0.5),
        plot.subtitle = element_text(colour = "grey35", size = 10, hjust = 0.5))
print(fig_partition)
save_fig(fig_partition, "fig1_multivariate_partitioning", 7, 5)

# --- Figure 2A: variance explained per response -------------------------------
resp_levels <- regen_vars
fig2a_data <- bind_rows(
  data.frame(response = uni_fractions$response, fraction = "Light (unique)",
             value = uni_fractions$light_unique, p = uni_fractions$p_light_unique_holm),
  data.frame(response = uni_fractions$response, fraction = "Shared",
             value = uni_fractions$shared, p = NA),
  data.frame(response = uni_fractions$response, fraction = "Adults (unique)",
             value = uni_fractions$adult_unique, p = uni_fractions$p_adult_unique_holm)) %>%
  mutate(term = c("Light (unique)" = "light_unique", "Shared" = "shared",
                  "Adults (unique)" = "adult_unique")[fraction]) %>%
  left_join(uni_frac_jack %>% select(response, term, ci_low, ci_high), by = c("response", "term")) %>%
  mutate(fraction = factor(fraction, levels = names(fraction_colours)),
         response = factor(response, levels = resp_levels),
         negative = value < 0)

# offset <- 0.04 * max(abs(c(fig2a_data$value, fig2a_data$ci_high)), na.rm = TRUE)
# fig2a_data$y_star <- pmax(fig2a_data$ci_high, fig2a_data$value, 0, na.rm = TRUE) + offset
offset <- 0.04 
fig2a_data$y_star <- pmax( fig2a_data$value, 0, na.rm = TRUE) + offset



# Full-model label above each response
fig2a_top <- uni_fractions %>%
  mutate(response = factor(response, levels = resp_levels),
         label = sprintf("Model: adj. R² = %.2f %s", full_adj_r2, stars(p_full)))
y_top <- max(fig2a_data$y_star) + 3 * offset

fig2a <- ggplot(fig2a_data, aes(x = response, y = value, fill = fraction)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  # geom_errorbar(aes(ymin = ci_low, ymax = ci_high, group = fraction),
  #               position = position_dodge(width = 0.8), width = 0.25, colour = "grey25") +
  geom_text(aes(y = y_star, label = stars(p, ns = "n.s."), group = fraction),
            position = position_dodge(width = 0.8), vjust = 0, size = 3.6, colour = "grey20") +
  geom_text(data = fig2a_top, aes(x = response, y = y_top, label = label),
            inherit.aes = FALSE, size = 3.3, colour = "grey25") +
  scale_fill_manual(values = fraction_colours, name = NULL) +
  scale_alpha_manual(values = c(`FALSE` = 1, `TRUE` = 0.4), guide = "none") +
  scale_x_discrete(labels = pretty_name) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.12))) +
  labs(x = NULL, y = "Variance explained (adjusted R²)",
       title = "Variance explained by light and adults",
       subtitle = "Stars: permutation tests of unique fractions,\nHolm-corrected across responses") +
  theme(panel.grid.major.x = element_blank())

# --- Figure 2B: importance ranking per response -------------------------------
var_order <- c(rev(light_vars), rev(adult_vars))   # adults on top, then light
fig2b_data <- uni_importance %>%
  mutate(response = factor(response, levels = resp_levels),
         variable = factor(variable, levels = var_order),
         group    = factor(group, levels = c("Adults", "Light")),
         label    = sprintf("#%d%s", rank, if_else(p_perm>alpha,"",paste0("\n", stars(p_perm)))))

fig2b <- ggplot(fig2b_data, aes(x = response, y = variable, fill = individual)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(aes(label = label), size = 3.6, lineheight = 0.85, colour = "grey10") +
  facet_grid(group ~ ., scales = "free_y", space = "free_y") +
  scale_fill_gradient2(low = "#B2443A", mid = "white", high = "#2B4C7E", midpoint = 0,
                       name = "Individual importance\n(adjusted R²)") +
  scale_x_discrete(labels = pretty_name, position = "top") +
  scale_y_discrete(labels = pretty_name) +
  labs(x = NULL, y = NULL,
       title = "Importance ranking of predictors",
       subtitle = "#: rank within each response  |  Stars: permutation test, Holm-corrected") +
  theme(panel.grid = element_blank(),
        strip.text.y = element_text(face = "bold", angle = 0, size = 11),
        legend.key.width = unit(1.2, "cm"))

# --- Figure 2: combined --------------------------------------------------------
fig_univariate <- fig2a + fig2b +
  plot_layout(widths = c(1.3, 1)) +
  plot_annotation(tag_levels = "A")
print(fig_univariate)

# --- Figure 3: multivariate RDA triplot (scaling 2) ---------------------------
# Reading: angle between arrows ~ correlation (acute = positive, 90° = none,
# obtuse = negative); only interpret significant axes (see axis_test).
axis_test <- anova(model, by = "axis", permutations = n_perm)
print(axis_test)

regen_colour <- "#2B4C7E"
axis_pct   <- round(100 * model$CCA$eig[1:2] / model$tot.chi, 1)
axis_p     <- axis_test$`Pr(>F)`[1:2]
rda_scores <- scores(model, scaling = 2, choices = 1:2, display = c("sites", "species", "bp"))
rename_axes <- function(d) { colnames(d)[1:2] <- c("axis1", "axis2"); d }
plots     <- rename_axes(as.data.frame(rda_scores$sites))
responses <- rename_axes(as.data.frame(rda_scores$species)); responses$name <- rownames(responses)
arrows_bp <- rename_axes(as.data.frame(rda_scores$biplot));  arrows_bp$name <- rownames(arrows_bp)
arrows_bp$group <- ifelse(arrows_bp$name %in% adult_vars, "Adults", "Light")
spread <- 0.85 * max(abs(plots[, 1:2]))   # scale arrows to the spread of plots
arrows_bp[, 1:2] <- arrows_bp[, 1:2] * spread / max(abs(arrows_bp[, 1:2]))
responses[, 1:2] <- responses[, 1:2] * spread / max(abs(responses[, 1:2]))

fig_triplot <- ggplot() +
  geom_hline(yintercept = 0, colour = "grey85") +
  geom_vline(xintercept = 0, colour = "grey85") +
  geom_point(data = plots, aes(axis1, axis2), colour = "grey60", size = 2, alpha = 0.8) +
  geom_segment(data = arrows_bp, aes(x = 0, y = 0, xend = axis1, yend = axis2, colour = group),
               arrow = arrow(length = unit(0.18, "cm")), linewidth = 0.8) +
  geom_segment(data = responses, aes(x = 0, y = 0, xend = axis1, yend = axis2),
               colour = regen_colour, arrow = arrow(length = unit(0.22, "cm")), linewidth = 1.2) +
  geom_text_repel(data = arrows_bp, aes(axis1, axis2, label = pretty_name(name), colour = group),
                  size = 3.4, show.legend = FALSE, seed = 1) +
  geom_text_repel(data = responses, aes(axis1, axis2, label = pretty_name(name)),
                  colour = regen_colour, fontface = "bold", size = 3.8, seed = 1) +
  scale_colour_manual(values = group_colours, name = NULL) +
  coord_equal() +
  labs(x = sprintf("RDA axis 1 (%.1f %%, p = %.3f)", axis_pct[1], axis_p[1]),
       y = sprintf("RDA axis 2 (%.1f %%, p = %.3f)", axis_pct[2], axis_p[2]),
       title = "Multivariate RDA triplot",
       subtitle = "Blue: regeneration variables  |  Points: plots  |  Scaling 2")
print(fig_triplot)

```

# Species analysis

```{r eval=FALSE, include=FALSE}
data.frame(plot=unique(regen_metrics_subplot$species$plot)) %>% 
  expand_grid(subplot=unique(regen_metrics_subplot$species$subplot)) %>% 
  expand_grid(Species=unique(regen_metrics_subplot$species$Species)) %>% 
  filter(Species!="NA") %>% 
  left_join(regen_metrics_subplot$species) %>% 
  dplyr::select(plot,subplot,species=Species,abundance) %>% 
  mutate(abundance=if_else(is.na(abundance),0,abundance)) %>% 
  left_join(metric3d_all %>% 
              filter(time==12,subplot!="plot") %>% 
              dplyr::select(plot,subplot,rb_mean,rb_cv,occupied_gf)) %>% 
  left_join(H_ext_all %>% 
              filter(time==12,extinction==0.5) %>% 
              dplyr::select(plot,subplot,mean,mean_slope)) %>% 
  group_by(species) %>% 
  mutate(n=n()) %>% 
  filter(n>30) %>% 
  dplyr::select(-n) %>% 
  pivot_longer(cols=-c("plot","subplot","species","abundance")) %>% 
  ggplot(aes(abundance,value))+
  geom_point()+
  facet_wrap(species~name,scales="free", ncol=5)



regen_metrics_plot$species %>% 
  dplyr::select(plot,species=Species,abundance) %>% 
  left_join(metric3d_all %>% 
              filter(time==12,subplot=="plot") %>% 
              dplyr::select(plot,subplot,slice_num,rb_mean,rb_cv,occupied_gf)) %>% 
  left_join(H_ext_all %>% 
              filter(time==12,extinction==0.5) %>% 
              dplyr::select(plot,subplot,mean,mean_slope)) %>% View()
  filter(plot!="GER10") %>% 
  group_by(species) %>% 
  mutate(n=n()) %>% 
  filter(n>30) %>% 
  dplyr::select(-n) %>% 
  ggplot(aes(log(abundance),log(rb_mean),color=species))+
  geom_point()+
  geom_smooth(method="gam",se = FALSE)+
  facet_grid(species~slice_num)
  
  
data.frame(plot=unique(regen_metrics_subplot$species$plot)) %>% 
  # expand_grid(subplot=unique(regen_metrics_subplot$species$subplot)) %>% 
  expand_grid(Species=unique(regen_metrics_subplot$species$Species)) %>% 
  filter(Species!="NA") %>% 
  left_join(regen_metrics_subplot$species) %>% 
  dplyr::select(plot,species=Species,abundance) %>% 
  mutate(abundance=if_else(is.na(abundance),0,abundance)) %>% 
  left_join(metric3d_all %>% 
              filter(time==12,subplot=="plot") %>% 
              dplyr::select(plot,subplot,slice_num,rb_mean,rb_cv,occupied_gf)) %>% 
  left_join(H_ext_all %>% 
              filter(time==12,extinction==0.5) %>% 
              dplyr::select(plot,subplot,mean,mean_slope)) %>%
  filter(plot!="GER10") %>% 
  group_by(species) %>% 
  mutate(n=n()) %>% 
  filter(n>30) %>% 
  dplyr::select(-n) %>% 
  ggplot(aes(abundance,log(rb_mean),color=species))+
  geom_point()+
  geom_smooth(method="gam",se = FALSE)+
  facet_grid(species~slice_num)
  



df <- regen_metrics_subplot$all %>% dplyr::select(plot,subplot) %>% unique() %>% 
  expand_grid(Species=unique(regen_metrics_subplot$species$Species))%>%
  filter(Species != "NA") %>%
  left_join(regen_metrics_subplot$species) %>%
  dplyr::select(plot, subplot, species = Species, abundance) %>%
  mutate(abundance=if_else(is.na(abundance),0,abundance),
         presence = if_else(is.na(abundance), "absent", "present")) %>%
  left_join(metric3d_all %>%
              filter(time == 12, subplot != "plot", slice_num == 1) %>%
              dplyr::select(plot, subplot, rb_mean, rb_cv, occupied_gf)) %>%
  mutate(log_rb_mean = log(rb_mean)) %>%
  filter(is.finite(log_rb_mean))



df %>%
  group_by(species) %>%
  filter(n_distinct(presence) == 2, all(table(presence) >= 1)) %>%  # need both groups, a few obs each
  wilcox_test(log_rb_mean ~ presence) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

regen_metrics_subplot$all %>% dplyr::select(plot,subplot) %>% unique() %>% 
  expand_grid(Species=unique(regen_metrics_subplot$species$Species)) %>% 
  filter(Species!="NA") %>% 
  left_join(regen_metrics_subplot$species) %>% 
  dplyr::select(plot,subplot,species=Species,abundance) %>% 
  mutate(presence=if_else(is.na(abundance),"absent","present")) %>% 
  left_join(metric3d_all %>% 
              filter(time==12,subplot!="plot",slice_num==1) %>% 
              dplyr::select(plot,subplot,rb_mean,rb_cv,occupied_gf)) %>% 
  ggplot(aes(species,log(rb_mean),color=presence))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter()+
  ylim(c(5.5,9))
  



df <- regen_metrics_plot$all %>% dplyr::select(plot) %>% unique() %>% 
  expand_grid(Species=unique(regen_metrics_plot$species$Species)) %>% 
  filter(Species!="NA") %>% 
  left_join(regen_metrics_plot$species) %>% 
  dplyr::select(plot,species=Species,abundance) %>% 
  mutate(presence=if_else(is.na(abundance),"absent","present")) %>% 
  left_join(metric3d_all %>% 
              filter(time==12,subplot=="plot",slice_num==1) %>% 
              dplyr::select(plot,rb_mean,rb_cv,occupied_gf)) %>% 
  mutate(log_rb_mean = log(rb_mean)) %>%
  filter(is.finite(log_rb_mean))

regen_metrics_plot$all %>% dplyr::select(plot) %>% unique() %>% 
  expand_grid(Species=unique(regen_metrics_plot$species$Species)) %>% 
  filter(Species!="NA") %>% 
  left_join(regen_metrics_plot$species) %>% 
  dplyr::select(plot,species=Species,abundance) %>% 
  mutate(presence=if_else(is.na(abundance),"absent","present")) %>% 
  left_join(metric3d_all %>% 
              filter(time==12,subplot=="plot",slice_num==1) %>% 
              dplyr::select(plot,rb_mean,rb_cv,occupied_gf)) %>% 
  ggplot(aes(species,log(rb_mean),color=presence))+
  geom_boxplot(outlier.colour = NA)+
  geom_jitter()+
  ylim(c(5.5,9))
  
  
library(lme4); library(lmerTest)

species_list <- df %>%
  group_by(species) %>%
  filter(n_distinct(presence) == 2, n_distinct(plot) >= 3) %>%
  pull(species) %>% unique()

results <- lapply(species_list, function(sp) {
  d <- df[df$species == sp, ]
  tryCatch({
    m  <- lm(log_rb_mean ~ presence , data = d)
    co <- summary(m)$coefficients["presencepresent", ]
    data.frame(species = sp,
               estimate  = co["Estimate"],
               std.error = co["Std. Error"],
               statistic = co["t value"],
               p.value   = co["Pr(>|t|)"])
  }, error = function(e)
    data.frame(species = sp, estimate = NA, std.error = NA,
               statistic = NA, p.value = NA))
})

res <- do.call(rbind, results)
res$p.adj <- p.adjust(res$p.value, method = "BH")
res




df$rb_c <- df$log_rb_mean - mean(df$log_rb_mean, na.rm = TRUE)   # centred predictor
df$species <- factor(df$species)

library(glmmTMB)

mnb <- glmmTMB(abundance ~ species * rb_c + (1 | plot),
               data = df, family = nbinom2)

# if there are more zeros than even NB expects (common), add a zero-inflation part:
mzi <- glmmTMB(abundance ~ species * rb_c + (1 | plot),
               ziformula = ~1, data = df, family = nbinom2)

AIC(mnb, mzi)  


library(emmeans)

et <- emtrends(mnb, ~ species, var = "rb_c")   # or mp / mzi
et                                             # per-species slope on log scale

# as rate ratios (abundance multiplies by this per 1-unit rb_mean):
summary(et, infer = TRUE) |>
  transform(RR = exp(rb_c.trend))
```

